### 2011-02-08 doc/MVSeg.tex
ENH: project overview

\documentclass[11pt,english]{article}

% Set page margins correctly
\usepackage{geometry}
\usepackage{url}
\geometry{letterpaper,top=1.0in,left=1.0in,bottom=1.0in,top=1.0in,headsep=6pt,footskip=18pt}

\usepackage{lscape}
\usepackage[square,comma,numbers,sort&compress]{natbib}

% Use fancy header style
\usepackage{fancyhdr}
\pagestyle{empty}
\renewcommand{\headrulewidth}{0.75pt}
\renewcommand{\footrulewidth}{0.75pt}
\usepackage{setspace}
\usepackage{listings}
\usepackage{float}
\floatstyle{plain}
\newfloat{command}{thp}{lop}
\floatname{command}{Command}
\doublespacing
\usepackage{boxedminipage}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{babel,verbatim}
\usepackage{enumerate}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{sectsty}
\usepackage[compact]{titlesec}
\usepackage[usenames]{color}
\usepackage{ulem}
\usepackage{multirow,booktabs,ctable,array}
\graphicspath{{./Figures/}
                          }

%\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}

\long\def\symbolfootnote[#1]#2{\begingroup%
\def\thefootnote{\fnsymbol{footnote}}\footnote[#1]{#2}\endgroup}

    \usepackage{color}

    \definecolor{listcomment}{rgb}{0.0,0.5,0.0}
    \definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
    \definecolor{listnumbers}{gray}{0.65}
    \definecolor{listlightgray}{gray}{0.955}
    \definecolor{listwhite}{gray}{1.0}

\begin{document}
\normalem

\vspace*{5cm}

\begin{center}
{\Large \bf Multivariate EM Segmentation Project} \\
\vspace*{0.5cm}
{\normalsize Ben Kandel, Pengfei Zheng and Brian B. Avants$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.}
\end{singlespace}
\end{center}

\vfill

\begin{singlespace} 
\scriptsize
\flushleft
%\line(1, 0){250} \\
{\bf MVSeg}\\
Corresponding author: \\
Brian B. Avants\\
3600 Market Street, Suite 370\\
Philadelphia, PA  19104\\
avants@picsl.upenn.edu\\
\end{singlespace} 
\clearpage
\begin{abstract} 
Neuroanatomical coordinate systems are essential for the
interpretation of structural and functional imaging studies.
However, manual delineation of the neuroanatomical complex is time
consuming and prone to random performance variability.  This
work describes an open source, image-based approach to white matter parcellation which uses training data to propagate
structural labelings to individual images.  The Bayesian formulation of the segmentation problem is 
solved using the Expectation Maximization (EM) algorithm with a
variety of different multivariate distance metrics.  We evaluate our
ability to segment DTI data based on comparison with
registration-based approaches and biological validity of results. 
\end{abstract}

\section{Introduction}
The Expectation-Maximization (EM) framework \citep{Dempster1977} is a
powerful optimization method for parameter estimation.

\section{Data}

\section{Methods}

\section{Results}

\section{Discussion}

\paragraph{Information Sharing Statement}
{\color{blue}{Atropos software is available in ANTs
    \url{http://www.picsl.upenn.edu/ANTs} 
which depends on ITK \url{http://www.itk.org/Wiki/ITK/Git/Download}.
The data used in this work is available in the ANTs software
repository, BrainWeb 
\url{http://mouldy.bic.mni.mcgill.ca/BrainWeb/} and at \url{www.brain-development.org}.
We employed itk-SNAP for visualization \url{www.itksnap.org}.}}

\paragraph{Acknowledgments}
{This work was supported in part by NIH (AG17586, AG15116, NS44266, and
NS53488).}

\newpage

\bibliographystyle{plain}
\bibliography{MVSeg} 
\end{document}

### 2011-02-09 doc/MVSeg.tex
ENH: notes and references on relevant work

@@
 Neuroanatomical coordinate systems are essential for the
-interpretation of structural and functional imaging studies.
-However, manual delineation of the neuroanatomical complex is time
-consuming and prone to random performance variability.  This
-work describes an open source, image-based approach to white matter parcellation which uses training data to propagate
-structural labelings to individual images.  The Bayesian formulation of the segmentation problem is 
+interpretation of structural and functional imaging studies.  However,
+manual delineation of the neuroanatomical complex is time consuming
+and prone to random performance variability.  This work describes an
+open source, image-based approach to white matter parcellation which
+uses training data to propagate structural labelings to individual
+images.  The Bayesian formulation of the segmentation problem is
 solved using the Expectation Maximization (EM) algorithm with a
@@
 The Expectation-Maximization (EM) framework \citep{Dempster1977} is a
-powerful optimization method for parameter estimation.
+powerful optimization method for parameter estimation.  This project
+will investigate EM with application to diffusion tensor imaging and
+will test our ability to segment major white matter tracts in
+different subjects.
 
 \section{Data}
+\subsection{DTI}
+We can start small with a couple of interesting datasets.
 
-\section{Methods}
+\noindent $I$ High-resolution human DT atlas in nifti format: \url{http://www.nitrc.org/frs/?group_id=432}
+
+\noindent $J$ A second human DT atlas
+\url{http://www.nitrc.org/projects/dtitk} which also has labels that
+may be used to guide our parcellation of the white matter into major
+tracts.
+
+\noindent $K$ A similar atlas for macaque \url{http://www.nitrc.org/projects/rmdtitemplate}
+
+Segmenting these templates into similar regions will let us quantify
+differences in white matter due to aging in humans ($I$ vs $J$) and
+between humans and macaques ($I$ and $J$ vs $K$).
+
+Once we establish a working knowledge of DTI segmentation, we can
+evaluate reproducibility of our methods on the Kirby dataset
+\url{http://www.nitrc.org/projects/multimodal/}.  This unique
+collection of images has repeat imaging on the same subject within a
+week.  Thus, our segmentation tools should provide repeatable
+measurements of volume or tracts across all subjects.  Once
+repeatability is established, we can apply the segmentation approach
+to other datasets where investigating population differences is of
+interest.
+
+\subsection{Neonate T1 and T2} There is also an interesting neonatal
+dataset for which there is some multivariate data (T1 and T2).  The
+goal with the neonatal segmentation is to increase repeatability
+between automated and manual volume measurements.  We do fairly well
+already but could improve results with some additional work on, for
+instance, distance metrics and/or partial volume estimation.
+
+\section{Methods} Our goal will be to provide reproducible
+measurements derived from DTI segmentation.
+\begin{description}
+\item[RBG:] The principal direction of the tensor can be visualized in
+  RGB space.  Some segmentation methods rely on this feature 
+  modulated by the FA.
+\item[FA:] Fractional anisotropy derived from the tensor.  This is
+  equal to the variance of the eigenvalues of the tensor matrix.  It
+  is the most common measure used in DTI studies, along with
+  tractography. 
+\item[Tractography:] White matter tracts provide a physical connection
+  between and within cortical regions.  At the coarse scale of DTI, we mostly
+  look at connections between cortical regions.  A simple way to
+  define a tract from DTI is to follow the principal eigenvector until
+  some stopping criterion based on curvature or FA is reached. 
+\end{description}
+An EM segmentation method can use any of these features.  One project
+that may be similar to our own is here
+\url{http://www.nitrc.org/projects/dots} though there is no clear
+reference other than the website.  Prior DT segmentation work is here
+\citep{Lenglet2005} and here \citep{Awate2007}.  Classic work on
+partial volume estimation \citep{Santago1993,Shattuck2001}.
+
+We can also explore the effect of simulated and/or deterministic
+annealing which is a hidden feature of Atropos.
 
@@
 \paragraph{Information Sharing Statement}
-{\color{blue}{Atropos software is available in ANTs
-    \url{http://www.picsl.upenn.edu/ANTs} 
-which depends on ITK \url{http://www.itk.org/Wiki/ITK/Git/Download}.
-The data used in this work is available in the ANTs software
-repository, BrainWeb 
-\url{http://mouldy.bic.mni.mcgill.ca/BrainWeb/} and at \url{www.brain-development.org}.
-We employed itk-SNAP for visualization \url{www.itksnap.org}.}}
+{Atropos software is available in ANTs
+    {\color{blue}{\url{http://www.picsl.upenn.edu/ANTs}}}
+which depends on ITK {\color{blue}{\url{http://www.itk.org/Wiki/ITK/Git/Download}}}.
+BrainWeb provides good evaluation data
+{\color{blue}{\url{http://mouldy.bic.mni.mcgill.ca/BrainWeb/}}}.
+ITK-SNAP is useful for scalar and RGB visualization {\color{blue}{\url{www.itksnap.org}}}.
 

### 2011-02-09 src/ANTS/Documentation/atropos/ResponseToReviewers.tex
ENH: Adding ants repository from February 9, 2011

\documentclass[11pt]{article}
\newcommand{\X}{{\bf X}}
\newcommand{\x}{{\bf x}}
\newcommand{\y}{{\bf y}}
\renewcommand{\v}{{\bf v}}
\renewcommand{\textfraction}{0.05}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}

\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\usepackage{verbatim}
\usepackage[usenames]{color}
\usepackage{listings}
\lstset{breaklines=true,basicstyle=\ttfamily,breakindent=0pt}
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\begin{document}
%\maketitle
\noindent
David Kennedy, Ph.D.,\\
Co-Editor-in-Chief, Neuroinformatics\\

\noindent
Dear Dr. Kennedy and Reviewers,\\

\noindent
My coauthors and I would like to thank the editors and reviewers for
handling our manuscript entitled ``An Open Source Multivariate Framework for $n$-Tissue
  Segmentation with Evaluation on Public Data.''  The reviews were extremely helpful in highlighting both
the strengths and weaknesses of the previous version of the paper.  In
summary, the review found the work to be worthy of publication, with
revisions that focused on improving clarity of some key points. We make every effort, in this response letter and the
associated revision, to clarify our work and improve the paper
according to the reviewers' suggestions. We believe that we carefully
revised the manuscript to address all concerns and provide responses
to the reviewers' comments, which you will find below.   We
also elected to add detail regarding the ICM strategy
available in Atropos. Thus, we would
like to resubmit the revised manuscript which we hope will be suitable
for publication. \\

\noindent
The revision of the paper includes:
\begin{enumerate}
\item  A new figure revision highlighting ICM and providing a simple
  starter example for Atropos usage. 
\item Typo fixes in the equations. 
\item  Many minor changes in the body of the text.  
\item  The references were altered according to reviewer suggestions.
\item  An information sharing statement. 
\end{enumerate}
We express our sincere gratitude to
the editorial staff and to the reviewers for their efforts in
evaluating this paper. \\

\noindent
Sincerely,\\

\noindent
Brian Avants, Ph.D.\\
Penn Image Computing and Science Laboratory \\ 
University of Pennsylvania\\
3600 Market Street, Ste 370\\
Philadelphia, PA 19104
\newpage
\begin{center}
\LARGE{{\bf Response To Reviewers}}
\end{center}
\section*{Reviewer 1}
{\em This paper summarised a semi-automatic segmentation tool called Atropos primarily targeting the brain. It uses EM based methods and can deal with a high number of labels. Overall it is well written, and the availability of the source code is great.
Although the techniques used are not novel, the paper is primarily a description of the Atropos tool and does this very well. It gives good background information as well as explains the theoretical aspects of the work clearly.}
\newline
\newline
{\em Furthermore, I obtained the source code and compiled it successfully. Unfortunately, I could not get it to work on our in-house data before submitting this review.}
\newline
\newline
{\color{black}{Please let us know through Sourceforge (http://sourceforge.net/projects/advants/) if you run into any compilation troubles and/or troubles running the code.  In addition to Atropos adhering to strict ITK coding practices and standards, we recently turned ensured that all compilation warnings using gcc 4.2.1 on Mac OS X 10.6 were completely resolved.
One of our motivations for making Atropos open source is the same for most of software---interaction with and feedback from users can only make our code better and more useful to us.  }}
\newline
\newline
{\em Below are my comments split into two sections. The first are primarily technical whereas the second section contains grammatical corrections some of which may be ignored at the authors discretion.}
\newline
\newline
We thank reviewer 1 for his/her careful reading for technical content
and for helping us to improve our work.\\
\newline
\newline
{\em 1) Page 5, 2nd line from bottom. What are silver standard
  labellings?}
\newline
\newline
{\color{black}{We realize that the use of this term in distinguishing levels of confidence in a given data set might be a source of unnecessary confusion so we have eliminated its use.  The relevant passage currently reads (first paragraph, page 6):
\begin{quote}
Although Atropos may be applied to multivariate data from arbitrary modalities, we limit our evaluation to tissue classification in T1 neuroimaging in part due to the abundance of “gold-standard” data for this modality.
\end{quote}
}}
\noindent
{\em 2) Page 10, Equations 8--11. These follow standard
  approaches. However, Equation 11 does not seem consistent with [8]
  and [10]. If delta(x\_i,x\_j)=1 for x\_i=x\_j, then U(x) as defined in
  [10] increases, and p(x) as defined in [8] decreases - hence
  labelling similar regions seems to be penalised. However, things
  seem consistent if equation 11 is changed as in the standard graph
  cuts approaches - e.g. see equation 3 Boykov \& Jolly ICCV 2001.}
\newline
\newline
We appreciate the reviewer catching this error in the equation.  If $x_i = x_j$ then $V=0$ and is
1 otherwise, as is standard in Besag's approach. 
\newline
\newline
{\em 3) Page 10, Equation 11. Computing the value of U(x) only
  considers if neighbours are the same or not. However, as this paper
  deals with segmenting a large number of labels why was an approach
  such as that taken in Noe \& Gee IPMI 2001 not considered -
  i.e. learning the probability of labels being neighbours and using
  these?}
\newline
\newline
This is an interesting point and we intend to investigate this
direction of research in future work.  In fact, some basic
functionality along these lines is available within the guts of the
code but additional comparative research into the myriad partial volume
estimation models that have been proposed is required to determine the
best approach.  Thus, we added comments in the discussion to indicate
our interest in pursuing partial volume models but elect not to pursue
this task for this resubmission. 
\newline
\newline
{\em 4) Page 12, lines 10-14. Authors comment that applying graph cuts
  to more than one label is not globally optimal and that EM is a
  reasonable and efficient alternative. Although this is true,another
  property of graph cuts is that it is fast. How does the EM approach
  compare in terms of speed?}
\newline
\newline
Fast is a relative term and we cannot comment on the speed of a method
for which we do not have an optimal implementation.  
The main computational challenge with graph cuts is memory.  Because one needs to
solve multiple shortest path problems---and store the solution of
each---the potential for memory bloat is high.  Even in EM, memory
concerns emerge when enough labels are used (e.g. in our examples).
In theory, though, EM computations are embarrassingly parallel which is
not a property shared by graph-based methods.  As such, one would
expect even an optimally implemented multi-label graph cut method to
lag in speed behind an optimally implemented EM segmentation.

{\color{black}{ We would also
like to point out that we have had prior experience with graph cuts (see,
for example, our implementation available at http://hdl.handle.net/1926/1503
and related publications:  Zhuang Song, Nicholas J. Tustison, Brian B. Avants, 
James C. Gee: Integrated Graph Cuts for Brain MRI Segmentation. MICCAI (2) 2006: 831-838
and Zhuang Song, Nicholas J. Tustison, Brian B. Avants, James C. Gee: 
Integrated Graph Cuts for Brain MRI Segmentation. MICCAI (2) 2006: 831-838) and
our preference for our current EM-based approach is certainly not restricted to 
speed considerations.  As enthusiastic supporters of open source software,
patent issues constrain the use of graph cuts where no such limitations exist
for EM-based approaches.}}
\newline
\newline
{\em 5) Page 13, Equation 17. This looks like a multivariate equation to me. I'm also not familiar with the form of Equation 19 - why is it different from 17?'}
\newline
\newline
The difference between equations 17 and 19 is that the input to the
data term is a vector (rather than scalar) and the scalar variance
parameter is replaced by the covariance matrix.  This is a standard
extension from univariate to multivariate probability. 
\newline
\newline
{\em 6) Page 17. Penultimate line. ``Three likelihood classes have been
  developed, one parametric and one ...''. Should this be ``Three
  likelihood classes have been developed, one parametric and [two]
  ...''?'}
\newline
\newline
Yes, thank you.  The fix is made.  
\newline
\newline
{\em 7) Page 26 Figure 5. In the sign column the signs for the Left
  and Right Caudate Nucleus and Thalamus should be -ve as according to
  the data in the table, Atropos performed significantly worse than
  the majority vote.}
\newline
\newline
Fixed.  
\newline
\newline
{\em 8) I could not verify the validity of all the references as they were not numbered neither were they in alphabetical order. }
\newline
\newline
{\color{black}{We apologize.  In preparing the original
manuscript, we were unable to find a .bst for the
Neuroinformatics journal so we created our own using
natbib and did not use all the options to facilitate
the reviewing process.  The references are now numbered
in the manuscript and in the bibliography section and
are ordered as they appear in the manuscript.   }}
\newline
\newline
The reviewer also mentioned not being able to locate the 19 Hammers
datasets.  It is unfortunate that the brain-development website is not
well annotated.  We suspect the reviewer looked at the IXI data which
is not the Hammers data.  We suggest the reviewer look at the adult
atlas section of the website which, currently, contains the data we
used.  This is noted in the revised text. 
\newline
\newline
We also followed each of the reviewer's suggestions on the structure
and grammar of the text. 
\newline
\newline
We appreciate reviewer 1's time and effort.  
\newline
\newline
\section*{Reviewer 2}
We thank reviewer 2 for his/her attention to the manuscript and overall 
positive comments.  We made the reviewer's suggested minor changes in this revision.  
The following comments were addressed: 
\newline
\newline
{\em I was very pleased with the clarity of most of the article,
 particularly with the careful introductory explanations,
 demonstrations of the command-line arguments for Atropos, and the
 excellent review of the literature. I was also happy to see the
 multivariate results in Figure 2, and would have liked to have seen
 accompanying quantitative results. If such results were to assert the
 significant improvement possible when including a complementary
 modality (i.e., without bias), that would be a very nice contribution
 in itself.}
\newline
\newline
We agree with the reviewer that establishing the benefit of
multivariate segmentation in the presence of bias would be
significant.  {\color{black}{Such exploration would probably lengthen
the current manuscript beyond its intended scope.  However, we did 
want to provide suggestive results with the single data set that 
we currently have which will obviously segue into a possible future
publication focusing on multivariate segmentation with Atropos
on specific data sets for which multi-modal data sets are essential.}}
\newline
\newline
{\em I have only two concerns with the article. First, there is almost
  no explanation of any of the figures in the body of the text, which
  raises questions about the significance of the findings, such as
  those of Figure 4 (``The results show that the PriorProbabilityMaps
  with w = 0.5 (far right) gives the best performance for all
  tissues.''). }
\newline
\newline
The revision adds explanatory material in the body of the text
associated with each figure.  {\color{black}{These revisions are printed in blue font and are sprinkled throughout the manuscript.}}
\newline
\newline
{\em Second, why is it that ``In this study, we labeled the brain web (sic) template very sparsely, quickly and crudely and did not expect highly accurate results.''
The results depending on these manual prior labels aren't informative
-- they don't give good results, and don't tell us whether the results
would be better if better manual labels were used.}
\newline
\newline
We agree with the reviewer and removed this comment and the evaluation
results.  We added (bottom of page 12): 
    \begin{quote}
    We provide this capability to allow the user to implement
    an interactive editing and segmentation loop.  The user may
    run Atropos with sparse manual label guidance, evaluate the
    results, update the manual labels and repeat until achieving the desired
    outcome.  This processing loop be performed easily with, e.g., ITK-SNAP.
    \end{quote}
{\em On page 22: ``The advancements introduced with N4 permit such an
  adaptive integration with Atropos.'' What relevant advances are
  introduced with respect to N3?}
\newline
\newline
{\color{black}{We added the following elaboration on the benefits of N4 in section 3.4.3 (page 21):
\begin{quote}
N4 extends the popular
nonparametric nonuniform intensity normalization (N3) algorithm [63]
in two principal ways:
\begin{itemize}
  \item We replace the least squares B-spline fitting with a parallelizable alternative 
  (which we also made publicly available in the Insight Toolkit)---
  the advantages being that 1) computation is much faster 
  and 2) smoothing is not susceptible to outliers as is characteristic with
  standard least squares fitting algorithms.
  \item The fitting algorithm permits a multi-resolution approach so whereas standard 
  N3 practice is to select a single resolution at which bias correction occurs, the N4
  framework permits a multi-resolution correction where a base resolution is chosen and
  correction can then occur at multiple resolution levels each resolution being twice 
  the resolution of the previous level.
\end{itemize}  
Specifically, with respect to segmentation, there exists a third advantage with N4 over
N3 in that the former permits the specification of a probabilistic mask as opposed 
to a binary mask.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 [64].  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.
\end{quote} }}
{\em Equations: Please remove punctuation following equations, because it came sometimes be distracting. For example, on pg.9, eq.1's comma might also appear to be a prime/apostrophe for the variable y.}
\newline
\newline
Done except in cases where the equation ends a sentence.   
\newline
\newline
Again, we thank each of the reviewers and the editor for their
efforts.  We truly feel that their reading of the paper was excellent
and greatly improved the work.

% \bibliographystyle{elsarticle-harv.bst}
% \bibliography{ftdcbdlong,avantsAll}

\end{document}

### 2011-02-09 doc/MVSeg.tex
ENH: an example figure
ENH: Adding ants repository from February 9, 2011

@@
 
+\begin{figure}
+\begin{center}
+\begin{tabular}{c}
+\includegraphics[width=6in]{Figures/004_vec_vs_dti.pdf}
+\end{tabular}
+\caption{The tensor-based segmentation on the left and the
+  RGB vector segmentation is on the right.  Both use Mahalanobis
+  distance and EM optimization with spatial priors.}
+\label{fig:example1}
+\end{center}
+\end{figure}
+
 \subsection{Neonate T1 and T2} There is also an interesting neonatal
@@
 \section{Methods} Our goal will be to provide reproducible
-measurements derived from DTI segmentation.
+measurements derived from multivariate segmentation.
 \begin{description}

### 2011-02-09 src/ANTS/Documentation/Tex/metric.tex
ENH: Adding ants repository from February 9, 2011

ANTS supports both volumetric registration and point set registration. The image / point set similarity metrics in ANTS are unified in the form of a function on the images or the point sets:       $$\textbf{Simlarity}[\textbf{fixedImage}, \textbf{movingImage}, \textbf{weight}, \textbf{parameters}].$$

The similarity type for the deformation transformation is specified by
the \verb -m  option, which contains two parts: simarity type and the parameters inside the brackets. (Note: no white spaces exist between parameters.) The possible similarity metrics for volumetric images are: 
\begin{itemize}
    \item Cross correlation estimate:  \verb " -m  CC[fixedImage,movingImage,weight,radius]". This metric works
      well for intra-modality and some inter-modality image registration. For exxample, \verb"-m CC[fixed.nii,moving.nii,1,5]" specifies:
    \begin{itemize}
        \item the fixed image: fixed.nii
        \item the moving image: moving.nii             
        \item weight for this metric is 1 (i.e. only this metric drives the registration).      
        \item the region radius for computing cross correlation is 5
    \end{itemize}

    \item Mutual information: \verb"-m MI[fixedImage,movingImage,weight,number-of-histogram-bins]". This metric works both well for intra-modality and inter-modality image registration. For example, the first three parameters in \verb"-m MI[fixed.nii,moving.nii,1,32]" is similar to the example above in cross correlation, except that the last parameter means that the number of bins in computing mutual information is 32.

    \item PR: \verb"-m PR[fixedImage,movingImage,weight,radius]". This
      metric is deprecated.  It still ``works'' but CC is preferred.

    \item Mean square difference: \verb"-m MSQ[fixedImage,movingImage,weight,0]". This metric works for intra-modality image registration. The last parameter \verb"0" is a padding value of no real meaning. For example, \verb"-m MSQ[fixed.nii,moving.nii,1,0]".

\end{itemize}

ANTS also support registration of point sets. The supported formats for point sets can be found in I/O section. The similarity metrics for point sets are:
\begin{itemize}
    \item Point set expectation: 
     \begin{verbatim}
-m PSE [fixedImage,movingImage,fixedPoints,movingPoints,weight,
pointSetPercentage,pointSetSigma,boundaryPointsOnly,
kNeighborhood,PartialMatchingIterations=100000]
    \end{verbatim}
    \begin{itemize}
        \item \verb"fixedImage": defines the space domain of the fixed point set.
        \item \verb"movingImage": defines the space domain of the moving point set.
        \item \verb"fixedPoints/Image": defines the coordinates of the fixed point set or label image. It can be an image with discrete positive labels, a VTK format point set file, or a text file. Details can be found in I/O section (TODO).
        \item \verb"movingPoints/Image": defines the coordinates of the moving point set or label image. 
        \item \verb"weight": weight for this metric. \verb"1" means that only this metric drives the registration.
        \item \verb"pointSetPercentage": the percentage of points to be randomly sampled used in the registration.
        \item \verb"pointSetSigma": the standard deviation of the Parzen window used to estimate the expectation. 
        \item \verb"boundaryPointsOnly":  \verb"1" (or ``true'') means only the boundary points in the label image is used to drive registration.
        \item \verb"kNeighborhood" is a positive discrete number. The first $k$ neighbors are used to compute the deformation during the registration. 
        \item \verb"PartialMatchingIterations" controls the symmtry in the matching. This option assumes the complete labeling is in the first set of label parameters ... more iterations leads to more symmetry in the matching  - 0 iterations means full asymmetry 
    \end{itemize}

    \item Jensen-Tsallis BSpline
    \begin{verbatim}
-m JTB[fixedImage,movingImage,fixedPoints,movingPoints,weight,
pointSetPercentage,pointSetSigma,boundaryPointsOnly,kNeighborhood,alpha,
meshResolution,splineOrder,numberOfLevels,useAnisotropicCovariances] 
    \end{verbatim}
    \begin{itemize}
        \item \verb"fixedImage": defines the space domain of the fixed point set.
        \item \verb"movingImage": defines the space domain of the moving point set.
        \item \verb"fixedPoints": defines the coordinates of the fixed point set. It can be an image with discrete positive labels, a VTK format point set file, or a text file. Details can be found in I/O section (TODO).
        \item \verb"movingPoints": defines the coordinates of the moving point set.
        \item \verb"weight": weight for this metric. \verb"1" means that only this metric drives the registration.
        \item \verb"pointSetPercentage": the percentage of points to be randomly sampled used in the registration.
        \item \verb"pointSetSigma": the sigma for the Parzen window used to estimate probabilities.  
        \item \verb"boundaryPointsOnly": [TODO] \verb"1" (or ``true'') means only the boundary points in the point sets are used to drive registration.
        \item \verb"kNeighborhood" is a positive discrete number. The first $k$ neighbors are used to compute the deformation during the registration. 
        \item \verb"alpha"
        \item \verb"meshResolution"
        \item \verb"splineOrder"
        \item \verb"numberOfLevels"
        \item \verb"useAnisotropicCovariances"
    \end{itemize}

\end{itemize}

In current implementation, the affine registration only supports two types of similarity metrics on volumetric images, which are specified using \verb"--affine-metric-type":
\begin{itemize}
    \item Mutual information, specified as \verb"--affine-metric-type MI". This usually works for both inter and intra-modalities in 3D. Also, the options used in computing mututal information in affine registration can be specified as \verb"--MI-option N1xN2". The first parameter \verb"N1" specifies the number of bins. The second parameter \verb"N2" specifies the number of samples. For example: \verb"--MI-option 32x8000".

    \item Mean square difference, specified using \verb"--affine-metric-type MSE". There is no options for this metric.
\end{itemize}

Fig. \ref{fig:metric_example} shows the registration result using intensity difference with the following command:
\begin{verbatim}
ANTS 2 -m MSQ[r16slice.nii,r64slice.nii,1,0] -r Gauss[3,0] -t SyN[0.5] -i 50x50x30
\end{verbatim}

%        \begin{wrapfigure}{r}{0.5\textwidth}
\begin{figure}
    \label{fig:metric_example}
    \centering
    \begin{tabular}[h]{c|c|c}
%        \includegraphics[width=0.16\textwidth]{Figures/r16slice.jpg} &
%        \includegraphics[width=0.16\textwidth]{Figures/r64slice.jpg} &
 %       \includegraphics[width=0.16\textwidth]{Figures/resMSQ.jpg} \\
        fixed image &
        moving image & 
        MSE  \\
    \end{tabular} 
    \itkcaption{registration using mean square intensity difference}
\end{figure}
    %    \end{wrapfigure}
Here is also an example script to register a pair of images using mean square intensity difference and computing the metrics of the registration image.
\begin{verbatim}
#use intensity difference with radius 0 -- radius no effect on intensity difference
ANTS 2 -m MSQ[r16slice.nii,r64slice.nii,1,0] -r Gauss[3,0] -t SyN[0.5] -i 50x50x30
WarpImageMultiTransform 2 r64slice.nii resMSQ.nii Warp.nii Affine.txt
MeasureImageSimilarity 2 0 r16slice.nii resMSQ.nii metricexamplelog.txt
MeasureImageSimilarity 2 1 r16slice.nii resMSQ.nii metricexamplelog.txt
MeasureImageSimilarity 2 2 r16slice.nii resMSQ.nii metricexamplelog.txt
ConvertToJpg resMSQ.nii resMSQ.jpg
\end{verbatim}

### 2011-02-09 src/ANTS/Documentation/atropos/_region_.tex
ENH: Adding ants repository from February 9, 2011

\message{ !name(atropos.tex)}%\documentclass[11pt,english]{article}
%
%% Set page margins correctly
%\usepackage{geometry}
%%\geometry{letterpaper,top=0.5in,left=0.5in,bottom=0.5in,top=0.5in,headsep=6pt,footskip=18pt}
%\geometry{letterpaper,top=0.5in,left=0.5in,bottom=0.5in,top=0.5in,headsep=6pt,footskip=18pt}
%
%% Use fancy header style
%\usepackage{fancyhdr}
%\pagestyle{fancy}
%\renewcommand{\headrulewidth}{0.75pt}
%\renewcommand{\footrulewidth}{0.75pt}
%\lhead{\footnotesize Principal Investigator/Program Director (Last, First, Middle): }
%\rhead{Yushkevich, Paul A.} \lfoot{\footnotesize PHS 398/2590 (Rev. 04/06)} \rfoot{\footnotesize
%\textbf Continuation Format Page} \cfoot{\footnotesize Page \underline{~~\thepage~~}}
%
%% Use pslatex fonts
%\usepackage{pslatex}
%\renewcommand{\familydefault}{\sfdefault}
%\renewcommand{\baselinestretch}{.9}
%

% START JUNK
\documentclass[11pt,english]{article}

% Set page margins correctly
\usepackage{geometry}
\usepackage{url}
\geometry{letterpaper,top=1.0in,left=1.0in,bottom=1.0in,top=1.0in,headsep=6pt,footskip=18pt}

\usepackage{lscape}
\usepackage[round,semicolon,authoryear,]{natbib}

% Use fancy header style
\usepackage{fancyhdr}
\pagestyle{empty}
\renewcommand{\headrulewidth}{0.75pt}
\renewcommand{\footrulewidth}{0.75pt}
%\lhead{\footnotesize Principal Investigator/Program Director (Last, First, Middle): }
%\rhead{Gee, James C.} \lfoot{\footnotesize PHS398 (06/07)} \rfoot{\footnotesize
%\textbf Continuation Format Page} \cfoot{\footnotesize Page \underline{~~\thepage~~}}

\usepackage{setspace}
\usepackage{listings}
\usepackage{float}

\floatstyle{plain}
\newfloat{command}{thp}{lop}
\floatname{command}{Command}

% \doublespacing

% Use pslatex fonts
%\usepackage[T1]{fontenc}
%\usepackage{mathptmx}

% END JUNK

% All other packages
\usepackage{boxedminipage}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{babel,verbatim}
\usepackage{enumerate}
%\usepackage{dsfont}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{sectsty}
\usepackage[compact]{titlesec}
\usepackage[usenames]{color}
\usepackage{ulem}
\usepackage{multirow,booktabs,ctable,array}

%GATHER{../../../shared/bibtex/biblio.bib}
%GATHER{../../../shared/bibtex/brainmri.bib}
%GATHER{../../../shared/bibtex/registration.bib}
\graphicspath{{./Figures/}
                          }

%\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}

\long\def\symbolfootnote[#1]#2{\begingroup%
\def\thefootnote{\fnsymbol{footnote}}\footnote[#1]{#2}\endgroup}

    \usepackage{color}

    \definecolor{listcomment}{rgb}{0.0,0.5,0.0}
    \definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
    \definecolor{listnumbers}{gray}{0.65}
    \definecolor{listlightgray}{gray}{0.955}
    \definecolor{listwhite}{gray}{1.0}

\newcommand{\lstsetcpp}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
       numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={\small \baselineskip 12pt Atropos short command line menu which is invoked using the `{\ttfamily -h}' option. 
        The expanded menu, which provides additional details regarding the possible parameters and usage 
        options, is elicited using the `{\ttfamily {-}{-}help}' option.},
        captionpos=b,
        label=listing:command
        }
}

\newcommand{\lstsetcppnfour}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={\small \baselineskip 12pt N4 short command line menu which is invoked using the `{\ttfamily -h}' option. 
        The expanded menu, which provides additional details regarding the possible parameters and usage 
        options, is elicited using the `{\ttfamily {-}{-}help}' option.  },
        captionpos=b,
        label=listing:n4
        }
}

%\renewcommand{\topfraction}{0.85}
%\renewcommand{\textfraction}{0.1}
%\renewcommand{\floatpagefraction}{0.75}

% Different font in captions
%\newcommand{\captionfonts}{\small}
%\makeatletter  % Allow the use of @ in command names
%\long\def\@makecaption#1#2{%
%  \vskip\abovecaptionskip
%  \sbox\@tempboxa{{\captionfonts #1: #2}}%
%  \ifdim \wd\@tempboxa >\hsize
%    {\captionfonts #1: #2\par}
%  \else
%    \hbox to\hsize{\hfil\box\@tempboxa\hfil}%
%  \fi
%  \vskip\belowcaptionskip}
%\makeatother   % Cancel the effect of \makeatletter

%Automated Segmentation of Ventilation Defects on $^3$He Lung MRI}
%\date{}
%\author{ Nicholas J. Tustison, DSc,$^{1*}$ Talissa A. Altes, MD,$^2$ Eduard E. de Lange, MD,$^2$
%  Brian B. Avants, PhD, $^1$ John P. Mugler III, PhD,$^2$ and James C. Gee, PhD,$^1$ \\
%  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
%  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
%\address{3600 Market Street, Suite 370 \\
%                 Philadelphia, PA  19104\\
%                 tustison@picsl.upenn.edu}
%\advisor{James Gee}
\begin{document}

\message{ !name(atropos.tex) !offset(-3) }

\normalem

\vspace*{5cm}

\begin{center}
{\Large \bf An Open Source Framework for Multivariate $n$-Tissue
  Segmentation and Evaluation Using Open Data Sets} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Jue Wu$^1$, Philip A.~Cook$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

\vfill

\begin{singlespace} 
\scriptsize
\flushleft
%\line(1, 0){250} \\
{\bf Atropos:  $n$-Tissue Segmentation}\\
Corresponding author: \\
Brian B. Avants\\
3600 Market Street, Suite 370\\
Philadelphia, PA  19104\\
avants@picsl.upenn.edu\\
\end{singlespace}

%\clearpage
%
%\vspace*{7cm}
%
%\begin{center}
%{\Large \bf An Open Source Framework for Multivariate $n$-Tissue Segmentation and Volumetric Cortical Parcellation:  Evaluation Using the BrainWeb, NIREP, and LPBA Data Sets}\\
%\end{center}

\clearpage

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation tool based on either parametric or non-parametric finite mixture modeling (FMM) of the
class intensities.  Atropos, available in ANTs
\footnote{
http://www.picsl.upenn.edu/ANTs
}, is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 69 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 69 class EM segmentation
problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that exercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

%\begin{keyword}
%segmentation \sep expectation maximization \sep spatial prior \sep brain 
%\end{keyword}

\clearpage

\section{Introduction} As medical image acquisition technology has advanced,
concomitant investment has been made towards adapting classification
techniques for neuroanatomy.  Some of the earliest work appropriated NASA
satellite image processing software for statistical classification of
head tissues in 2-D MR images \citep{Vannier1985}.  A proliferation of
techniques ensued with increasing sophistication in both core
methodology and degree of refinement for specific problems.  The
chronology of progress in segmentation may be tracked through both
technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid of the primate brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  Given the ``missing data'' aspect of this
problem, use of the Expectation-Maximization (EM) framework is natural\citep{Dempster1977}.  The
work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended nonparametric
probability model, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

Subsequent development included the use of Markov Random Field (MRF)  modeling \citep{Geman1984}
to regularize the classification results \citep{Held1997} with later work adding heuristics concerning neuroanatomy to prevent 
over-regularization and the resulting loss of fine structural details \citep{Leemput1999,Leemput1999a}.  
A more formalized integration of generic MRF spatial priors was employed in the work of \cite{Zhang2001}, 
commonly referred to as FAST (FMRIB's Automated Segmentation Tool), which is in widespread use
given its public availability and good performance.  More recently, a uniform distribution of local MRFs within the brain volume and their subsequent integration into a global solution has been proposed obviating the need for an 
explicit bias correction solution \citep{Scherrer2009}.

EM may easily be caught in local optima. Consequently, a good
initialization is critical.  Common low-level initialization steps 
include uniform probability assignment \citep{Wells1996},
Otsu thresholding \citep{Zhang2001}, and K-means clustering
\citep{Pappas1992}.  More sophisticated low-level initialization
schemes include that of \cite{Greenspan2006} in which a dense spatial
distribution of Gaussians is used to capture the complex
neuroanatomical layout with subsequent processing used to conjoin
subsets of Gaussians belonging to the same tissue classes.  Recently,
anatomically specific strategies for incorporating prior knowledge
into the segmentation solution have included the use of spatial prior
probability maps of the various structures of interest
\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
probability maps can also be used to provide an initial segmentation.
Related technological developments model partial volume effects for
increased accuracy in brain segmentation
\citep{Ruan2000,Ballester2002,Leemput2003}.

A general trend towards more integrative neuroanatomical image processing led to the work described in
\cite{Ashburner2005} which is publicly available within SPM5, a
large-scale Matlab module in which registration, segmentation, and
bias field correction can be simultaneously modeled within a single
optimization scheme. The roots of this very popular software package
stem back to early work by Karl Friston in which the basis for
statistical parametric mapping was developed \citep{Friston1990} and
subsequently provided to the research community.  Similar integrative brain processing was provided in \cite{Pohl2006} in which
segmentation and registration parameters were optimized simultaneously
while casting the inhomogeneity model parameters of \cite{Wells1996}
as nuisance variables.  Continued work involved recursive parcellation
of the brain volume by considering sub-structures in a hierarchical
manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
\citep{Pieper2006}---an open source package with developmental
contributions from multiple agencies including both private and
academic institutions. Researchers in aging often focus on
accurately segmenting the T1 MRI of elderly controls and subjects
suffering from neurodegeneration, for instance, via SIENA \citep{Smith2007}.  A recent evaluation study 
compared kNN segmentation, SPM Unified Segmentation and SIENA and
found different performance
characteristics under different evaluation criteria
\citep{Bresser2011}.  \cite{Klauschen2009} had similar findings when
comparing SPM5, FSL and FreeSurfer.  These studies suggest that no
single method performs best under every measurement and, along with the No
Free Lunch theorem, highlight the need for segmentation tools that are tunable for different
problems and research goals.

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} with aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
time-consumption and tedium associated with manual effort towards the
latter, various techniques have been proposed of which a small subset
has been publicly availed to the research community such as the
popular software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
Advantages of surface representations include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, as well as Klein's
Mindboggle \cite{Klein2005}, would usually follow an initial segmentation by a volumetric method such as Atropos.

%{\color{red}{{\em Needs work}:  Mindboggle, Heckemann  One advantage
%of the pipeline we propose is that we parcellate cortex entirely in
%the image space, thus avoiding the difficulty of transferring labels
%from the mesh space back to the image space---a problem that is a confound of surface-based methods\citep{Klein2010}.}}

Our open source segmentation tool, which we have dubbed {\em
Atropos}, 
\footnote{ Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs toolkit, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements a $n$-tissue paradigm for
voxel-based image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid.
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
also borrow an idea from \cite{Boykov2004} and use sparse spatial
priors to provide initialization and boundary conditions for Atropos
EM segmentation in a semi-interactive manner.  In short, Atropos seeks
to provide a segmentation toolbox that may be modified,
tuned and refined by the user for different problems.

We also allow Atropos to address the bias field issue by incorporating
the recently developed N4 bias correction software \citep{Tustison2010} in an adaptive manner.  Coupled with the
registration \citep{Avants2010b} and template building 
\citep{Avants2010} already included in the ANTs toolkit, Atropos is a
versatile and powerful software tool which touches multiple aspects of
our brain processing pipeline from brain extraction
\citep{Avants2010a} to grey matter/white matter/cerebrospinal fluid
segmentation to label fusion and propagation for achieving a refined
cortical parcellation based on both prior information and
subject-specific image data.

Atropos is multi-platform, free, stand-alone and able to
manage a wide range of scalar and vector image datatypes.  To highlight the
value of this open-source contribution, we performed a search of
software attributes on NITRC and found that as of November 2010 no
stand-alone EM methods are currently listed.
Atropos's memory efficiency enables EM-segmentation of many classes at
once, a capability that is due to a sparse internal representation of
prior probability images that minimizes the need for random access
memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 69 neuroanatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  Atropos
also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
segmentation, as in neonatal brain tissue classification
\cite{Weisenfeld2009}.  Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

Organization of this work is as follows: we first describe the theory
behind the various components of Atropos while
acknowledging that theoretical discussion is available elsewhere,
e.g. in previously cited literature.  This is followed by a thorough
discussion of implementation which, though often overlooked, is of
immense practical utility.  We then report evaluation results on
public T1 neuroimaging datasets that have reference gold/silver
standard labelings.  Finally, we provide a discussion of our results
and our open-source contribution in the context of the remainder of
this paper and of previous and future work.

\section{Theoretical Foundations for Atropos Segmentation}

Atropos encodes a family of Bayesian segmentation techniques that may be configured in an application-specific manner.  The theory underlying Atropos dates back 20$+$ years and is representative of some of the most innovative work in the field.  Although we synthesize some of the theoretical work in this section, we recommend to the interested reader consultation with the literature cited previously for further insight.

Bayes' theorem provides a powerful mechanism for making inductive
inferences assuming the availability of quantities defining the
relevant conditional probabilities, specifically the likelihood and
prior probability terms.  Bayesian paradigms for brain image
segmentation employ a user-selected observation model defining the
likelihood term and one or more prior probability terms.  The product
of likelihood(s) and prior(s) is proportional to the posterior
probability.  The likelihood term has been previously defined both 
parametrically (e.g. a Gaussian model) and non-parametrically
(e.g. Parzen windowing of the sample histogram).  The prior term, as
given in the literature, has often been formed either as
MRF-based or template-based.  An image segmentation solution in this
context is an assignment of one label to each voxel
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  The next sections
introduce notation and provide a formal description of three essential
components in Bayesian segmentation, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized
    MRF and template-based prior terms, and 
%, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These components are common across most EM segmentation algorithms.

\subsection{Notation}
Assume a field, $\mathcal{F}$, whose values are known at discrete
locations, i.e. sites, within a regular voxel lattice that makes up an
image domain, $\mathcal{I}$.  Note that $\mathcal{F}$ can be a scalar
field in the case of unimodal data (e.g. T1 image only) or a vector
field in the case of multimodal data (e.g. T1, T2, and proton density
images).  A specific set of observed values, denoted by $\mathbf{y}$,
are indexed at $N$ discrete locations in $\mathcal{I}$ by $i \in \{1,
2, \ldots, N\}$.  This random field, $Y = \{y_1, y_2, \ldots, y_N \}$,
serves as a discrete representation of an observed image's intensities.  A labeling
of this image, also known as a hard segmentation, assigns to each site
in $\mathcal{I}$ one of $K$ labels from the finite set $\mathcal{L} =
\{l_1, l_2, \ldots, l_K\}$.  Also considered a random field, this
discrete labeling is $X = \{x_1, x_2, \ldots, x_N\}$ where each
$x_i \in \mathcal{L}$.   We use $\mathbf{x}$ to denote a specific set of labels in
$\mathcal{I}$ and a valid, though not necessarily optimal, solution to the segmentation problem.

%Furthermore, it is assumed that $\mathcal{F}$ is a random field whose true state, $\mathcal{S}$, is `hidden' beneath its observed values but fully represented by $\mathcal{L}$.%
%\footnote{
%For example, in the classic 3-tissue segmentation problem $\mathcal{L}$ is defined as the set $ \{\mathrm{csf}, \mathrm{gm}, \mathrm{wm}\}$ and each voxel in the brain region is assigned one and only one of these three labels.  
%}

\subsection{Segmentation Objective Function} Atropos optimizes a class
of user-selectable maximization objective functions each of which may
be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
requires {\em likelihood models} and {\em prior models} which enter
into Bayes' formula,
\begin{equation}\label{eq:bayes}
p(\mathbf{x}|\mathbf{y})=\underbrace{
p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{\mathbf{y}}, 
\end{equation} 
where the normalization term, $1/\mathbf{y}$, is a constant that does
not affect the optimization \citep{Sanjay-Gopal1998}.
Given choices for likelihood models and prior probabilities, the Bayesian
segmentation solution is the labeling $\hat{\mathbf{x}}$ which
maximizes the posterior probability, i.e.
\begin{align} \hat{\mathbf{x}} = \argmax_{\mathbf{x}}
\left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align} Similar to its predecessors, Atropos employs
the EM framework \citep{Dempster1977} to find maximum likelihood
solutions to this problem.  The following sections detail the 
Atropos EM along with choices for the likelihood(s), prior(s).

\subsection{Likelihood or Observation Models}
To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normal distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  When $y_i$ is a vector
quantity, we replace the Euclidean distance by by Mahalanobis distance
and define multivariate Gaussian parameters via a mean vector,
$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma_k}$.

A common technique for the non-parametric variant is to define $p_k$ using Parzen windowing of the sample observation histogram of $\mathbf{y}$, i.e.
\begin{align} \label{eq:nonparam}
  p_k\left(y_i|l_k\right) &= \frac{1}{N_B} \sum_{j=1}^{N_B} \frac{1}{\sqrt{2\pi \sigma_j^2}}\exp\left( \frac{ -(y_i - c_j)^2 }{2\sigma_j^2} \right)
\end{align}
where $N_B$ is the number of bins used to define the histogram of the sample observations (in Atropos the default is $N_B = 32$) and $c_j$ is the center of $j^{th}$ bin in the histogram.  $\sigma_j$ is the width of each of the $N_B$ Gaussian kernels.  For multi-modal data in which the number of components of $y_i$ is greater than one, a Parzen window function is constructed for each component.  The likelihood value is determined by the joint probability given by their product.

Atropos segmentation likelihood estimates are based on the classical finite mixture model (FMM).
FMM assumes independency between voxels to calculate the probability
associated with the entire set of observations, $\mathbf{y}$.  Spatial
interdependency between voxels is modeled by the prior probabilities
discussed in the next section.  Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
\begin{align}\label{eq:likelihood}
  p(\mathbf{y}|\mathbf{x}) = \prod_{i=1}^N \left(      
                                                        \sum_{k=1}^K \gamma_k p_k(y_i|l_k)
                                                        \right)
\end{align}
where $\gamma_k$ is the mixing parameter \citep{Ashburner2005}.

\subsection{Prior Probability Models}
By modeling $\mathcal{F}$ via the set of observation models $\Phi$,
this so called finite-mixture model could be used to produce a
labeling or segmentation \citep[e.g.][]{Wells1996}.  However, as
pointed out by \cite{Zhang2001}, exclusive use of the intensity profile produces a less than optimal
solution because spatial contextual considerations are ignored.  This
has been remedied by the introduction of a host of prior probability
models including those characterized by use of MRF theory and
template-based information.  For example, in the works of
\cite{Leemput1999a} and \cite{Weisenfeld2009}, the original global
prior term given in \cite{Wells1996} is replaced by the product of the
template-based and the MRF-based prior terms.  In addition to their
descriptions below, we discuss a third possible
prior/objective combination in the form of a (sparse) prior labeling which fixes specific points
of the segmentation and uses EM to propagate this information
elsewhere in the image.  
% extension of template-based Euclidean and geodesic distance label propagation.

\subsubsection{Generalized MRF Prior}
One may incorporate spatial coherence into the segmentation by
favoring labeling configurations in which voxel neighborhoods tend
towards homogeneity. This intuition is formally described by MRF
theory in which spatial interactions in voxel neighborhoods can be
modeled \citep{Li2001}.

We assume the random field introduced earlier, $X$, is an MRF characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., 
 \begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x} 
 \end{align}
and where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
other words, any labeling permutation on $X$ is {\em a priori}
possible).  The MRF locality condition is then, 
\begin{align}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right),
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This locality property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Following these two assumptions, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
\end{align}
$Z$ is a normalization factor known as the {\em partition function}
and $U(\mathbf{x})$ is the {\em energy function} which can take
several forms \citep{Li2001}. In Atropos, as is the case with many
other segmentation algorithms of the same family, we choose
$U(\mathbf{x})$ such that it is only composed of a sum over pairwise
interactions between neighboring sites across the image,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta,
$\delta_{ij}$, based on the classical Ising potential (also known as a Potts model) \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            1 & \text{if } x_i = x_j \\
                            0 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more heavily than sites further away.

\subsubsection{Template-Based Priors}
A number of researchers have used templates to both
ensure spatial coherence and incorporate prior knowledge in
segmentation.  A common technique is to select labeled subjects from a
population from which a template is constructed \citep[e.g.][which is
also available in the ANTs toolkit]{Avants2010}.  Each labeling can
then be warped to the template where the synthesis of warped labeled
regions produces a prior probability map or prior label map encoding
the spatial distribution of labeled anatomy which can be harnessed in
joint segmentation/registration or, specifically, Atropos/ANTs hybrids involving unlabeled subjects.

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$
(cf. Eqn. (\ref{eq:likelihood})), 
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions,
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}.
\end{align}
The $t_{ik}$ is the prior probability value at site $i$ which was
mapped, typically by image registration, to the local image
from a template data set.  The user may also
choose mixing proportions equal to
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}}.
\end{align}
via the command line interface to the posterior formulation.

\subsubsection{Supervised Semi-Interactive Segmentation}
Brain segmentation methods have relied on user interaction for many
years \citep{Lim1989,Julin1997,Freeborough1997a,Yushkevich2006}.
Atropos is capable of benefitting from user knowledge via an
initialization and optimization that depends upon a spatially
varying prior label image passed as input.  Rapid, sparse
labeling---with visualization provided by ITK-SNAP
(www.itksnap.org)---enables an interaction and execution processing
loop that can be critical to solving segmentation problems with
challenging clinical data in which automated approaches fail.  This
part of Atropos design is inspired by the interactive graph cuts
pioneered by \cite{Boykov2001} and which has spawned many follow-up
applications.  The Atropos prior label
image pre-specificies the segmentation results at a subset of the
spatial domain by fixing the priors and likelihood (and, thus, the
posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
and $0$ for each other label at the same site.  The user-input
therefore not only initializes the optimization, but also gives boundary conditions
that influence the EM solution outside of the known sites.  While the
graph-based min-cut max-flow solution is globally optimal for two
labels, only locally optimal optimizers are available for 3 or more
classes.  Thus, in most practical applications, EM is a
reasonable and efficient alterative to Boykov's solution.
Furthermore, one may automate the
initialization process.

\begin{comment}
{
\subsubsection{Distance Prior for Label Propagation}

In order to provide a dense segmentation even in regions where the template-based prior probability is 0, we use the template-based prior label images or probability maps described previously to formulate an optional distance prior probability.
Consider a subset of $\mathcal{I}$, which we denote $R_k$, defined by $\mathrm{Pr}(x_i = l_k) > \mathrm{Pr}(x_i = l_j) \,\,\forall j \in K, j \neq k$ and $\mathrm{Pr}(x_i = l_k) > 0$. 
We also denote the boundary of $R_k$ as $\partial R_k$ and the `peak' as $\rho_{R_k}$ where $i$ and $j$ are sites in $R_k$:
\begin{align}
  \rho_{R_k} = \argmax_{i \in R_k} \left\{ \min_{j \in \partial {R_k}} d\left(i,j\right)\right\}.
\end{align}
In other words, the peak is an interior site (or multiple interior sites) in $R_k$ characterized as having the maximum distance of all sites to any corresponding closest point on the boundary of $R_k$.  We denote this max-min distance as $\Delta_k$.  This allows us to define the distance prior probability for label propagation as follows:
\begin{align}\label{eq:prop}
  \mathrm{Pr}(x_i = l_k) = \left\{
                                               \begin{array}{ll}
                                                \alpha_k \exp \left( -\frac{d(i, \partial R_k)}{\sigma_k} \right) & \text{if } i \text{ is outside the region } R_k \\
                                                \alpha_k + d\left(i, \partial R_k\right)\left(\frac{1.0 - \alpha_k}{\Delta_k}\right) & \text{otherwise}
                                                \end{array}
                                                \right.
\end{align}
The probability calculations are illustrated in Fig.~\ref{fig:distancePrior} and were designed so that the most interior point, or set of points, of region $R_k$ would have a distance prior probability of 1.0 which would linearly decrease within the region to the boundary with a user-selected probability value of $\alpha_k$.  Outside $R_k$, the distance prior exponentially decays with a decay rate governed by another user-selected parameter $\sigma_k$.  This permits a dense segmentation throughout the segmentation region of $\mathcal{I}$ even where the template prior probabilities are zero.

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=150mm]{distancePrior.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Graphical representation of the distance prior probability calculation associated with label $l_k$.  This extension of the popular template-based prior probability permits a dense labeling with the user specified mask. }
\label{fig:distancePrior}
\end{center}
\end{figure}

A crucial matter, particularly for propagating labels in the cortical
regions, is the choice of distance function.  One of the advantages to
surface-based methods is that distances can be calculated along the
surface mesh, thus discriminating between proximal cortical surface
points and those points which are situated close together due to
cortical folding.  In Atropos, we provide two possibilities---(1) a
geodesic distance function using a fast marching construction
\citep{Osher1988} of the distance function with Atropos's masked
region of interest and (2) a Euclidean distance function using the algorithm described in \cite{Maurer2003}.  The differences in label propagation results are illustrated in Fig.~\ref{fig:label_propagation}.

Figure~\ref{fig:neonate} illustrates the benefit of geodesic distance
priors in segmenting T2 MRI of the 
neonatal brain.  Geodesic distance priors may be used to reduce ambiguity at
tissue interfaces, where partial voluming leads to mixed intensity.  
In neonatal brain segmentation from T2 MRI, the
interface of gray matter (low intensity) and cerebrospinal fluid
(bright intensity) often matches that of white matter (medium intensity). 
However, the white matter class should never appear near the surface
of the brain.  Thus, the distance from the brain surface may be used
to disambiguate voxels such as these.  
In this case, the distance of a voxel from a known
or expected tissue location serves to differentiate it from classes
that have similar intensity signature.  The geodesic distance is therefore an
alternative to explicit models of partial volume \citep{Ruan2000,Ballester2002,Leemput2003}.

\begin{figure}
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=50mm]{sparse_labels.png} &
\includegraphics[width=50mm]{prop_euclidean.png} &
\includegraphics[width=50mm]{prop_geodesic.png} \\
(a) & (b) & (c) \\
\end{tabular}
\caption{\baselineskip 12pt \small  Illustration of label propagation differences between the Euclidean and geodesic distances which
are particularly acute in sinuous structures such as the cortex.  (a) Given a mask, represented in white, and a sparse labeling,
Atropos can be used to propagate the labels in a dense manner throughout the masked region.  (b) This propagation can occur using Euclidean distances in the image space via an Euclidean distance transform which can cause labels to jump meaningful anatomical boundaries.   (c) Given the cortical geometry, a more intuitive approach would be to propagate the labelings in a geodesic manner using a product of the fast marching paradigm which can be used in a manner that respects anatomical boundaries.  }
\label{fig:label_propagation}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{sparse_labels.png} 
\caption{\baselineskip 12pt \small  Neonatal brain segmentation.  }
\label{fig:neonate}
\end{center}
\end{figure}

}
\end{comment}

\subsection{Optimization}
Atropos uses expectation maximization to find a locally optimal
solution for the user-selected version of the Bayesian segmentation
problem (cf. Eqn. (\ref{eq:bayes})).
After initial estimation of the
likelihood model parameters, EM iterates between calculation of the
missing optimal labels $\hat{\mathbf{x}}$ and subsequent re-estimation of the model
parameters by maximizing the expectation of the complete data
log-likelihood (cf. Eqn. (\ref{eq:likelihood})).  
%\begin{align}
%\log( p(\mathbf{y}) ) = \sum_{i=1}^N \log \left( \sum_{k=1}^K \gamma_k  p(\mathbf{y}|l_k)  \right)
%\end{align}
The expectation maximization procedure is derived in various
publications including \cite{Zhang2001} which yields the optimal mean
and variance (or covariance), but sets the mixing parameter $\gamma_k$
as a constant.  The Atropos implementation estimates $\gamma_k$ at
each iteration as in \cite{Ashburner2005}.
\footnote{
Due to the lack of parameters in the non-parametric approach, it is not technically an EM algorithm (as described in \cite{Wells1996}).  However, the same iterative maximization is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric models (see the Introduction).
}  
When spatial coherence constraints are included as an MRF prior in Atropos, the optimal segmentation solution becomes intractable.%
\footnote{
Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
}
Although many optimization techniques exist \citep[see the
introduction in][for a concise summary of the myriad optimization
possibilities]{Marroquin2002}---each with their characteristic
advantages and disadvantages in terms of computational complexity and
accuracy---Atropos uses the well-known {\em Iterated Conditional
  Modes} (ICM)  \citep{Besag1986} which is greedy, computationally
efficient and provides good performance.  The EM employed in
Atropos may therefore be written as a series of steps:
\begin{description}
\item[Initialization:] In all cases, the user defines the number of
classes to segment.  The simplest initialization is by the classic
K-means algorithm with only the number of classes as a prior.
Otherwise, the user must provide prior information for each class in
the form of either a single $n$-ary prior label image or a series of
prior probability images, one for each class.
\item[E-Step: ICM Label Update:] Given the initialization and fixed
model parameters, Atropos will evaluate the posterior probability for
each possible label and assign, to each voxel, the label that
maximizes the posterior.  {\bf \textcolor{red}{ FIXME more specific?}}.
\item[M-Step: Parameter Update:] 
The posteriors used in this step are called $p^{old}$ to make it
clear that we use the previous parameter estimates.
We use a common and elementary estimate of the mixing parameters:
\begin{align}
\gamma^{new}_k \leftarrow  \sum_i \frac{p^{old}_{ik}}{\sum_{j=1}^K p^{old}_{ij}}. 
\end{align}
For parametric model update, for each of
  $K$ labels, we compute the mean,
$$ \mu^{new}_k \leftarrow \frac{ \sum_{i=1}^N  y_i p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) },
$$
and variance {\bf \textcolor{red}{ FIXME should this be the full
    covariance matrix?}},
$$ \sigma^{new}_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) }.
$$
This type of update is known as {\em soft} EM.  Hard EM, in
contrast, only uses sites containing label $l_k$ to update model $k$ parameters.
A similar pattern is used in the multivariate and non-parametric cases.
\end{description}
EM will iterate toward a local maximum.  We track
convergence by summing up the maximum posterior
probability at each site over the segmentation domain.  
%\begin{align}
%P(l_k | y_i, t_i, m_i,d_i) &= \frac{P(y_i|l_k ,t_i,m_i,d_i)   P(l_k|t_i,m_i,d_i)   }{P(y_i|t_i,m_i,d_i)}
%\end{align}
%
%
%An image, $I$, maps a domain, $\Omega$ into the
%positive real numbers, such that $I \colon \Omega \rightarrow
%\mathbb{R}^+$.  The goal of segmentation, in general, is to define the
%spatial distribution of a finite set of labels over this domain.  We
%denote the segmentation itself as $\eta \colon \Omega \rightarrow L$
%where $L = \{ L_1 = 1 , \cdots , L_N=N \}$, a set of integer indexed
%segmentation labels.  Note that $\eta$ may be formed as $\eta =
%\sum_{i=1}^{i=N} L_i \eta_i $ where $\eta_i$ is the binary
%segmentation for label $L_i$.  A prior estimate for the label image
%$\eta_i$ is here denoted $\eta^s_i$ with a complete set of priors
%denoted $\eta^s$.  The boundary of the binary segmentation -- where a
%$0/1$ transition edge exists -- is denoted $\partial \eta^s_i$, for
%the prior, and $\partial \eta_i$ for the label image.
%
%\subsection{Apocrita Theory} A general maximum a posteriori criterion
%for segmentation seeks,
%\begin{equation} {\hat \eta} = \argmax _{\eta} \Pr( \eta | I )(\x) =
%\Pr( I | \eta )(\x) \Pr(\eta)(\x),
%\end{equation} 
%where $I$ is the input image, ${\eta}$ represents the
%label set configuration taken from the set $L$, $\Pr$ is the
%probability, $\x$ is the spatial index and the optimal solution is
%$\hat \eta$.  The input image $I$, here, is an unlabeled T1 MRI
%indexed by the value $\x \in \Omega$ where $\Omega$ is the image's
%spatial domain.  This probability is composed of the likelihood (first
%term) and the prior (second term).  Atropos ~ uses a spatially varying
%likelihood term and a two-component prior term that takes into account
%both spatial distribution and label smoothness, the latter via a
%standard MRF prior.
%
%The likelihood term for a single label value $L_i \in L$ is,
%\begin{eqnarray} \Pr( I | \eta_i )(\x)=\frac{1}{Z_i} \exp(- \| I(\x) -
%\mu_i(\x) \|^2 / \sigma_i^2 ),
%\end{eqnarray} where $Z_i$ is a normalizing constant, $\mu(\x)$ is a
%spatially varying estimate of the tissue mean and $\sigma$ is a
%standard deviation.  The prior term is given by,
%\begin{eqnarray} \Pr(\eta_i)(\x) = \frac{1}{Z^\prime_i} \exp( -f( \x -
%\y_{\partial \eta^s_i} ) / \sigma_{\eta^s_i}^2 ) p(\eta_i |
%\eta^\aleph_i), \\ \notag f( \x - \y_{\partial \eta^s_i} ) = (1 -
%\eta^s_i (\x) ) \| \x - \y_{\partial \eta^s_i} \|,
%\end{eqnarray} where $p(\eta_i | \eta^N_i)$ is the MRF smoothness
%probability based on the local neighborhood $\aleph$, the $Z$ is a
%normalizing constant and $\y_{\partial \eta^s_i}$ is the nearest point
%to $\x$ on the boundary of this labeling. Atropos ~requires a user or
%template-defined $\eta^s_i$ and the standard deviation
%$\sigma_{\eta^s_i}$ if a non-unity spatial prior component is desired
%for that label.  The free parameters, that must be estimated
%iteratively, are therefore $\mu_i, \sigma_i$ and the label set itself
%$\hat \eta$, which defines the (locally) optimal spatial distribution
%of $L$ through $\Omega$.  Figure~\ref{fig:spatp} shows the
%distribution of the spatial prior as a function of the distance from
%prior-defined object boundary.  Note that $f$ may easily be varied for
%other applications or that fixed probability images may also be
%substituted here.  This choice of $f$ is motivated by the fact that it
%allows compressed storage of the priors in a single image, $\eta^s$,
%while also maintaining the ability to manipulate -- for each
%$\eta^s_i$ -- the spatial influence of the prior via
%$\sigma_{\eta^s_i}$.  Practically, this is especially valuable when,
%$N$, the number of labels, is large.

\section{Implementation}

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=160mm]{AtroposFlowchart.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Flowchart illustrating Atropos usage typically beginning with bias correction via N4.  
Initialization provides an estimate before the iterative optimization in which the 
likelihood models for each class are tabulated from the current estimate followed by a recalculation of the 
posterior probabilities associated with each class.  The multiple options associated with the different algorithmic 
components are indicated by the colored rounded rectangles connected to their respective core Atropos 
processes via curved dashed lines.  }
\label{fig:flowchart}
\end{center}
\end{figure}
\subsection{The Atropos User Interface}
As with other classes that comprise ANTs, Atropos uses the Insight
Toolkit as a developmental foundation.  This allows us to take
advantage of the mature portions of ITK (e.g. image IO) and ensures
the integrity of the ancillary processes such as those facilitated by
the underlying statistical framework.  Although Atropos is publicly
distributed with the rest of the ANTs package, we plan to contribute
the Atropos to the Insight Toolkit where it can be vetted and further
improved by other interested researchers.

An overview of the core components of Atropos can be gleaned, in part, from the flowchart depicted in 
Fig.~\ref{fig:flowchart}.  To provide a more intuitive interface without the overhead costs of a graphical user interface, a set of unique command line parsing classes were developed which can also provide insight to the functionality of Atropos.  
The short version of the command line help menu is given in Listing \ref{listing:command} which is invoked by typing `{\ttfamily Atropos -h}' at the command prompt.  Both short and long option flags are available and each option has its own set of possible values and parameters introduced in a more formal way in both previous discussion and related papers cited in the introduction.  Here we  describe these options from the unique perspective of  implementation.

\singlespacing 
\begin{command}
\lstsetcpp
\begin{lstlisting}
COMMAND: 
     Atropos

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -a, --intensity-image [intensityImage,<adaptiveSmoothingWeight>]
     -b, --bspline [<numberOfLevels=6>,<initialMeshResolution=1x1x...>,<splineOrder=3>]
     -i, --initialization Random[numberOfClasses]
                          KMeans[numberOfClasses]
                          Otsu[numberOfClasses]
                          PriorProbabilityImages[numberOfClasses,
                            fileSeriesFormat(index=1 to numberOfClasses) or vectorImage,
                            priorWeighting,<priorProbabilityThreshold>]
                          PriorLabelImage[numberOfClasses,labelImage,priorWeighting]
     -x, --mask-image maskImageFilename
     -c, --convergence [<numberOfIterations=5>,<convergenceThreshold=0.001>]
     -k, --likelihood-model Gaussian
                            HistogramParzenWindows[<sigma=1.0>,<numberOfBins=32>]
     -m, --mrf [<smoothingFactor=0.3>,<radius=1x1x...>]
     -o, --output [classifiedImage,<posteriorProbabilityImageFileNameFormat>]
     -u, --minimize-memory-usage (0)/1
     -w, --winsorize-outliers BoxPlot[<lowerPercentile=0.25>,<upperPercentile=0.75>,
                               <whiskerLength=1.5>]
                              GrubbsRosner[<significanceLevel=0.05>,<winsorizingLevel=0.10>]
     -e, --use-euclidean-distance (0)/1
     -l, --label-propagation whichLabel[sigma=0.0,<boundaryProbability=1.0>]
     -h 
     --help 
\end{lstlisting} 
\end{command}

\subsection{Initializing the Atropos Objective}
Atropos has a number of parameters defined within
Listing \ref{listing:n4} and will function on 2, 3 or 4 dimensional
data.  However, the majority of the time, users will be concerned with
a smaller set of input parameters.  Here, we list the recommended input and
an example definition for each parameter:
\begin{description}
\item[Input images to be segmented:] If more than
  one input image is passed, then a multivariate model will be
  instantiated.  E.g.  {\ttfamily -a Image.nii.gz} for one image and
  {\ttfamily -a Image1.nii.gz -a Image2.nii.gz} for multiple images.
\item[Input image mask:] This binary image
  will define the spatial segmentation domain.  Regions outside the
  mask receive the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
\item[Convergence criteria:] The algorithm terminates if it reaches max-iterations or
  produces a change less than the minimum change in the posterior.
  E.g. {\ttfamily -c [5,1.e-5]}. 
\item[MRF prior:] The key parameter to increase or decrease the
  spatial smoothness of the label map is $\beta$.  A useful range of
  $\beta$ values is $0$ to $0.5$ where we usually use $0.05, 0.1, 0.2$
  in brain segmentation.  E.g.  {\ttfamily -m [ 0.1, 1x1x1 ]} would
  define $\beta=0.1$ with a MRF radius of one voxel in each of three dimensions.
\item[Initialization:] The initialization options include (where the
  first parameter defines $K$, here 3 for each below),
\begin{itemize}
\item  {\ttfamily -i  kmeans[3]} standard kmeans initialization for
  three classes.
\item  {\ttfamily -i  PriorLabelImage[3, label\_image.nii.gz , w ]}
  where scalar $w \in [0,1]$ is unused. 
\item  {\ttfamily -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz ,
    w ]} where $w=0$~(use the prior probability images only for
  initialization) or $w=0.5$ (use the prior probability images
  throughout the optimization).  If one chooses $0 < w < 0.5$ then one
  will increase (from zero) the weight on the priors.  These images,
  like the PriorLabelImage, should be defined with the same domain as
  the input images to be segmented.  
\end{itemize}
\item[Posterior formulation:]  The user may choose to estimate the
  mixture proportions (or not) by setting  {\ttfamily -p Socrates[1]}
  or {\ttfamily -p Socrates[0] }.  Fixed label boundary conditions may be employed by
  selecting the PriorLabelImage initialization and   {\ttfamily -p Plato[0] }.
\item[Output:] Atropos will output the hard segmentation and the
  probability image for each model.  E.g.  {\ttfamily -o
    [segmentation.nii.gz , seg\_prob\%02d.nii.gz] } will write
  out the hard segmentation in the first output parameter and a
  probability image for each class named, here, seg\_prob01.nii.gz ,
  seg\_prob02.nii.gz, etc. 
\end{description}
Higher dimensions than 4 are possible although we have not yet encountered such 
application-specific need.  Multiple images (assumed to be of the same
dimension, size, origin, etc.), will automatically enable multivariate
likelihoods.  In that case, the first image specified on the 
command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
follow-up multivariate K-means labeling.    More details on each of
the key implementation options are given below.

\subsection{Likelihood Implementation}
As mentioned previously in the introduction, different groups have
opted for different likelihood models which have included either
parametric, specifically Gaussian, or nonparametric variations.
However, these approaches are similar in that they require a list
sample of intensity data from the input image(s) and a list of
weighting values for each observation of the list sample from which
the model is constructed.  In general, one may query model
probabilities by passing a given pixel's single intensity (for
univariate segmentation) or multiple intensities (for multivariate
segmentation) to the modeling function, regardless of whether the
function is parameter or non-parametric.  These similarities permitted
the creation of a generic plug-in architecture where classes
describing both parametric and nonparametric observational models are
all derived from an abstract list sample function class.  Three
likelihood classes have been developed, one parametric and two
nonparametric, and are available for usage although one of the
nonparametric classes is still in experimental development.  The
plug-in architecture even permits mixing likelihood models with
different classes during the same run for a hybrid
parametric/non-parametric model although this possibility has yet
to be fully explored.

If the Gaussian likelihood model is chosen, the list sample of
intensity values and corresponding weights comprised of the posterior
probabilities are used to estimate the Gaussian model parameters, i.e.
the mean and variance.  For the non-parametric model, the list sample
and posteriors are used in a Parzen windowing scheme on a weighted
histogram to estimate the observational model \citep{Awate2006}.

\subsection{Prior Probability Models}
Consistent with our previous discussion, we offer both an MRF-based
prior probability for modeling spatial coherence and the possibility
of specifying a set of prior probability maps or a prior label map
with the latter extendable to creating a dense labeling.  To invoke
the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing
factor (or the granularity parameter, $\beta$, given in
Eqn. (\ref{eq:U}), and the radius (in voxels) of the neighborhood
system using the vector notation `{\ttfamily 1x1x1}' for a
neighborhood radius of 1 in all 3 dimensions.  This radius is defined
such that voxels including but not limited to those that are face
connected will influence the MRF.

\begin{comment}{
The user also has an option to propagate initial labels, set by the
prior options, by a distance map tool.  
To propagate labels with the specification of prior probability images
or a prior label image, one needs to select the distance function (set
by the `{\ttfamily -e/--use-euclidean-distance}' boolean option)
which, by default, is determined by the geodesic distance.  The user
also needs to set the label propagation parameters for one or more of
the classes.   This  includes both the boundary probability and the exponential
 decay parameter, respectively $\alpha_k$ and $\sigma_k$ in Eqn. (\ref{eq:prop}).  
}\end{comment}

\subsection{Integration with Registration and N4 Bias Correction} 
Much of the image processing in our lab takes advantage of multiple
open source tools.  Most critically, these involve registration and
bias correction, in addition to segmentation. 
\subsubsection{Registration of Probability Maps} 
Image registration enables the ability to transform information
between spatial domains which may aid in both segmentation and bias
correction.  We rely heavily on template-building strategies
\citep{Avants2010,Avants2010a} which are also offered in ANTs.  Since
aligned prior probability images and prior label maps are often
associated with such templates, Atropos can be initialized with these
data with their influence regulated by a prior probability weighting
term.  Although prior label maps can be specified as a single
multi-label image, prior probability data are often represented as
multiple scalar images with a single image corresponding to a
particular label.  For relatively small classifications, such as the
standard 3-tissue segmentation (i.e. white matter, gray matter, and
cerebrospinal fluid), this does not typically present computational
complexities using modern hardware.  However, when considering dense
cortical parcellations where the number of labels can range upwards of
74 per hemisphere \citep{Destrieux2010}, the memory load can be
prohibitive if all label images are loaded into run-time memory
simultaneously.  A major part of minimizing memory usage in Atropos,
which corresponds to the boolean `{\ttfamily
-u/--minimize-memory-usage}' option, is the sparse representation of
each of the prior probability images.  Motivated by the observation
that these spatial prior probability maps tend to be highly localized
for large quantities of cortical labels, a threshold is specified on
the command line (default = 0.0) and only those probability values
which exceed that threshold are stored in the sparse representation.
During the course of optimization, the prior probability image for a
given label is reconstructed on the fly as needed.  For instance, the
NIREP (www.nirep.org) evaluation images are on the order of $300
\times 300 \times 256$ with 32 cortical labels.  Our novel memory
minimizing image representation typically shrinks run-time memory usage from a
peak of 10+ GB to approximately 1.5 GB and enable these datasets to be
used for training/prior-based cortical parcellation.

\singlespacing 
\begin{command}
\lstsetcppnfour
\begin{lstlisting}
COMMAND: 
     N4BiasFieldCorrection

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\subsubsection{Integration of N4 Bias Correction}
The typical segmentation processing pipeline begins with an intensity
normalization/bias correction step using the recently developed N4
algorithm \citep{Tustison2010} which is based on the popular
nonparametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
several methods proposed in the literature have taken an integrative
view of the segmentation problem by incorporating an intrinsic bias
correction step into the actual workflow.  The
advancements introduced with N4 permit such an adaptive
integration with Atropos.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.  This is done by setting
the `{\ttfamily --weight-image}' option on the N4 command line call
(see Listing \ref{listing:n4}) to the posterior probability image
corresponding to the white matter produced as output in the Atropos
call, i.e. `{\ttfamily Atropos --output}'.  N4 is
currently being added as a standard part of the Insight ToolKit.  The
evaluation section will illustrate inclusion of Atropos, N4 and ANTs
in a brain processing pipeline.  An example of this procedure, using brainweb data with
40\% RF bias, is in figure~\ref{fig:bwebrf40}.  We supply the
information necessary to repeat this example in the script entitled
`{\ttfamily  atroposBwebRF40FigureExample.sh}' which is available in the ANTs toolkit's Atropos
documentation folder as of svn commit 711.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=6in]{Figures/bwebN4ex.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
sequential processing and apply to brainweb single-subject data with
40\% RF bias and 3\% noise.  The $\beta$ for the MRF term is, here,
$0.2$.  Slice 71 of the input data is in (a).  The initial K-means
($K=3$) segmentation is in panel (b).  We use the brain mask to guide
N4 bias correction and produce the image, shown in (c).  We repeat the
K-means segmentation, but with the N4 corrected image as input, and
produce the segmentation in (d).  The average 3-tissue Dice overlap of
result (b) is 0.906 while the average overlap for (d) is 0.954.
Arrows highlight a region of large before-after segmentation
discrepancy.}
\label{fig:bwebrf40}
\end{center}
\end{figure}

\section{Evaluation}
Atropos~encodes a family of segmentation techniques that may be
instantiated for different applications but here we evaluate only two
of the many possibilites. First, we perform an evaluation on the
BrainWeb dataset using both the standard T1 image with multiple bias
and noise levels and also the BrainWeb20 data
\citep{Aubert-Broche2006a,Battaglini2008}.  In combination, this data
allows one to vary not only noise and bias but also the underlying
anatomy.  Second, we evaluate the use of Atropos in improving whole
brain parcellation and exercise its ability to efficiently solve {\em
many class} expectation maximization problem.  We choose this
evaluation problem in part to illustrate the flexibility of Atropos
and also the benefits of the novel, efficient implementation that
allows many-class problems to be solved with low memory usage ($<$2GB
for a 69 class model on 1mm$^3$ brain data).

\subsection{BrainWeb Evaluation}
\label{sec:bweb} The BrainWeb data is freely available at
\url{http://mouldy.bic.mni.mcgill.ca/brainweb/}.  We employ both the
individual subject data and the BrainWeb20 data in this evaluation.
\subsubsection{Single-Subject Evaluation} We use the single-subject
data with 3\% noise and bias 0, 20 and 40 \% RF inhomogeneity.  We
study the effect of the MRF prior term and initialization on
the Dice overlap between ground truth and the segmentation result for
each tissue.  We test both K-means and prior label image
initialization with MRF $\beta \in \{ 0.00 , 0.05 , 0.10 , 0.15 , 0.20
, 0.25 , 0.30 \}$ at each bias field.  We also feed the white matter
probability map derived from K-means into N4 to guide the bias
correction.  Segmentation is then repeated, with the same parameters,
but with the N4 corrected image as input.  The resulting algorithm is
similar to those that fix segmentation parameters while
estimating bias and fix bias while estimating segmentation parameters.
Thus, with this simple evaluation, we are able to compare the impact
of bias on our N4 and Atropos and also the validity of our
prior label image initialization.  The prior label image manual
initialization was based on a crude ITK-SNAP labeling of a small
set of voxels for each tissue.  Results, in terms of
Dice overlap, are shown in figure~\ref{fig:bweb1}  Because overlap
ratios with N4 bias correction approximate those of the zero bias data, we may
conclude that simple N4 pre-processing is adequate to correct even the 40\%
bias level.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small Brainweb single-subject results for
  each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
  appears to be between $0.1$ and $0.2$.  The prior label image
  initialization is competitive with but inferior to K-means.  The
  legend is in the same position in each graph, allowing a visual
  comparison of the results.  As one may see, the N4 assisted overlap
  values are consistent across bias field/RF inhomogeneity.}
\label{fig:bweb1}
\end{center}
\end{figure}
\subsubsection{20 Subject Evaluation} The single-subject brainweb
study in the previous section tested the basic Atropos options and the
benefit of N4 for segmentation in the presence of bias.  The 20
subject brainweb data allows us to use 2-fold cross-validation to test
our ability to segment different individuals reliably.  In this study,
we divide the 20 subjects equally into training and testing groups.
We then exploit the ground-truth labeling of the training data to
build both a group template \citep{Avants2010} and also prior
probability maps for each of the three major tissues in the cerebrum.
Each prior probability map is gained by deforming the ground truth
labels from each of the 10 training subjects to their template and
averaging component by component.  We then deform the template---and
priors---to the ten testing subjects and run Atropos with not only
{\ttfamily KMeans[3]} initialization but also {\ttfamily
  PriorProbabilityMap[3,priors\%02d.nii.gz,$w$] } where $w \in
\{0.0,0.5\}$.  We then switch the roles of testing and training sets to
gain 3-tissue segmentation for each of the twenty subjects.  
When $w=0$, the priors are only used in initializing
the model parameters but not during subsequent EM iterations.  
When $w=1$, the priors are only maintained in the product with the
likelihood during all EM iterations.  Results, in terms of bar
plots for Dice overlap mean and standard deviation, are shown in figure~\ref{fig:bweb20}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
\end{tabular}
\caption{\baselineskip 12pt \small Brainweb 20-subject results for
  each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
  is in $\{0,0.05,0.1,0.15,0.2\}$ and increases left to right.  The results show that the PriorProbabilityMaps with
  $w=0.5$ (far right) gives the best performance for all tissues.}
\label{fig:bweb20}
\end{center}
\end{figure}

\subsection{The Hammers Dataset Evaluation}
\label{sec:hammers}
We evaluate the ability to improve multi-template labeling results by
converting the group labels to probability maps and using them to
drive many-class EM segmentation.  The ground truth labels cover 69
classes and much of the brain.  Some unlabeled regions remain which we
assign to label 69 such that all brain parenchyma contains a unique
label.  Following \cite{Avants2010a}, the initialization of our
evaluation applies the script {\verb ants_multitemplate_labeling.sh }
(available in the ANTs toolkit) to the 19 Hammers evaluation datasets
located at \url{http://www.brain-development.org/}
\citep{Hammers2003,Heckemann2006}.  These initial majority voting
results are competitive with prior work
\citep{Heckemann2006,Heckemann2010} and serve as a baseline against
which we compare.

We first convert each of the 69 labels within the original evaluation
dataset to an individual image.  The remaining steps,
summarized briefly, are the same for each of the 19 subjects.  We
select one subject as an unlabeled target.  The other 18 datasets are
then mapped (as in the script above) to that subject.  We then deform,
individually, the 69 $\times$ 18 label images to the unlabeled
subject.  The label probability map is gained by averaging the 18
deformed images associated with each label.  We repeat this for each
subject.  The following parameters are the most relevant to this discussion:
{\ttfamily  -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz , 0.5
  ] -m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
Dice overlap, are shown in figure~\ref{fig:hammers}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=8in]{Figures/final_results_majority_vote_vs_atropos.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small The figure compares the Dice
overlap results from Atropos versus the raw results from majority
voting for each of 68 neuroanatomical regions and, in addition, the
unlabeled portions of the brain from the Hammers evaluation dataset.
We evaluated Atropos via N-fold cross-validation and employed
PriorProbabilityImages for each class where probabilities are gained
by averaging mapped subject labels. The color coding highlights those
regions that have the highest (yellow) and lowest (pink) improvement.
The significance of the improvement, measured by pairwise T-test, is
also shown as is a trinary coding of that improvement as: $+$
significant improvement, $-$ performance reduction, $\sim$ no change.}
\label{fig:hammers}
\end{center}
\end{figure}

\subsection{Reproducibility of this Evaluation}
The brainweb data is freely available.  We used single-subject
brainweb data as is.  The 20 subject data, however, required excluding
non-cerebrum tissue classes.  The Hammers data was also used as is.
The scripts for the single-subject brainweb study is available in
supplementary material.  The initialization procedure for the brainweb 20 and
the Hammers evaluation data is based on freely available scripts
included with the ANTs toolkit, specifically  {\verb ants_multitemplate_labeling.sh } and  {\verb atroposBwebRF40FigureExample.sh }.

\section{Discussion}
We introduced Atropos, the theory and implementation
details and documented its performance in a variety of use cases.  We
also showed evidence that the openly available N4 bias correction can easily used with Atropos to improve segmentation.
Furthermore, we used multiple subject brainweb data to build dataset
specific priors that provided the most consistent segmentation
performance across tissues.  Finally, we used majority voting to
initialize an Atropos EM solution to a 69 class brain parcellation
problem.  Significant improvements were gained in multiple brain
regions, in particular in temporal lobe cortex, the hippocampi and
amygdalae and the lateral ventricles.  This work, in summary, proves
the applicability of Atropos in both basic and extended use cases.

\subsection{Performance on Brainweb Data}
Atropos results are competitive with the state of the art.  For
instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF (bias field) 3\%
noise brainweb single subject data finding 0.932 (GM) and .961 (WM)
Dice. Results on 40\% RF were .934 (GM) and .961 (WM). SPM5 exhibits
insensitivity to bias similar to our own best results on the 40\%RF, 3
\% noise case (MRF-$\beta$=0.2, Kmeans $+$ N4) with Dice for GM is
0.950737 and for WM is 0.962586.  \cite{Nakamura2009} gave GM Dice
results (brainweb single 3\% noise) of 0.962 (0\%RF), 0.964(20\%RF)
and .956(40\%RF) which are slightly higher than either SPM5 or Atropos
results.  However, \cite{Nakamura2009} do not report WM or CSF numbers
for comparison.  Topology preserving methods also perform well.
\cite{Shiee2010} achieved Dice results for 3\% noise 20\% RF brainweb
single subject with 0.927, 0.912 and 0.900 for WM, GM and CSF,
respectively.  These are excellent numbers given the additional
constraint applied to the segmentation.  
\cite{Bazin2007c} proposed TOADS and, estimating from the
paper's graph, showed that the average Dice accuracy for 3\% noise various RF was
.95-.96 for WM, .93-.95 for GM and .92-.94 for CSF.
Perhaps the most recent balanced evaluation was performed in
\citep{Klauschen2009}, which reports confusion matrix numbers, rather
than Dice overlap.  Because the absolute true number of GM and WM
voxels for brainweb are known, we can convert the confusion matrix to
Dice.  In that case, the SPM5 Dice for brainweb single GM and WM is .8847
and .9085, respectively, while FreeSurfer and FSL's accuracy is
lower. The best GM result for 20 subjects is obtained by SPM5: 0.9295;
best WM Dice is from FSL: 0.9499.  We note that \cite{Klauschen2009}
used a comprehensive evaluation where quality of brain extraction also
contributed to the outcome.  Thus, the results must be interpreted
slightly differently than those from other papers.

\subsection{Performance on Hammers Data}
Our prior work, \citep{Avants2010b}, showed that the majority vote
initialization provided to Atropos by ANTs template mapping is
competitive with \cite{Heckemann2006}.  The Atropos EM extension of
improved these results further.  However, the Atropos EM segmentation
performed significantly worse.  This is not surprising, in that
Atropos EM assumes that signal from the likelihood and MRF term is
valuable in improving the segmentation.  This assumption held for
amygdala and lateral ventricles among other areas.  However, in
pallidum and corpus callosum (the most significant areas with loss of
performance), this is not true.  We believe the explanation is that
the intensity varies within these structures and that a more complex
intensity model (or finer parcellation) would be needed here.  An
alternative solution would be to use boundary conditions for these
structures, as in the PriorLabelImage Atropos option.

\subsection{Clinically-Related Evaluation}
While specifying performance on brainweb is highly
valuable, clinical validation is a second important aspect of
segmentation evaluation.  For instance, 
\citep{Freeborough1997,Westlye2009,Sanchez-Benavides2010,Chou2009,Bresser2011}
are only a few of the papers that evaluate segmentation performance with
respect to a known neurobiological outcome measure.  Atropos is
currently used in clinical studies and a number of clinically-focused,
application-specific evaluations are ongoing and will constitute
future work.  One early example of a clinically-focused Atropos
neuroimaging application is in \citep{Avants2010c}.  An additional successful
application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
\citep{Tustison2010a} which also used the open-source  Glamorous Glue
algorithm to impose topology constraints.  Thus, future work may
incorporate topology more closely into the Atropos methodology.

A more general advantage which extends beyond the scope of the experimental 
evaluation section of this paper is the flexibility of Atropos.   This includes 
not only $n$-tissue segmentation and dense volumetric cortical parcellation,
as reported in this work, but Atropos is also used in conjunction with our ANTs registration tools for robust
brain extraction which has reported good performance in comparison with other
popular, publicly available brain extraction tools \citep{Avants2010a}.

\section{Conclusion}
The Atropos software is freely available to the public.  We release
this code not only to make it available to clinical researchers but
with the hope that other researchers in segmentation will provide
feedback about the implementation decisions that we made.  EM
segmentation is non-trivial and there are numerous design alternatives
available not only in the models selected but also in the ICM coding, alternatives to ICM and the specific method in which prior
and likelihood are combined.  Due to the flexibility of Atropos, we
also hope that some of its capabilities, though not evaluated here,
are explored by the segmentation or clinical community.

\paragraph{Acknowledgments}
{This work was supported in part by NIH  {\bf {\textcolor{red}{ FIXME}}} }

\newpage

\bibliographystyle{neuroinformatics}
\bibliography{atropos}

\end{document}

\message{ !name(atropos.tex) !offset(-1465) }

### 2011-02-09 src/ANTS/Documentation/atropos/param_study_rf40ex.sh

@@ -1,0 +1,20 @@
+IMG=" T1_ICBM_1mm_3perc_40_bias/t1_icbm_normal_1mm_pn3_rf40.mhd "
+EX=Atropos
+EX=~/code/bin/ants_git/bin/Release/antsSegmentation 
+NC=3
+CONV=[3,0]
+MRF=0.20
+SOC=1
+# kmeans with N4
+NAMING=bwebRF40example
+WTIM=test_init_prob03.nii.gz
+# $EX  -d 3 -a  $IMG -i KMeans[$NC] -o [test_init.nii.gz,test_init_prob%02d.nii.gz] -m [${MRF},1x1x1] -x mask.nii.gz -p Socrates[${SOC}]  -c $CONV 
+BS=" N4BiasFieldCorrection -d 3  -h 0 " ; ITS=20 ; ITS2=3
+$BS  -i $IMG   -o  n4.nii.gz -s 2 -b [200] -w mask.nii.gz -c [${ITS}x${ITS}x${ITS}x${ITS}x${ITS2},0.0001] -x mask.nii.gz 
+exe="$EX  -d 3 -a  n4.nii.gz -i KMeans[$NC] -o [test.nii.gz,test_prob%02d.nii.gz] -m [${MRF},1x1x1] -x mask.nii.gz -p Socrates[${SOC}]  -c $CONV "
+echo $exe > ${NAMING}.txt 
+$exe #  >> ${NAMING}.txt 
+ImageMath 3 dice_out.txt DiceAndMinDistSum test.nii.gz ground_truth.nii.gz 
+cat dice_out.txt >> ${NAMING}.txt 
+rm dice* 
+

### 2011-02-09 src/ANTS/Documentation/atropos/runtex.sh

@@ -1,0 +1,13 @@
+#!/bin/sh
+
+for FL in $1  # antstheory # ants
+  do 
+for i in 1 2 3 
+do
+pdflatex $FL 
+bibtex   $FL
+done
+# cp ${FL}.pdf ../
+done 
+rm -f *.bbl *.log *.aux *.toc *.blg *.brf *.out *.spl *~ 
+

### 2011-02-09 src/ANTS/Examples/Data/test.sh

@@ -1,0 +1,29 @@
+~/code/bin/ants_git/bin/Release/antsSegmentation -d 2 -a r16slice.nii -i PriorLabelImage[3,r16slice_prior2.nii.gz,0.5]  -p Plato[1] -x mask.nii.gz -c [5,-0.1] -m [0.2] -o test.nii.gz  # -l 1[0.1,1] -l 2[0.1,1] -l 3[0.1,1]
+exit
+# 0.  Here, we use r16 as a template and r64 as a subject T1 image. 
+# 1. Segment the input images 
+  ThresholdImage 2 r16slice.nii mask.nii.gz 0.001 Inf
+  Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
+      -c [10,0.001] -m [0.05,1] -i kmeans[3] -x mask.nii.gz
+  ThresholdImage 2 r64slice.nii mask.nii.gz 0.001 Inf
+  Atropos -d 2 -a r64slice.nii -o [r64slice_seg.nii.gz,r64slice_prob_%02d.nii.gz] \
+       -c [10,0.001] -m [0.05,1] -i kmeans[3] -x mask.nii.gz
+# 2. Compute the deformation
+  sh ants.sh 2 r16slice_prob_02.nii.gz r64slice_prob_02.nii.gz SPM 100x100x100
+# 3. Smooth the segmented image 
+  SmoothImage 2 r64slice_prob_02.nii.gz  1. sr64slice_prob_02.nii.gz  
+# 3. Warp the segmented image by the 
+  WarpImageMultiTransform 2 sr64slice_prob_02.nii.gz  sr64slice_prob_02_norm.nii.gz \ 
+     -R r16slice.nii SPMWarp.nii.gz SPMAffine.txt 
+# 4. Compute the log Jacobian from the transformation.
+  ANTSJacobian 2 SPMWarp.nii.gz SPM 1  # take the log
+  gzip SPMlogjacobian.nii 
+# 5. Multiply the log-jacobian by the gray matter probability. 
+ ImageMath 2 SPMlogjacobianmask.nii.gz m SPMlogjacobian.nii.gz sr64slice_prob_02_norm.nii.gz
+# 6.  Estimate cortical thickness --- use svn revision 684 or greater (not in official release yet). 
[…]

### 2011-02-09 src/ANTS/ImageRegistration/ANTS_affine_registration2.h

@@ -1,0 +1,1356 @@
+#ifndef ANTS_AFFINE_REGISTRATION2_H_
+#define ANTS_AFFINE_REGISTRATION2_H_
+
+
+#include <vector>
+#include <stdlib.h>
+#include <time.h>
+#include "itkImage.h"
+#include "itkPoint.h"
+#include "itkCastImageFilter.h"
+#include "itkImageMaskSpatialObject.h"
+#include "itkCenteredEuler3DTransform.h"
+#include "itkRegularStepGradientDescentOptimizer.h"
+#include "itkGradientDescentOptimizer.h"
+#include "itkCenteredTransformInitializer.h"
+#include "itkMattesMutualInformationImageToImageMetric.h"
+#include "itkCorrelationCoefficientHistogramImageToImageMetric.h"
+#include "itkMultiResolutionImageRegistrationMethod.h"
+#include "itkLinearInterpolateImageFunction.h"
+#include "itkANTSAffine3DTransform.h"
+#include "itkCenteredRigid2DTransform.h"
+#include "itkANTSCenteredAffine2DTransform.h"
+#include "itkTransformFactory.h"
+#include "itkTransformFileReader.h"
+#include "itkTransformFileWriter.h"
+#include "itkImageRegionIteratorWithIndex.h"
+#include  "itkResampleImageFilter.h"
+#include "itkLinearInterpolateImageFunction.h"
+#include "itkWarpImageFilter.h"
+#include "itkWarpImageWAffineFilter.h"
+#include "itkImageMomentsCalculator.h"
+#include <vector>
+#include "ReadWriteImage.h"
+#include "itkMeanSquaresImageToImageMetric.h"
+#include "itkGradientDifferenceImageToImageMetric.h"
+#include "itkNormalizedCorrelationImageToImageMetric.h" 
+
+typedef enum{AffineWithMutualInformation=1, AffineWithMeanSquareDifference , AffineWithHistogramCorrelation, AffineWithNormalizedCorrelation , AffineWithGradientDifference } AffineMetricType;
+
+template<class TAffineTransformPointer, class TMaskImagePointer>
+class OptAffine{
+public:
+  typedef TAffineTransformPointer AffineTransformPointerType;
+  typedef TMaskImagePointer MaskImagePointerType;
+  typedef typename AffineTransformPointerType::ObjectType AffineTransformType;
+
+  OptAffine(){
+    MI_bins = 32;
+    MI_samples = 6000;
[…]

### 2011-02-09 src/ANTS/ImageRegistration/itkANTSImageTransformation.h

@@ -1,0 +1,169 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkANTSImageTransformation.h,v $
+  Language:  C++
+  Date:      $Date: 2008/11/15 23:46:06 $
+  Version:   $Revision: 1.18 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkANTSImageTransformation_h
+#define __itkANTSImageTransformation_h
+
+#include "itkObject.h"
+#include "itkObjectFactory.h"
+
+#include "antsCommandLineParser.h"
+#include "itkImage.h"
+#include "itkMacro.h"
+
+#include "itkCenteredEuler3DTransform.h"
+#include "itkQuaternionRigidTransform.h"
+#include "itkANTSCenteredAffine2DTransform.h"
+#include "itkANTSAffine3DTransform.h"
+// #include "itkCenteredRigid2DTransform.h"
+#include "itkCenteredTransformInitializer.h"
+#include  "itkTransformFileReader.h"
+#include "itkTransformFileWriter.h"
+namespace itk
+{
+
+template<unsigned int TDimension = 3, class TReal = float> 
+class ITK_EXPORT ANTSImageTransformation
+: public Object
+{
+  public:
+  /** Standard class typedefs. */
+  typedef ANTSImageTransformation                Self;
+  typedef Object                                           Superclass;
+  typedef SmartPointer<Self>                               Pointer;
+  typedef SmartPointer<const Self>                         ConstPointer;
+  
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self ); 
+  
+  typedef double TComp;
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( ANTSImageTransformation, Object ); 
+  itkStaticConstMacro( Dimension, unsigned int, TDimension );
+  itkStaticConstMacro( ImageDimension, unsigned int, TDimension );
+  
[…]

### 2011-02-09 src/ANTS/ImageRegistration/itkPICSLAdvancedNormalizationToolKit.h

@@ -1,0 +1,130 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkPICSLAdvancedNormalizationToolKit.h,v $
+  Language:  C++
+  Date:      $Date: 2008/11/15 23:46:06 $
+  Version:   $Revision: 1.18 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkPICSLAdvancedNormalizationToolKit_h
+#define __itkPICSLAdvancedNormalizationToolKit_h
+
+#include "itkObject.h"
+#include "itkObjectFactory.h"
+
+#include "itkANTSImageTransformation.h"
+#include "itkANTSImageRegistrationOptimizer.h"
+#include "itkANTSSimilarityMetric.h"
+#include "antsCommandLineParser.h"
+#include "itkImage.h"
+#include "itkMacro.h"
+#include "itkANTSLabeledPointSet.h"
+
+namespace itk
+{
+
+template<unsigned int TDimension = 3, class TReal = float> 
+class ITK_EXPORT PICSLAdvancedNormalizationToolKit
+: public Object
+{
+public:
+  /** Standard class typedefs. */
+  typedef PICSLAdvancedNormalizationToolKit                Self;
+  typedef Object                                           Superclass;
+  typedef SmartPointer<Self>                               Pointer;
+  typedef SmartPointer<const Self>                         ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self ); 
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( PICSLAdvancedNormalizationToolKit, Object ); 
+  itkStaticConstMacro( Dimension, unsigned int, TDimension );
+typedef double TComp;
+  typedef TReal                                            RealType; 
+  typedef Image<RealType, 
+    itkGetStaticConstMacro( Dimension )>                   ImageType;
+  typedef typename ImageType::Pointer                      ImagePointer;
[…]

### 2011-02-09 src/ANTS/ImageSegmentation/antsAtroposSegmentationImageFilter.h

@@ -1,0 +1,910 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: antsAtroposSegmentationImageFilter.h,v $
+  Language:  C++
+  Date:      $Date: $
+  Version:   $Revision: $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or
+  http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt
+  for details.
+
+  This software is distributed WITHOUT ANY WARRANTY; without even
+  the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+  PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __antsAtroposSegmentationImageFilter_h
+#define __antsAtroposSegmentationImageFilter_h
+
+#include "itkImageToImageFilter.h"
+
+#include "antsListSampleFunction.h"
+#include "antsListSampleToListSampleFilter.h"
+
+#include "itkArray.h"
+#include "itkBSplineScatteredDataPointSetToImageFilter.h"
+#include "itkConstNeighborhoodIterator.h"
+#include "itkFixedArray.h"
+#include "itkListSample.h"
+#include "itkMersenneTwisterRandomVariateGenerator.h"
+#include "itkNeighborhoodIterator.h"
+#include "itkPointSet.h"
+#include "itkVector.h"
+
+#include <algorithm>
+#include <vector>
+#include <map>
+#include <utility>
+
+namespace itk {
+namespace ants {
+
+/** \class AtroposSegmentationImageFilter
+ * \brief Atropos:  A Priori Classification with Registration Initialized
+ *  Template Assistance
+ *
+ * This filter provides an Expectation-Maximization framework for statistical
+ * segmentation where the intensity profile of each class is modeled as a
+ * mixture model and spatial smoothness is enforced by an MRF prior.
+ *
+ * Initial labeling can be performed by otsu thresholding, kmeans clustering,
+ * a set of user-specified prior probability images, or a prior label image.
+ * If specified, the latter two initialization options are also used as
+ * priors in the MRF update step.
+ *
+ * The assumed labeling is such that classes are assigned consecutive
+ * indices 1, 2, 3, etc.  Label 0 is reserved for the background when a
+ * mask is specified.
+ *
+ */
+
+template<class TInputImage, class TMaskImage
[…]

### 2011-02-09 src/ANTS/ImageSegmentation/antsGrubbsRosnerListSampleFilter.h

@@ -1,0 +1,121 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: antsGrubbsRosnerListSampleFilter.h,v $
+  Language:  C++
+  Date:      $Date: $
+  Version:   $Revision: $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or
+  http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt
+  for details.
+
+  This software is distributed WITHOUT ANY WARRANTY; without even
+  the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+  PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __antsGrubbsRosnerListSampleFilter_h
+#define __antsGrubbsRosnerListSampleFilter_h
+
+#include "antsListSampleToListSampleFilter.h"
+
+#include <vector>
+
+namespace itk {
+namespace ants {
+namespace Statistics {
+
+/** \class GrubbsRosnerListSampleFilter
+ * \brief Base class of filters intended to generate scalar samples from
+ * intensity samples.
+ *
+ */
+
+template<class TScalarListSample>
+class ITK_EXPORT GrubbsRosnerListSampleFilter
+: public ListSampleToListSampleFilter<TScalarListSample, TScalarListSample>
+{
+public:
+  /**
+   * Standard class typedefs.
+   */
+  typedef GrubbsRosnerListSampleFilter                        Self;
+  typedef ListSampleToListSampleFilter
+    <TScalarListSample, TScalarListSample>                    Superclass;
+  typedef SmartPointer<Self>                                  Pointer;
+  typedef SmartPointer<const Self>                            ConstPointer;
+
+  /**
+   * Standard macros
+   */
+  itkTypeMacro( GrubbsRosnerListSampleFilter,
+    ListSampleToScalarListSampleFilter );
+
+  /**
+   * Method for creation through the object factory.
+   */
+  itkNewMacro( Self );
+
+  /**
+   * Conveneient typedefs
+   */
+  typedef double                                      RealType;
+  typedef TScalarListSample                           ScalarListSampleType;
+  typedef typename ScalarListSampleType
+    ::MeasurementVectorType                           MeasurementVectorType;
[…]

### 2011-02-09 src/ANTS/ImageSegmentation/antsHistogramParzenWindowsListSampleFunction.h

@@ -1,0 +1,103 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: antsHistogramParzenWindowsListSampleFunction.h,v $
+  Language:  C++
+  Date:      $Date: $
+  Version:   $Revision: $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or
+  http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt
+  for details.
+
+  This software is distributed WITHOUT ANY WARRANTY; without even
+  the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+  PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __antsHistogramParzenWindowsListSampleFunction_h
+#define __antsHistogramParzenWindowsListSampleFunction_h
+
+#include "antsListSampleFunction.h"
+
+#include "itkImage.h"
+
+namespace itk {
+namespace ants {
+namespace Statistics {
+
+/** \class HistogramParzenWindowsListSampleFunction.h
+ * \brief point set filter.
+ */
+
+template <class TListSample, class TOutput = double, class TCoordRep = double>
+class ITK_EXPORT HistogramParzenWindowsListSampleFunction
+: public ListSampleFunction<TListSample, TOutput, TCoordRep>
+{
+public:
+  typedef HistogramParzenWindowsListSampleFunction         Self;
+  typedef ListSampleFunction
+    <TListSample, TOutput, TCoordRep>                      Superclass;
+  typedef SmartPointer<Self>                               Pointer;
+  typedef SmartPointer<const Self>                         ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( HistogramParzenWindowsListSampleFunction, ListSampleFunction );
+
+  typedef typename Superclass::InputListSampleType          InputListSampleType;
+  typedef typename Superclass::InputMeasurementVectorType   InputMeasurementVectorType;
+  typedef typename Superclass::InputMeasurementType         InputMeasurementType;
+
+  /** List sample typedef support. */
+  typedef TListSample                                       ListSampleType;
+
+  /** Other typedef */
+  typedef TOutput                                           RealType;
[…]

### 2011-02-09 src/ANTS/ImageSegmentation/antsListSampleFunction.h

@@ -1,0 +1,143 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: antsListSampleFunction.h,v $
+  Language:  C++
+  Date:      $Date: 2008/10/18 00:20:04 $
+  Version:   $Revision: 1.1.1.1 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or
+  http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt
+  for details.
+
+  This software is distributed WITHOUT ANY WARRANTY; without even
+  the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+  PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __antsListSampleFunction_h
+#define __antsListSampleFunction_h
+
+#include "itkFunctionBase.h"
+
+#include "itkArray.h"
+
+namespace itk {
+namespace ants {
+namespace Statistics {
+
+/** \class ListSampleFunction
+ * \brief Evaluates a function of an image at specified position.
+ *
+ * ListSampleFunction is a baseclass for all objects that evaluates
+ * a function of a list sample at a measurement
+ * This class is templated over the input list type, the type
+ * of the function output and the coordinate representation type
+ * (e.g. float or double).
+ *
+ * The input list sample is set via method SetInputListSample().
+ * The methods Evaluate() evaluates the function at a measurement vector.
+ *
+ * \ingroup ListSampleFunctions
+ */
+template <class TInputListSample, class TOutput, class TCoordRep = float>
+class ITK_EXPORT ListSampleFunction :
+  public FunctionBase<typename TInputListSample::MeasurementVectorType, TOutput>
+{
+public:
+
+  /** Standard class typedefs. */
+  typedef ListSampleFunction                              Self;
+  typedef FunctionBase<TInputListSample, TOutput>         Superclass;
+  typedef SmartPointer<Self>                              Pointer;
+  typedef SmartPointer<const Self>                        ConstPointer;
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( ListSampleFunction, FunctionBase );
+
+  /** InputListSampleType typedef support. */
+  typedef TInputListSample InputListSampleType;
+
[…]

### 2011-02-09 src/ANTS/ImageSegmentation/antsListSampleToListSampleFilter.h

@@ -1,0 +1,95 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: antsListSampleToListSampleFilter.h,v $
+  Language:  C++
+  Date:      $$
+  Version:   $Revision: $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or
+  http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt
+  for details.
+
+  This software is distributed WITHOUT ANY WARRANTY; without even
+  the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+  PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __antsListSampleToListSampleFilter_h
+#define __antsListSampleToListSampleFilter_h
+
+#include "itkProcessObject.h"
+
+namespace itk {
+namespace ants {
+namespace Statistics {
+
+/** \class ListSampleToListSampleFilter
+ * \brief Base class for filters that take a list sample as an input and output
+ * another list sample.
+ *
+ * ListSampleToListSampleFilter is the base class for all process objects that output
+ * list sample data, and require list sample data as input. Specifically, this class
+ * defines the SetInput() method for defining the input to a filter.
+ *
+ * \ingroup ListSampleFilters
+ *
+ */
+template <class TInputListSample, class TOutputListSample = TInputListSample>
+class ITK_EXPORT ListSampleToListSampleFilter : public ProcessObject
+{
+public:
+  /** Standard class typedefs. */
+  typedef ListSampleToListSampleFilter  Self;
+  typedef ProcessObject Superclass;
+  typedef SmartPointer<Self>  Pointer;
+  typedef SmartPointer<const Self>  ConstPointer;
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( ListSampleToListSampleFilter, ProcessObject );
+
+  /** Some convenient typedefs. */
+  typedef TInputListSample                       InputListSampleType;
+  typedef TOutputListSample                      OutputListSampleType;
+
+  /** Set the list sample input of this object.  */
+  void SetInput( const InputListSampleType *input );
+
+  /** Get the list sample input of this object.  */
+  InputListSampleType * GetInput();
+
+  /** Get the list sample output of this object.  */
[…]

### 2011-02-09 src/ANTS/ImageSegmentation/antsLogEuclideanGaussianListSampleFunction.h

@@ -1,0 +1,94 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: antsLogEuclideanGaussianListSampleFunction.h,v $
+  Language:  C++
+  Date:      $Date: $
+  Version:   $Revision: $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or
+  http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt
+  for details.
+
+  This software is distributed WITHOUT ANY WARRANTY; without even
+  the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+  PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __antsLogEuclideanGaussianListSampleFunction_h
+#define __antsLogEuclideanGaussianListSampleFunction_h
+
+#include "antsListSampleFunction.h"
+
+#include "itkVariableSizeMatrix.h"
+
+namespace itk {
+namespace ants {
+namespace Statistics {
+
+/** \class LogEuclideanGaussianListSampleFunction.h
+ * \brief
+ */
+
+template <class TListSample, class TOutput = double, class TCoordRep = double>
+class ITK_EXPORT LogEuclideanGaussianListSampleFunction
+: public ListSampleFunction<TListSample, TOutput, TCoordRep>
+{
+public:
+  typedef LogEuclideanGaussianListSampleFunction           Self;
+  typedef ListSampleFunction
+    <TListSample, TOutput, TCoordRep>                      Superclass;
+  typedef SmartPointer<Self>                               Pointer;
+  typedef SmartPointer<const Self>                         ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( LogEuclideanGaussianListSampleFunction, ListSampleFunction );
+
+  typedef typename Superclass::InputListSampleType          InputListSampleType;
+  typedef typename Superclass::InputMeasurementVectorType   InputMeasurementVectorType;
+  typedef typename Superclass::InputMeasurementType         InputMeasurementType;
+
+  /** Other typedef */
+  typedef TOutput                                           RealType;
+  typedef TOutput                                           OutputType;
[…]

### 2011-02-09 src/ANTS/ImageSegmentation/antsManifoldParzenWindowsListSampleFunction.h

@@ -1,0 +1,129 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: antsManifoldParzenWindowsListSampleFunction.h,v $
+  Language:  C++
+  Date:      $Date: $
+  Version:   $Revision: $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or
+  http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt
+  for details.
+
+  This software is distributed WITHOUT ANY WARRANTY; without even
+  the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+  PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __antsManifoldParzenWindowsListSampleFunction_h
+#define __antsManifoldParzenWindowsListSampleFunction_h
+
+#include "antsListSampleFunction.h"
+
+#include "itkGaussianMembershipFunction.h"
+#include "itkWeightedCentroidKdTreeGenerator.h"
+
+#include <vector>
+
+namespace itk {
+namespace ants {
+namespace Statistics {
+
+/** \class ManifoldParzenWindowsListSampleFunction.h
+ * \brief point set filter.
+ */
+
+template <class TListSample, class TOutput = double, class TCoordRep = double>
+class ITK_EXPORT ManifoldParzenWindowsListSampleFunction
+: public ListSampleFunction<TListSample, TOutput, TCoordRep>
+{
+public:
+  typedef ManifoldParzenWindowsListSampleFunction          Self;
+  typedef ListSampleFunction
+    <TListSample, TOutput, TCoordRep>                      Superclass;
+  typedef SmartPointer<Self>                               Pointer;
+  typedef SmartPointer<const Self>                         ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( ManifoldParzenWindowsListSampleFunction, ListSampleFunction );
+
+  typedef typename Superclass::InputListSampleType          InputListSampleType;
+  typedef typename Superclass::InputMeasurementVectorType   InputMeasurementVectorType;
+  typedef typename Superclass::InputMeasurementType         InputMeasurementType;
+
+  /** List sample typedef support. */
[…]

### 2011-02-09 src/ANTS/Scripts/ants.sh

@@ -1,0 +1,153 @@
+#!/bin/bash
+
+NUMPARAMS=$#
+
+MAXITERATIONS=30x90x20
+
+export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"} # EDIT THIS 
+if [ $NUMPARAMS -lt 3  ]
+then 
+echo " USAGE ::  "
+echo "  sh   ants.sh  ImageDimension  fixed.ext  moving.ext  OPTIONAL-OUTPREFIX   OPTIONAL-max-iterations  OPTIONAL-Labels-In-Fixed-Image-Space-To-Deform-To-Moving-Image     Option-DoANTSQC " 
+echo " be sure to set ANTSPATH environment variable " 
+echo " Max-Iterations in form :    JxKxL where " 
+echo "  J = max iterations at coarsest resolution (here, reduce by power of 2^2) "  
+echo " K = middle resolution iterations ( here, reduce by power of 2 ) "
+echo " L = fine resolution iterations ( here, full resolution ) -- this level takes much more time per iteration " 
+echo " an extra  Ix before JxKxL would add another level "
+echo " Default ierations is  $MAXITERATIONS  -- you can often get away with fewer for many apps " 
+echo " Other parameters are updates of the defaults used in the A. Klein evaluation paper in Neuroimage, 2009 " 
+exit
+fi
+
+#ANTSPATH=YOURANTSPATH
+if [  ${#ANTSPATH} -le 0 ]
+then
+echo " Please set ANTSPATH=LocationOfYourAntsBinaries "  
+echo " Either set this in your profile or directly, here in the script. " 
+echo " For example : " 
+echo " ANTSPATH=/home/yourname/bin/ants/ "
+exit
+else 
+echo " ANTSPATH is $ANTSPATH " 
+fi
+
+#initialization, here, is unbiased 
+DIM=$1
+
+if  [ ${#DIM} -gt 1 ]
+then
+echo " Problem with specified ImageDimension => User Specified Value = $DIM " 
+exit
+fi
+
+FIXED=$2
+
[…]

### 2011-02-09 src/ANTS/Scripts/antsIntroduction.sh

@@ -1,0 +1,677 @@
+#!/bin/bash
+
+VERSION="0.0.7"
+
+# Uncomment the line below in case you have not set the ANTSPATH variable in your environment.
+export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"}
+
+function Usage {
+    cat <<USAGE
+
+Usage: 
+
+$0 -d ImageDimension -r fixed.ext -i moving.ext 
+
+Compulsory arguments:
+
+-d:  ImageDimension: 2 or 3 (for 2 or 3 Dimensional registration)
+
+-i:  Input image
+
+-r:  Reference image 
+
+Optional arguments
+
+-f:  Force script to proceed even if headers may be incompatible (0 = no, 1 = yes (default))
+
+-l:  Labels-In-Fixed-Image-Space-To-Deform-To-Moving-Image (default is off)
+
+-m:  Max-iterations
+ 
+-n:  N3BiasFieldCorrection of moving image ( 0 = off (default); 1 = on )
+
+-o:  OUTPREFIX; A prefix that is prepended to all output files.
+
+-q:  Perform a Quality Check (QC) of the result ( 0 = off (default); 1 = on ).
+
+-s:  Type of similarity metric used for registration. 
+
+-t:  Type of transformation model used for registration. 
+
+ You can change the values for each of these methods in this script.
+
+ Use $0 -h for extended help.
+--------------------------------------------------------------------------------------
+ ANTS was created by:
+ Brian B. Avants, Nick Tustison and Gang Song
+ Penn Image Computing And Science Laboratory
+ University of Pennsylvania
+--------------------------------------------------------------------------------------
+ script written by N.M. van Strien, www.mri-tutorial.com | NTNU MR-Center
+--------------------------------------------------------------------------------------
+
+USAGE
+    exit 1
+}
+
+function Help {
+    cat <<Help
+
+ $0 will perform an elastic or diffeomorphic transformation of the moving image 
+ into the fixed image space. 
+
+ this script encodes some reasonable defaults for the parameters and is useful for testing and comparing methods
+
+ ANTS will test for convergence after 10 iterations to prevent the program from running forever. 
+ Convergence is determined by occurrence of a flat or inflection point in the similarity term time series. 
+
+ Usage: 
+
+ $0 -d ImageDimension -r fixed.ext -i moving.ext 
+
+ Compulsory arguments:
+
[…]

### 2011-02-09 src/ANTS/Scripts/antsaffine.sh

@@ -1,0 +1,79 @@
+#!/bin/sh 
+export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"}
+
+NUMPARAMS=$#
+
+if [ $NUMPARAMS -lt 3  ]
+then 
+echo " USAGE ::  "
+echo "  sh   ants.sh  ImageDimension  fixed.ext  moving.ext  OPTIONAL-OUTPREFIX   PURELY-RIGID  " 
+echo " be sure to set ANTSPATH environment variable " 
+echo " affine only registration " 
+exit
+fi
+
+#ANTSPATH=YOURANTSPATH
+if [  ${#ANTSPATH} -le 0 ]
+then
+echo " Please set ANTSPATH=LocationOfYourAntsBinaries "  
+echo " Either set this in your profile or directly, here in the script. " 
+echo " For example : " 
+echo $ANTSPATH 
+echo " ANTSPATH=/home/yourname/bin/ants/ "
+exit
+else 
+echo " ANTSPATH is $ANTSPATH " 
+fi
+
+#initialization, here, is unbiased 
+DIM=$1
+
+if  [ ${#DIM} -gt 1 ]
+then
+echo " Problem with specified ImageDimension => User Specified Value = $DIM " 
+exit
+fi
+
+FIXED=$2
+
+if  [ ${#FIXED} -lt 1 -o  ! -f $FIXED ]
+then
+echo " Problem with specified Fixed Image => User Specified Value = $FIXED " 
+exit
+fi
+
+MOVING=$3
+
+if  [ ${#MOVING} -lt 1 -o  ! -f $MOVING ]
+then
+echo " Problem with specified Moving Image => User Specified Value = $MOVING " 
+exit
+fi
+
+OUTPUTNAME=` echo $MOVING | cut -d '.' -f 1 `
+
+if [ $NUMPARAMS -gt 3  ]
+then 
+  OUTPUTNAME=${4}
+fi
+
+RIGID=" --rigid-affine false  " 
+if [ $NUMPARAMS -gt 4 ]
+then 
+  RIGID=" --rigid-affine true  --affine-gradient-descent-option  0.5x0.95x1.e-4x1.e-4  " 
+fi
+echo " Will this mapping be purely rigid?  $RIGID " 
+
+echo  " ANTSPATH  is $ANTSPATH     "
+ #below, some affine options 
[…]

### 2011-02-09 src/ANTS/Scripts/antswithdt.sh

@@ -1,0 +1,150 @@
+#!/bin/bash
+
+NUMPARAMS=$#
+
+MAXITERATIONS=30x90x20
+export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"}
+if [ $NUMPARAMS -lt 3  ]
+then 
+echo " USAGE ::  "
+echo "  sh   ants.sh  ImageDimension  fixed.ext  moving.ext Subject/Moving-DT-To-Deform-To-Fixed-Image  OPTIONAL-Subject/Moving-BZero-To-DistortionCorrect-To-Moving-T1-Image " 
+echo " be sure to set ANTSPATH environment variable " 
+echo " Max-Iterations in form :    JxKxL where " 
+echo "  J = max iterations at coarsest resolution (here, reduce by power of 2^2) "  
+echo " K = middle resolution iterations ( here, reduce by power of 2 ) "
+echo " L = fine resolution iterations ( here, full resolution ) -- this level takes much more time per iteration " 
+echo " an extra  Ix before JxKxL would add another level "
+echo " Default ierations is  $MAXITERATIONS  -- you can often get away with fewer for many apps " 
+echo " Other parameters are defaults used in the A. Klein evaluation paper in Neuroimage, 2009 " 
+echo " " 
+echo " The DT component is distortion corrected to the T1 image either by the B-Zero Image / Average-DWI Image " 
+echo " or whatever is passed as the last parameter --- Alternatively, we compute the FA from the DTI and then "
+echo" distortion correct to it.   One should test the validity of this approach on your data before widely applying ."
+echo " We use the cross-correlation to perform the distortion correction. " 
+echo " Some parameter tuning may be required, depending on the distortions present in your acquisition. "
+exit
+fi
+
+#initialization, here, is unbiased 
+DIM=$1
+
[…]

### 2011-02-09 src/ANTS/Scripts/guidedregistration.sh

@@ -1,0 +1,65 @@
+#!/bin/sh
+if [ $# -lt 7  ]
+then
+echo " USAGE \n  sh command.sh  fixed.nii fixedhipp.nii  moving.nii movinghipp.nii outputname  iterations DIM "
+echo " the template = fixed.nii   ,   the individual = moving.nii " 
+echo " iterations should be of the form  100x100x10 " 
+exit
+fi
+
+FIX=$1
+FIXH=$2
+MOV=$3
+MOVH=$4
+OUT=$5
+ITS=$6
+DIM=$7
+
+#ANTSPATH='/Users/stnava/Code/bin/ants/'
+
+if  [ ${#FIX} -lt 1 -o  ! -f $FIX ]
+then
+echo " Problem with specified Fixed Image => User Specified Value = $FIX " 
+exit
+fi
+if  [ ${#MOV} -lt 1 -o  ! -f $MOV ]
+then
+echo " Problem with specified Moving Image => User Specified Value = $MOV " 
+exit
+fi
+if  [ ${#FIXH} -lt 1 -o  ! -f $FIXH ]
+then
+echo " Problem with specified Fixed Label Image => User Specified Value = $FIXH " 
+exit
+fi
+if  [ ${#MOVH} -lt 1 -o  ! -f $MOVH ]
+then
+echo " Problem with specified Moving Label Image => User Specified Value = $MOVH " 
+exit
+fi
+
+#  == Important Parameters Begin ==
+
+LMWT=$9  # weight on landmarks
+
+INTWT=$8   # weight on intensity --  twice the landmarks
+
+# PSE/point-set-expectation/PointSetExpectation[fixedImage,movingImage,fixedPoints,movingPoints,weight,pointSetPercentage,pointSetSigma,boundaryPointsOnly,kNeighborhood, PartialMatchingIterations=100000]   
+# the partial matching option assumes the complete labeling is in the first set of label parameters ... 
+# more iterations leads to more symmetry in the matching  - 0 iterations means full asymmetry 	
+PCT=0.1 # percent of labeled voxels to use 
+PARZ=100 # PARZEN sigma 
+LM=PSE[${FIX},${MOV},$FIXH,$MOVH,${LMWT},${PCT},${PARZ},0,25,100]
+
[…]

### 2011-02-09 src/ANTS/Temporary/itkAutumnColormapFunctor.h

@@ -1,0 +1,77 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkAutumnColormapFunctor.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 19:22:31 $
+  Version:   $Revision: 1.3 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkAutumnColormapFunctor_h
+#define __itkAutumnColormapFunctor_h
+
+#include "itkColormapFunctor.h"
+
+namespace itk {
+
+namespace Functor {  
+
+/**
+ * \class AutumnColormapFunctor
+ * \brief Function object which maps a scalar value into an RGB colormap value.
+ *
+ * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and James C. Gee
+ * 
+ * This code was contributed in the Insight Journal paper:
+ *
+ * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
+ * http://www.insight-journal.org/browse/publication/285
+ * http://hdl.handle.net/1926/1452
+ *
+ */
+template< class TScalar, class TRGBPixel >
+class ITK_EXPORT AutumnColormapFunctor
+: public ColormapFunctor<TScalar, TRGBPixel>
+{
+public:
+
+  typedef AutumnColormapFunctor                  Self;
+  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
+  typedef SmartPointer<Self>                     Pointer;
+  typedef SmartPointer<const Self>               ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  typedef typename Superclass::RGBPixelType      RGBPixelType;
+  typedef typename Superclass::ScalarType        ScalarType;  
+  typedef typename Superclass::RealType          RealType;
+  
[…]

### 2011-02-09 src/ANTS/Temporary/itkContourDirectedMeanDistanceImageFilter.h

@@ -1,0 +1,175 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkContourDirectedMeanDistanceImageFilter.h,v $
+  Language:  C++
+  Date:      $Date: 2009-04-25 12:27:20 $
+  Version:   $Revision: 1.7 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkContourDirectedMeanDistanceImageFilter_h
+#define __itkContourDirectedMeanDistanceImageFilter_h
+
+#include "itkImageToImageFilter.h"
+#include "itkNumericTraits.h"
+#include "itkArray.h"
+#include "itkImage.h"
+
+namespace itk {
+
+/** \class ContourDirectedMeanDistanceImageFilter
+ * \brief Computes the directed Mean distance between the boundaries of
+ * non-zero pixel regions of two images.
+ *
+ * ContourDirectedMeanDistanceImageFilter computes the distance between the set
+ * non-zero pixels of two images using the following formula:
+ * \f[ h(A,B) = \mathrm{mean}_{a \in A} \min_{b \in B} \| a - b\| \f]
+ * where \f$A\f$ and \f$B\f$ are respectively the set of non-zero pixels
+ * in the first and second input images. It identifies the point \f$ a \in A \f$
+ * that is farthest from any point of \f$B\f$ and measures the distance from \f$a\f$
+ * to the nearest neighbor in \f$B\f$. Note that this function is not
+ * is not symmetric and hence is not a true distance.
+ *
+ * In particular, this filter uses the SignedDanielssonDistanceMapImageFilter inside
+ * to compute distance map from all non-zero pixels in the second image. It
+ * then computes the mean distance (in pixels) within the boundary pixels of
+ * non-zero regions in the first image.
+ *
+ * Use MeanDistanceImageFilter to compute the full Mean distance.
+ *
+ * This filter requires the largest possible region of the first image and the
[…]

### 2011-02-09 src/ANTS/Temporary/itkCoolColormapFunctor.h

@@ -1,0 +1,79 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkCoolColormapFunctor.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 19:08:09 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkCoolColormapFunctor_h
+#define __itkCoolColormapFunctor_h
+
+#include "itkColormapFunctor.h"
+
+namespace itk {
+
+namespace Functor {  
+
+/**
+ * \class CoolColormapFunctor
+ * \brief Function object which maps a scalar value into an RGB colormap value.
+ *
+ * 
+ * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
+ * 
+ * This code was contributed in the Insight Journal paper:
+ *
+ * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
+ * http://www.insight-journal.org/browse/publication/285
+ * http://hdl.handle.net/1926/1452
+ *
+ */
+template< class TScalar, class TRGBPixel >
+class ITK_EXPORT CoolColormapFunctor
+: public ColormapFunctor<TScalar, TRGBPixel>
+{
+public:
+
+  typedef CoolColormapFunctor                    Self;
+  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
+  typedef SmartPointer<Self>                     Pointer;
+  typedef SmartPointer<const Self>               ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  typedef typename Superclass::RGBPixelType      RGBPixelType;
+  typedef typename Superclass::ScalarType        ScalarType;  
+  typedef typename Superclass::RealType          RealType;
+  
[…]

### 2011-02-09 src/ANTS/Temporary/itkCopperColormapFunctor.h

@@ -1,0 +1,79 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkCopperColormapFunctor.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 19:08:09 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkCopperColormapFunctor_h
+#define __itkCopperColormapFunctor_h
+
+#include "itkColormapFunctor.h"
+
+namespace itk {
+
+namespace Functor {  
+
+/**
+ * \class CopperColormapFunctor
+ * \brief Function object which maps a scalar value into an RGB colormap value.
+ *
+ *
+ * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
+ * 
+ * This code was contributed in the Insight Journal paper:
+ *
+ * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
+ * http://www.insight-journal.org/browse/publication/285
+ * http://hdl.handle.net/1926/1452
+ *
+ */
+template< class TScalar, class TRGBPixel >
+class ITK_EXPORT CopperColormapFunctor
+: public ColormapFunctor<TScalar, TRGBPixel>
+{
+public:
+
+  typedef CopperColormapFunctor                  Self;
+  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
+  typedef SmartPointer<Self>                     Pointer;
+  typedef SmartPointer<const Self>               ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  typedef typename Superclass::RGBPixelType      RGBPixelType;
+  typedef typename Superclass::ScalarType        ScalarType;  
+  typedef typename Superclass::RealType          RealType;
+  
[…]

### 2011-02-09 src/ANTS/Temporary/itkDijkstrasAlgorithm.h

@@ -1,0 +1,548 @@
+/*=========================================================================
+
+  Program:   Insight Segmentation & Registration Toolkit (ITK)
+  Module:    $RCSfile: itkDijkstrasAlgorithm.h,v $ Language:  C++
+  Date:      $Date: 2008/06/05 18:39:14 $
+  Version:   $Revision: 1.8 $
+
+Copyright (c) 2001 Insight Consortium
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without
+modification, are permitted provided that the following conditions are met:
+
+ * Redistributions of source code must retain the above copyright notice,
+   this list of conditions and the following disclaimer.
+
+ * Redistributions in binary form must reproduce the above copyright notice,
+   this list of conditions and the following disclaimer in the documentation
+   and/or other materials provided with the distribution.
+
+ * The name of the Insight Consortium, nor the names of any consortium members,
+   nor of any contributors, may be used to endorse or promote products derived
+   from this software without specific prior written permission.
+
+  * Modified source versions must be plainly marked as such, and must not be
+    misrepresented as being the original software.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDER AND CONTRIBUTORS ``AS IS''
+AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE
+ARE DISCLAIMED. IN NO EVENT SHALL THE AUTHORS OR CONTRIBUTORS BE LIABLE FOR
+ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
+DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR
+SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER
+CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY,
+OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+=========================================================================*/
[…]

### 2011-02-09 src/ANTS/Temporary/itkFEMElement3DMembrane1DOF.h

@@ -1,0 +1,132 @@
+/*=========================================================================
+
+  Program:   Insight Segmentation & Registration Toolkit
+  Module:    $RCSfile: itkFEMElement3DMembrane1DOF.h,v $
+  Language:  C++
+  Date:      $Date: 2005/11/08 15:34:16 $
+  Version:   $Revision: 1.1.1.1 $
+
+  Copyright (c) 2002 Insight Consortium. All rights reserved.
+  See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+
+#ifndef __itkFEMElement3DMembrane1DOF_h
+#define __itkFEMElement3DMembrane1DOF_h
+
+#include "itkFEMElementBase.h"
+#include "itkFEMMaterialLinearElasticity.h"
+
+namespace itk {
+namespace fem {
+
+
+
+
+/**
+ * \class Element3DMembrane1DOF
+ * \brief Class that is used to define a membrane energy problem in 3D space.
+ *
+ * This class only defines the physics of the problem. Use his class together
+ * with element classes that specify the geometry to fully define the element.
+ *
+ * You can specify one template parameter:
+ *
+ *   TBaseClass - Class from which Element3DMembrane1DOF is derived. TBaseClass must
+ *                be derived from the Element base class. This enables you
+ *                to use this class at any level of element definition.
+ *                If not specified, it defaults to the Element base class.
+ */
+template<class TBaseClass=Element>
+class Element3DMembrane1DOF : public TBaseClass
+{
+FEM_ABSTRACT_CLASS(Element3DMembrane1DOF,TBaseClass)
+public:
+
+  // Repeat the required typedefs and enums from parent class
+  typedef typename Superclass::Float Float;
+  typedef typename Superclass::MatrixType MatrixType;
+  typedef typename Superclass::VectorType VectorType;
+
+  /**
+   * Read data for this class from input stream
+   */
+  virtual void Read( std::istream&, void* info );
+
+  /**
+   * Write this class to output stream
+   */
[…]

### 2011-02-09 src/ANTS/Temporary/itkGreenColormapFunctor.h

@@ -1,0 +1,78 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkGreenColormapFunctor.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 19:08:09 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkGreenColormapFunctor_h
+#define __itkGreenColormapFunctor_h
+
+#include "itkColormapFunctor.h"
+
+namespace itk {
+
+namespace Functor {  
+
+/**
+ * \class GreenColormapFunctor
+ * \brief Function object which maps a scalar value into an RGB colormap value.
+ *
+ * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
+ * 
+ * This code was contributed in the Insight Journal paper:
+ *
+ * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
+ * http://www.insight-journal.org/browse/publication/285
+ * http://hdl.handle.net/1926/1452
+ *
+ */
+template< class TScalar, class TRGBPixel >
+class ITK_EXPORT GreenColormapFunctor
+: public ColormapFunctor<TScalar, TRGBPixel>
+{
+public:
+
+  typedef GreenColormapFunctor                   Self;
+  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
+  typedef SmartPointer<Self>                     Pointer;
+  typedef SmartPointer<const Self>               ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  typedef typename Superclass::RGBPixelType      RGBPixelType;
+  typedef typename Superclass::ScalarType        ScalarType;  
+  typedef typename Superclass::RealType          RealType;
+  
+  virtual RGBPixelType operator()( const TScalar & ) const;
+
[…]

### 2011-02-09 src/ANTS/Temporary/itkGreyColormapFunctor.h

@@ -1,0 +1,78 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkGreyColormapFunctor.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 19:22:31 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkGreyColormapFunctor_h
+#define __itkGreyColormapFunctor_h
+
+#include "itkColormapFunctor.h"
+
+namespace itk {
+
+namespace Functor {  
+
+/**
+ * \class GreyColormapFunctor
+ * \brief Function object which maps a scalar value into an RGB colormap value.
+ *
+ * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
+ * 
+ * This code was contributed in the Insight Journal paper:
+ *
+ * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
+ * http://www.insight-journal.org/browse/publication/285
+ * http://hdl.handle.net/1926/1452
+ *
+ */
+template< class TScalar, class TRGBPixel >
+class ITK_EXPORT GreyColormapFunctor
+: public ColormapFunctor<TScalar, TRGBPixel>
+{
+public:
+
+  typedef GreyColormapFunctor                    Self;
+  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
+  typedef SmartPointer<Self>                     Pointer;
+  typedef SmartPointer<const Self>               ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  typedef typename Superclass::RGBPixelType      RGBPixelType;
+  typedef typename Superclass::ScalarType        ScalarType;  
+  typedef typename Superclass::RealType          RealType;
+  
+  virtual RGBPixelType operator()( const TScalar & ) const;
[…]

### 2011-02-09 src/ANTS/Temporary/itkJetColormapFunctor.h

@@ -1,0 +1,78 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkJetColormapFunctor.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 19:22:31 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkJetColormapFunctor_h
+#define __itkJetColormapFunctor_h
+
+#include "itkColormapFunctor.h"
+
+namespace itk {
+
+namespace Functor {  
+
+/**
+ * \class JetColormapFunctor
+ * \brief Function object which maps a scalar value into an RGB colormap value.
+ *
+ * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
+ * 
+ * This code was contributed in the Insight Journal paper:
+ *
+ * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
+ * http://www.insight-journal.org/browse/publication/285
+ * http://hdl.handle.net/1926/1452
+ *
+ */
+template< class TScalar, class TRGBPixel >
+class ITK_EXPORT JetColormapFunctor
+: public ColormapFunctor<TScalar, TRGBPixel>
+{
+public:
+
+  typedef JetColormapFunctor                     Self;
+  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
+  typedef SmartPointer<Self>                     Pointer;
+  typedef SmartPointer<const Self>               ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  typedef typename Superclass::RGBPixelType      RGBPixelType;
+  typedef typename Superclass::ScalarType        ScalarType;  
+  typedef typename Superclass::RealType          RealType;
+  
+  virtual RGBPixelType operator()( const TScalar & ) const;
+
+protected:
[…]

### 2011-02-09 src/ANTS/Temporary/itkLabelContourImageFilter.h

@@ -1,0 +1,213 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkLabelContourImageFilter.h,v $
+  Language:  C++
+  Date:      $Date: 2009/03/10 17:30:41 $
+  Version:   $Revision: 1.1 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+
+#ifndef __itkLabelContourImageFilter_h
+#define __itkLabelContourImageFilter_h
+
+#include "itkInPlaceImageFilter.h"
+#include "itkImage.h"
+#include "itkConceptChecking.h"
+#include <vector>
+#include <map>
+#include "itkProgressReporter.h"
+#include "itkBarrier.h"
+
+namespace itk
+{
+
+/**
+ * \class LabelContourImageFilter
+ * \brief Give the pixels on the border of an object.
+ *
+ * LabelContourImageFilter labels the pixels on the borders
+ * of the objects in a binary image.
+ *
+ * \sa ImageToImageFilter 
+ */
+
+template <class TInputImage, class TOutputImage>
+class ITK_EXPORT LabelContourImageFilter : 
+    public InPlaceImageFilter< TInputImage, TOutputImage > 
+{
+public:
+  /**
+   * Standard "Self" & Superclass typedef.
+   */
+  typedef LabelContourImageFilter                         Self;
+  typedef InPlaceImageFilter< TInputImage, TOutputImage > Superclass;
+
+  /**
+   * Types from the Superclass
+   */
+  typedef typename Superclass::InputImagePointer InputImagePointer;
+
+  /**
+   * Extract some information from the image types.  Dimensionality
+   * of the two images is assumed to be the same.
+   */
+  typedef typename TOutputImage::PixelType         OutputPixelType;
+  typedef typename TOutputImage::InternalPixelType OutputInternalPixelType;
+  typedef typename TInputImage::PixelType          InputPixelType;
[…]

### 2011-02-09 src/ANTS/Temporary/itkManifoldIntegrationAlgorithm.h

@@ -1,0 +1,311 @@
+/*=========================================================================
+
+  Program:   Insight Segmentation & Registration Toolkit (ITK)
+  Module:    $RCSfile: itkManifoldIntegrationAlgorithm.h,v $ Language:  C++
+  Date:      $Date: 2008/03/11 16:33:05 $
+  Version:   $Revision: 1.2 $
+
+Copyright (c) 2001 Insight Consortium
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without
+modification, are permitted provided that the following conditions are met:
+
+ * Redistributions of source code must retain the above copyright notice,
+   this list of conditions and the following disclaimer.
+
+ * Redistributions in binary form must reproduce the above copyright notice,
+   this list of conditions and the following disclaimer in the documentation
+   and/or other materials provided with the distribution.
+
+ * The name of the Insight Consortium, nor the names of any consortium members,
+   nor of any contributors, may be used to endorse or promote products derived
+   from this software without specific prior written permission.
+
+  * Modified source versions must be plainly marked as such, and must not be
+    misrepresented as being the original software.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDER AND CONTRIBUTORS ``AS IS''
+AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
+IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE
+ARE DISCLAIMED. IN NO EVENT SHALL THE AUTHORS OR CONTRIBUTORS BE LIABLE FOR
+ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
+DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR
+SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER
+CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY,
+OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
[…]

### 2011-02-09 src/ANTS/Temporary/itkRedColormapFunctor.h

@@ -1,0 +1,78 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkRedColormapFunctor.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 19:22:31 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkRedColormapFunctor_h
+#define __itkRedColormapFunctor_h
+
+#include "itkColormapFunctor.h"
+
+namespace itk {
+
+namespace Functor {  
+
+/**
+ * \class RedColormapFunctor
+ * \brief Function object which maps a scalar value into an RGB colormap value.
+ *
+ * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
+ * 
+ * This code was contributed in the Insight Journal paper:
+ *
+ * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
+ * http://www.insight-journal.org/browse/publication/285
+ * http://hdl.handle.net/1926/1452
+ *
+ */
+template< class TScalar, class TRGBPixel >
+class ITK_EXPORT RedColormapFunctor
+: public ColormapFunctor<TScalar, TRGBPixel>
+{
+public:
+
+  typedef RedColormapFunctor                     Self;
+  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
+  typedef SmartPointer<Self>                     Pointer;
+  typedef SmartPointer<const Self>               ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  typedef typename Superclass::RGBPixelType      RGBPixelType;
+  typedef typename Superclass::ScalarType        ScalarType;  
+  typedef typename Superclass::RealType          RealType;
+  
+  virtual RGBPixelType operator()( const TScalar & ) const;
+
+protected:
[…]

### 2011-02-09 src/ANTS/Temporary/itkScalarToRGBColormapImageFilter.h

@@ -1,0 +1,139 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkScalarToRGBColormapImageFilter.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 12:51:13 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkScalarToRGBColormapImageFilter_h
+#define __itkScalarToRGBColormapImageFilter_h
+
+#include "itkImageToImageFilter.h"
+
+#include "itkColormapFunctor.h"
+
+namespace itk
+{
+
+/** \class ScalarToRGBColormapImageFilter
+ * \brief Implements pixel-wise intensity->rgb mapping operation on one image.
+ *
+ * This class is parameterized over the type of the input image and
+ * the type of the output image.
+ *
+ * ScalarToRGBColormapImageFilter
+ *
+ * \sa BinaryFunctorImageFilter TernaryFunctorImageFilter
+ *
+ * \ingroup   IntensityImageFilters     Multithreaded
+ */
+template <class TInputImage, class TOutputImage>
+class ITK_EXPORT ScalarToRGBColormapImageFilter
+: public ImageToImageFilter<TInputImage, TOutputImage>
+{
+public:
+  /** Standard class typedefs. */
+  typedef ScalarToRGBColormapImageFilter                 Self;
+  typedef ImageToImageFilter<TInputImage, TOutputImage>  Superclass;
+  typedef SmartPointer<Self>                             Pointer;
+  typedef SmartPointer<const Self>                       ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( ScalarToRGBColormapImageFilter, ImageToImageFilter );
+
+  /** Some typedefs. */
+  typedef TInputImage                                    InputImageType;
+  typedef typename InputImageType::ConstPointer          InputImagePointer;
[…]

### 2011-02-09 src/ANTS/Temporary/itkSummerColormapFunctor.h

@@ -1,0 +1,78 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkSummerColormapFunctor.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 19:22:31 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkSummerColormapFunctor_h
+#define __itkSummerColormapFunctor_h
+
+#include "itkColormapFunctor.h"
+
+namespace itk {
+
+namespace Functor {  
+
+/**
+ * \class SummerColormapFunctor
+ * \brief Function object which maps a scalar value into an RGB colormap value.
+ *
+ * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
+ * 
+ * This code was contributed in the Insight Journal paper:
+ *
+ * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
+ * http://www.insight-journal.org/browse/publication/285
+ * http://hdl.handle.net/1926/1452
+ *
+ */
+template< class TScalar, class TRGBPixel >
+class ITK_EXPORT SummerColormapFunctor
+: public ColormapFunctor<TScalar, TRGBPixel>
+{
+public:
+
+  typedef SummerColormapFunctor                  Self;
+  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
+  typedef SmartPointer<Self>                     Pointer;
+  typedef SmartPointer<const Self>               ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  typedef typename Superclass::RGBPixelType      RGBPixelType;
+  typedef typename Superclass::ScalarType        ScalarType;  
+  typedef typename Superclass::RealType          RealType;
+  
+  virtual RGBPixelType operator()( const TScalar & ) const;
+
[…]

### 2011-02-09 src/ANTS/Temporary/itkWinterColormapFunctor.h

@@ -1,0 +1,78 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkWinterColormapFunctor.h,v $
+  Language:  C++
+  Date:      $Date: 2009-05-15 19:22:31 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkWinterColormapFunctor_h
+#define __itkWinterColormapFunctor_h
+
+#include "itkColormapFunctor.h"
+
+namespace itk {
+
+namespace Functor {  
+
+/**
+ * \class WinterColormapFunctor
+ * \brief Function object which maps a scalar value into an RGB colormap value.
+ *
+ * \author Nicholas Tustison, Hui Zhang, Gaetan Lehmann, Paul Yushkevich and ̈James C. Gee
+ * 
+ * This code was contributed in the Insight Journal paper:
+ *
+ * "Meeting Andy Warhol Somewhere Over the Rainbow: RGB Colormapping and ITK"
+ * http://www.insight-journal.org/browse/publication/285
+ * http://hdl.handle.net/1926/1452
+ *
+ */
+template< class TScalar, class TRGBPixel >
+class ITK_EXPORT WinterColormapFunctor
+: public ColormapFunctor<TScalar, TRGBPixel>
+{
+public:
+
+  typedef WinterColormapFunctor                  Self;
+  typedef ColormapFunctor<TScalar, TRGBPixel>    Superclass;
+  typedef SmartPointer<Self>                     Pointer;
+  typedef SmartPointer<const Self>               ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  typedef typename Superclass::RGBPixelType      RGBPixelType;
+  typedef typename Superclass::ScalarType        ScalarType;  
+  typedef typename Superclass::RealType          RealType;
+  
+  virtual RGBPixelType operator()( const TScalar & ) const;
+
[…]

### 2011-02-09 src/ANTS/Tensor/itkDecomposeTensorFunction2.h

@@ -1,0 +1,95 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkDecomposeTensorFunction2.h,v $
+  Language:  C++
+  Date:      $Date: 2009/01/29 00:15:47 $
+  Version:   $Revision: 1.2 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkDecomposeTensorFunction2_h
+#define __itkDecomposeTensorFunction2_h
+
+#include "itkVariableSizeMatrix.h"
+
+namespace itk
+{
+/** \class DecomposeTensorFunction2
+ *
+ */
+template <typename TInput, 
+          typename TRealType = float, 
+          typename TOutput = itk::VariableSizeMatrix<TRealType>
+>
+class DecomposeTensorFunction2
+{
+public:
+  /** Standard class typedefs. */
+  typedef DecomposeTensorFunction2 Self;
+  typedef SmartPointer<Self> Pointer;
+  typedef SmartPointer<const Self>  ConstPointer;
+
+  /** Extract some information from the image types.  Dimensionality
+   * of the two images is assumed to be the same. */
+  typedef TInput                                          InputMatrixType;
+  typedef TOutput                                         OutputMatrixType;
+
+  /** Define the data type and the vector of data type used in calculations. */
+  typedef TRealType                                       RealType;
+
+  // Wrappers for vnl routines
+  void EvaluateEigenDecomposition( InputMatrixType&, 
+    OutputMatrixType&, OutputMatrixType& );
+
+  void EvaluateSymmetricEigenDecomposition( InputMatrixType&, 
+    OutputMatrixType&, OutputMatrixType& );
+
+  void EvaluateQRDecomposition( InputMatrixType&, 
+    OutputMatrixType&, OutputMatrixType& );
+
[…]

### 2011-02-09 src/ANTS/Tensor/itkLogTensorImageFilter.h

@@ -1,0 +1,128 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkLogTensorImageFilter.h,v $
+  Language:  C++
+  Date:      $Date: 2009/01/27 19:56:28 $
+  Version:   $Revision: 1.1 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkLogTensorImageFilter_h
+#define __itkLogTensorImageFilter_h
+
+#include "itkImageToImageFilter.h"
+#include "itkImage.h"
+#include "itkNumericTraits.h"
+#include "TensorFunctions.h"
+
+
+namespace itk
+{
+/** \class LogTensorImageFilter
+ * \brief Applies an averaging filter to an image
+ *
+ * Computes an image where a given pixel is the mean value of the
+ * the pixels in a neighborhood about the corresponding input pixel.
+ *
+ * A mean filter is one of the family of linear filters.  
+ *
+ * \sa Image
+ * \sa Neighborhood
+ * \sa NeighborhoodOperator
+ * \sa NeighborhoodIterator
+ * 
+ * \ingroup IntensityImageFilters
+ */
+template <class TInputImage, class TOutputImage>
+class ITK_EXPORT LogTensorImageFilter :
+    public ImageToImageFilter< TInputImage, TOutputImage >
+{
+public:
+  /** Extract dimension from input and output image. */
+  itkStaticConstMacro(InputImageDimension, unsigned int,
+                      TInputImage::ImageDimension);
+  itkStaticConstMacro(OutputImageDimension, unsigned int,
+                      TOutputImage::ImageDimension);
+
+  /** Convenient typedefs for simplifying declarations. */
+  typedef TInputImage InputImageType;
+  typedef TOutputImage OutputImageType;
+
+  /** Standard class typedefs. */
+  typedef LogTensorImageFilter Self;
+  typedef ImageToImageFilter< InputImageType, OutputImageType> Superclass;
+  typedef SmartPointer<Self> Pointer;
[…]

### 2011-02-09 src/ANTS/Utilities/BinaryImageToMeshFilter.h

@@ -1,0 +1,456 @@
+/********************************************************************************
+ * Create a VTK mesh from the white matter image
+ *******************************************************************************/
+#ifndef __BinaryImageToMeshFilter_h_
+#define __BinaryImageToMeshFilter_h_
+
+#include "itkImage.h"
+#include "itkResampleImageFilter.h"
+#include "itkWindowedSincInterpolateImageFunction.h"
+#include "itkNearestNeighborInterpolateImageFunction.h"
+#include "itkAntiAliasBinaryImageFilter.h"
+#include "itkUnaryFunctorImageFilter.h"
+//#include "itkBinaryWellComposed3DImageFilter.h"
+#include "itkMinimumMaximumImageCalculator.h"
+#include "itkCommand.h"
+
+#include <vtkCellArray.h>
+#include <vtkPolyData.h>
+#include <vtkTriangleFilter.h>
+#include <vtkImageMarchingCubes.h>
+#include <vtkImageImport.h>
+#include <itkVTKImageExport.h>
+#include <vtkDecimatePro.h>
+#include <vtkPolyDataConnectivityFilter.h>
+#include <vtkStripper.h>
+#include <vtkCleanPolyData.h>
+#include <vtkSmoothPolyDataFilter.h>
+#include <vtkImageData.h>
+
+#include <iostream>
+
+using namespace std;
+
+template<class TImage>
+void ConnectITKToVTK(itk::VTKImageExport<TImage> *fltExport,vtkImageImport *fltImport)
+{
+  fltImport->SetUpdateInformationCallback( fltExport->GetUpdateInformationCallback());
+  fltImport->SetPipelineModifiedCallback( fltExport->GetPipelineModifiedCallback());
+  fltImport->SetWholeExtentCallback( fltExport->GetWholeExtentCallback());
+  fltImport->SetSpacingCallback( fltExport->GetSpacingCallback());
+  fltImport->SetOriginCallback( fltExport->GetOriginCallback());
+  fltImport->SetScalarTypeCallback( fltExport->GetScalarTypeCallback());
+  fltImport->SetNumberOfComponentsCallback( fltExport->GetNumberOfComponentsCallback());
+  fltImport->SetPropagateUpdateExtentCallback( fltExport->GetPropagateUpdateExtentCallback());
+  fltImport->SetUpdateDataCallback( fltExport->GetUpdateDataCallback());
[…]

### 2011-02-09 src/ANTS/Utilities/ReadWriteImage.h

@@ -1,0 +1,550 @@
+#ifndef __ReadWriteImage_h_
+#define __ReadWriteImage_h_
+
+#include <iostream>           
+#include <fstream>       
+#include <stdio.h>                    
+#include "itkVector.h"
+#include "itkImage.h"                   
+#include "itkImageFileWriter.h"                   
+#include "itkImageFileReader.h"       
+#include "itkImageRegionIteratorWithIndex.h"        
+#include "itkWarpImageFilter.h"
+//#include "itkInverseWarpImageFilter.h"
+#include "itkAffineTransform.h"
+#include "itkImageRegionIterator.h"
+#include "itkResampleImageFilter.h"
+#include "itkVectorIndexSelectionCastImageFilter.h"
+#include "itkImageRegionIteratorWithIndex.h"
+#include "itkNearestNeighborInterpolateImageFunction.h"
+#include "itkVectorIndexSelectionCastImageFilter.h"
+#include "itkLogTensorImageFilter.h"
+#include "itkExpTensorImageFilter.h"
+#include <sys/stat.h> 
+
+bool ANTSFileExists(std::string strFilename) { 
+  struct stat stFileInfo; 
+  bool blnReturn; 
+  int intStat; 
+
+  // Attempt to get the file attributes 
+  intStat = stat(strFilename.c_str(),&stFileInfo); 
+  if(intStat == 0) { 
+    // We were able to get the file attributes 
+    // so the file obviously exists. 
+    blnReturn = true; 
+  } else { 
+    // We were not able to get the file attributes. 
+    // This may mean that we don't have permission to 
+    // access the folder which contains this file. If you 
+    // need to do that level of checking, lookup the 
+    // return values of stat which will give you 
+    // more details on why stat failed. 
+    blnReturn = false; 
+  } 
+   
+  return(blnReturn); 
+}
+
+// Nifti stores DTI values in lower tri format but itk uses upper tri
+// currently, nifti io does nothing to deal with this. if this changes
+// the function below should be modified/eliminated.
+
+template <class TImageType>
+void NiftiDTICheck(itk::SmartPointer<TImageType> &target, const char *file, bool makeLower)
+{
+  typedef typename TImageType::PixelType PixType;
+  
+  return; 
+  
+  //typedef itk::ImageFileWriter<TImageType> Writer;
[…]

### 2011-02-09 src/ANTS/Utilities/itkArrayConvertPixelTraits.h

@@ -1,0 +1,357 @@
+/*=========================================================================
+
+  Program:   Insight Segmentation & Registration Toolkit
+  Module:    $RCSfile: itkArrayConvertPixelTraits.h,v $
+  Language:  C++
+  Date:      $Date: 2008-01-21 18:30:58 $
+  Version:   $Revision: 1.20 $
+
+  Copyright (c) Insight Software Consortium. All rights reserved.
+  See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+
+#ifndef __itkArrayConvertPixelTraits_h
+#define __itkArrayConvertPixelTraits_h
+
+#include "itkArray.h"
+#include "itkOffset.h"
+#include "itkCovariantVector.h"
+#include "itkVariableLengthVector.h"
+#include "itkVector.h"
+#include "itkPoint.h"
+#include "itkMatrix.h"
+
+namespace itk
+{
+  
+/** \class ArrayConvertPixelTraits
+ *  \brief Traits class used to by ConvertPixels to convert blocks of pixels.
+ *
+ *  TOutputPixelType is the destination type. The input type is infered
+ *  by the templated static function Convert.
+ *
+ *  This implementaion, does a simple assignment operator, so if you are
+ *  going from from a higher bit representation to a lower bit one (int to
+ *  char), you may want to specialize and add some sort of transfer function.
+ */
+template<typename PixelType>
+class ArrayConvertPixelTraits
+{
+public:
+  /** Determine the pixel data type. */
+  typedef typename PixelType::ComponentType ComponentType;
+
+  /** Return the number of components per pixel. */
+  static unsigned int GetNumberOfComponents() 
+  { return PixelType::GetNumberOfComponents();}
+  
+  /** Return the nth component of the pixel. */
+  static ComponentType GetNthComponent(int c, const PixelType& pixel) 
+  { return pixel.GetNthComponent(c); }
+
+  /** Set the nth component of the pixel. */
+  static void SetNthComponent(int c, PixelType& pixel, const ComponentType& v) 
+  { pixel.SetNthComponent(c, v); }
+
+  /** Return a single scalar value from this pixel. */
+  static ComponentType GetScalarValue(const PixelType& pixel)
[…]

### 2011-02-09 src/ANTS/Utilities/itkConvertArrayPixelBuffer.h

@@ -1,0 +1,64 @@
+/*=========================================================================
+
+  Program:   Insight Segmentation & Registration Toolkit
+  Module:    $RCSfile: itkConvertArrayPixelBuffer.h,v $
+  Language:  C++
+  Date:      $Date: 2010-08-14 23:24:28 $
+  Version:   $Revision: 0.00 $
+
+  Copyright (c) Insight Software Consortium. All rights reserved.
+  See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkConvertArrayPixelBuffer_h
+#define __itkConvertArrayPixelBuffer_h
+
+#include "itkConvertPixelBuffer.h"
+#include "itkArray.h"
+
+namespace itk
+{
+/**
+ * \class ConvertPixelBuffer
+ *  \brief Class to convert blocks of data from one type to another.
+ *
+ * ConvertPixelBuffer has a static method Convert().  It is used by
+ * itkImageFileReader to convert from an unknown type at run-time to the
+ * compile-time template pixel type in itkImageFileReader.  To work with
+ * complex pixel types like RGB and RGBA a traits class is used.
+ * OutputConvertTraits() is the traits class.  The default one used is
+ * DefaultConvertPixelTraits.  
+ *
+ */
+                                        
+template <typename InputPixelType, typename T, typename OutputConvertTraits>
+class ConvertPixelBuffer<InputPixelType, Array<T>, OutputConvertTraits> 
+{
+public:
+  /** Determine the output data type. */
+  typedef typename OutputConvertTraits::ComponentType OutputComponentType;
+
+  /** General method converts from one type to another. */
+  static void Convert(InputPixelType* inputData, 
+                      int inputNumberOfComponents, 
+                      Array<T>* outputData , 
+                      size_t size);
+  static void ConvertVectorImage(InputPixelType* inputData, 
+                      int inputNumberOfComponents, 
+                      Array<T>* outputData , 
+                      size_t size);
+private:
+  ConvertPixelBuffer();
+  ~ConvertPixelBuffer();
+};
+} //namespace ITK
+
[…]

### 2011-02-09 src/ANTS/Utilities/itkConvertVariableLengthVectorPixelBuffer.h

@@ -1,0 +1,64 @@
+/*=========================================================================
+
+  Program:   Insight Segmentation & Registration Toolkit
+  Module:    $RCSfile: itkConvertVariableLengthVectorPixelBuffer.h,v $
+  Language:  C++
+  Date:      $Date: 2010-08-18 23:24:28 $
+  Version:   $Revision: 0.00 $
+
+  Copyright (c) Insight Software Consortium. All rights reserved.
+  See ITKCopyright.txt or http://www.itk.org/HTML/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkConvertVariableLengthVectorPixelBuffer_h
+#define __itkConvertVariableLengthVectorPixelBuffer_h
+
+#include "itkConvertPixelBuffer.h"
+#include "itkVariableLengthVector.h"
+
+namespace itk
+{
+/**
+ * \class ConvertPixelBuffer
+ *  \brief Class to convert blocks of data from one type to another.
+ *
+ * ConvertPixelBuffer has a static method Convert().  It is used by
+ * itkImageFileReader to convert from an unknown type at run-time to the
+ * compile-time template pixel type in itkImageFileReader.  To work with
+ * complex pixel types like RGB and RGBA a traits class is used.
+ * OutputConvertTraits() is the traits class.  The default one used is
+ * DefaultConvertPixelTraits.  
+ *
+ */
+                                            
+template <typename InputPixelType, typename T, typename OutputConvertTraits>
+class ConvertPixelBuffer<InputPixelType, VariableLengthVector<T>, OutputConvertTraits>
+{
+public:
+  /** Determine the output data type. */
+  typedef typename OutputConvertTraits::ComponentType OutputComponentType;
+
+  /** General method converts from one type to another. */
+  static void Convert(InputPixelType* inputData, 
+                      int inputNumberOfComponents, 
+                      VariableLengthVector<T>* outputData , 
+                      size_t size);
+  static void ConvertVectorImage(InputPixelType* inputData, 
+                      int inputNumberOfComponents, 
+                      VariableLengthVector<T>* outputData , 
+                      size_t size);
+private:
[…]

### 2011-02-09 src/ANTS/Utilities/itkGaussianInterpolateImageFunction.h

@@ -1,0 +1,268 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkGaussianInterpolateImageFunction.h,v $
+  Language:  C++
+  Date:      $Date: 2009/07/01 12:59:34 $
+  Version:   $Revision: 1.5 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkGaussianInterpolateImageFunction_h
+#define __itkGaussianInterpolateImageFunction_h
+
+#include "itkInterpolateImageFunction.h"
+#include "vnl/vnl_erf.h"
+
+namespace itk
+{
+
+/** \class GaussianInterpolateImageFunction
+ * \brief Gaussianly interpolate an image at specified positions.
+ *
+ * GaussianInterpolateImageFunction linearly interpolates image intensity at
+ * a non-integer pixel position. This class is templated
+ * over the input image type and the coordinate representation type 
+ * (e.g. float or double).
+ *
+ * This function works for N-dimensional images.
+ *
+ * \ingroup ImageFunctions ImageInterpolators 
+ */
+template <class TInputImage, class TCoordRep = double>
+class ITK_EXPORT GaussianInterpolateImageFunction : 
+  public InterpolateImageFunction<TInputImage,TCoordRep> 
+{
+public:
+  /** Standard class typedefs. */
+  typedef GaussianInterpolateImageFunction Self;
+  typedef InterpolateImageFunction<TInputImage,TCoordRep> Superclass;
+  typedef SmartPointer<Self> Pointer;
+  typedef SmartPointer<const Self>  ConstPointer;
+  
+  /** Run-time type information (and related methods). */
+  itkTypeMacro(GaussianInterpolateImageFunction, InterpolateImageFunction);
+
+  /** Method for creation through the object factory. */
+  itkNewMacro(Self);  
+
+  /** OutputType typedef support. */
[…]

### 2011-02-09 src/ANTS/Utilities/itkLabelOverlapMeasuresImageFilter.h

@@ -1,0 +1,209 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkLabelOverlapMeasuresImageFilter.h,v $
+  Language:  C++
+  Date:      $Date: $
+  Version:   $Revision: $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkLabelOverlapMeasuresImageFilter_h
+#define __itkLabelOverlapMeasuresImageFilter_h
+
+#include "itkImageToImageFilter.h"
+#include "itkFastMutexLock.h"
+#include "itkNumericTraits.h"
+
+#include "itk_hash_map.h"
+
+namespace itk {
+
+/** \class LabelOverlapMeasuresImageFilter
+ * \brief Computes overlap measures between the set same set of labels of
+ * pixels of two images.  Background is assumed to be 0.
+ *
+ * \sa LabelOverlapMeasuresImageFilter
+ *
+ * \ingroup MultiThreaded
+ */
+template<class TLabelImage>
+class ITK_EXPORT LabelOverlapMeasuresImageFilter :
+    public ImageToImageFilter<TLabelImage, TLabelImage>
+{
+public:
+  /** Standard Self typedef */
+  typedef LabelOverlapMeasuresImageFilter                Self;
+  typedef ImageToImageFilter<TLabelImage,TLabelImage>    Superclass;
+  typedef SmartPointer<Self>                             Pointer;
+  typedef SmartPointer<const Self>                       ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  /** Runtime information support. */
+  itkTypeMacro( LabelOverlapMeasuresImageFilter, ImageToImageFilter );
+
+  /** Image related typedefs. */
+  typedef TLabelImage                                   LabelImageType;
+  typedef typename TLabelImage::Pointer                 LabelImagePointer;
+  typedef typename TLabelImage::ConstPointer            LabelImageConstPointer;
+
[…]

### 2011-02-09 src/ANTS/Utilities/itkLabeledPointSetFileReader.h

@@ -1,0 +1,137 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkLabeledPointSetFileReader.h,v $
+  Language:  C++
+  Date:      $Date: 2009/03/04 23:10:58 $
+  Version:   $Revision: 1.17 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkLabeledPointSetFileReader_h
+#define __itkLabeledPointSetFileReader_h
+
+#include "itkMesh.h"
+#include "itkMeshSource.h"
+
+#include "itkArray.h"
+#include "itkImage.h"
+#include "itkVectorContainer.h"
+
+#include <vector>
+
+namespace itk
+{
+
+/** \class LabeledPointSetFileReader
+ * \brief
+ * Reads a file and creates an itkMesh.
+ *
+ */
+template <class TOutputMesh>
+class LabeledPointSetFileReader 
+: public MeshSource<TOutputMesh>
+{
+public:
+  /** Standard "Self" typedef. */
+  typedef LabeledPointSetFileReader               Self;
+  typedef MeshSource<TOutputMesh>                 Superclass;
+  typedef SmartPointer<Self>                      Pointer;
+  typedef SmartPointer<const Self>                ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  /** Extract dimension from the output mesh. */
+  itkStaticConstMacro( Dimension, unsigned int,
+                       TOutputMesh::PointType::Dimension );
+
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( LabeledPointSetFileReader, MeshSource );
+
+  /** Hold on to the type information specified by the template parameters. */
+  typedef TOutputMesh                             OutputMeshType;
+  typedef typename OutputMeshType::MeshTraits     MeshTraits;
+  typedef typename OutputMeshType::Superclass     PointSetType;
[…]

### 2011-02-09 src/ANTS/Utilities/itkLabeledPointSetFileWriter.h

@@ -1,0 +1,154 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkLabeledPointSetFileWriter.h,v $
+  Language:  C++
+  Date:      $Date: 2009/03/04 23:10:58 $
+  Version:   $Revision: 1.17 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkLabeledPointSetFileWriter_h
+#define __itkLabeledPointSetFileWriter_h
+
+#include "itkMesh.h"
+
+#include "itkArray.h"
+#include "itkImage.h"
+#include "itkVectorContainer.h"
+
+namespace itk
+{
+/** \class LabeledPointSetFileWriter
+ * \brief
+ * Writes an itkMesh to a file in various txt file formats.
+ *
+ */
+template <class TInputMesh>
+class LabeledPointSetFileWriter : public Object
+{
+public:
+  /** Standard "Self" typedef. */
+  typedef LabeledPointSetFileWriter               Self;
+  typedef Object                                  Superclass;
+  typedef SmartPointer<Self>                      Pointer;
+  typedef SmartPointer<const Self>                ConstPointer;
+
+  /** Method for creation through the object factory */
+  itkNewMacro( Self );
+
+  /** Write the Input mesh to the Output file.
+   * Use either Update() or Write(). */
+  void Update( void );
+  void Write( void );
+
+  /** Extract dimension from the output mesh. */
+  itkStaticConstMacro( Dimension, unsigned int,
+                       TInputMesh::PointType::Dimension );
+
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro( LabeledPointSetFileWriter, Object );
+
+  /** Hold on to the type information specified by the template parameters. */
+  typedef TInputMesh                             InputMeshType;
+  typedef typename TInputMesh::Pointer           InputMeshPointer;
[…]

### 2011-02-09 src/ANTS/Utilities/itkSurfaceCurvatureBase.h

@@ -1,0 +1,296 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkSurfaceCurvatureBase.h,v $
+  Language:  C++
+  Date:      $Date: 2008/11/15 23:46:06 $
+  Version:   $Revision: 1.12 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef _SurfaceCurvatureBase_h
+#define _SurfaceCurvatureBase_h
+
+#include <vnl/vnl_matops.h>
+#include <vnl/algo/vnl_svd.h>
+#include <vnl/algo/vnl_symmetric_eigensystem.h>
+#include <vnl/vnl_cross.h>
+// #include <vnl/algo/vnl_rpoly_roots.h>
+#include "itkObject.h"
+#include "itkProcessObject.h"
+
+#include "itkVectorContainer.h"
+#include "itkCastImageFilter.h"
+
+namespace itk
+{
+
+
+/** \class SurfaceCurvatureBase
+ * 
+ * This class takes a surface as input and creates a local
+ * geometric frame for each surface point.  
+ *
+ * \note The origin of a neighborhood is always taken to be 
+ *       the first point entered into and the 
+ *       last point stored in the list.
+ */
+template < typename TSurface, unsigned int TDimension=3 >
+           class SurfaceCurvatureBase : public ProcessObject
+{
+public:
+
+  /** Standard class typedefs. */
+  typedef SurfaceCurvatureBase Self;
+  typedef ProcessObject Superclass;
+  typedef SmartPointer<Self> Pointer;
+  typedef SmartPointer<const Self>  ConstPointer;
+
+  /** Run-time type information (and related methods). */
+  itkTypeMacro(SurfaceCurvatureBase,ProcessObject);
+
+  /** Method for creation through the object factory. */
+  itkNewMacro(Self);
+
+  /** Image types. */
+  typedef TSurface               SurfaceType;
+
+  /** Image dimension. */
[…]

### 2011-02-09 src/ANTS/Utilities/itkVectorFieldGradientImageFunction.h

@@ -1,0 +1,237 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkVectorFieldGradientImageFunction.h,v $
+  Language:  C++
+  Date:      $Date: 2008/11/15 23:46:06 $
+  Version:   $Revision: 1.16 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkVectorFieldGradientImageFunction_h
+#define __itkVectorFieldGradientImageFunction_h
+
+#include "itkImageFunction.h"
+#include "itkVariableSizeMatrix.h"
+
+namespace itk
+{
+/** \class VectorFieldGradientImageFunction
+ *
+ */
+template <typename TInputImage,
+          typename TRealType = float,
+          typename TOutput = 
+            itk::VariableSizeMatrix<TRealType>
+>
+class ITK_EXPORT VectorFieldGradientImageFunction :
+    public ImageFunction<TInputImage, TOutput>
+{
+public:
+  /** Standard class typedefs. */
+  typedef VectorFieldGradientImageFunction Self;
+  typedef ImageFunction<TInputImage, TOutput> Superclass;
+  typedef SmartPointer<Self> Pointer;
+  typedef SmartPointer<const Self>  ConstPointer;
+
+  /** Method for creation through the object factory. */
+  itkNewMacro( Self );
+
+  /** Run-time type information (and related methods) */
+  itkTypeMacro( VectorFieldGradientImageFunction, ImageFunction );
+  
+  /** Extract some information from the image types.  Dimensionality
+   * of the two images is assumed to be the same. */
+  typedef TInputImage                                          InputImageType;
+  typedef typename InputImageType::PixelType                   VectorType;
+  typedef TOutput                                              MatrixType;
+
+  /** Declare typedefs of superclass */
[…]

### 2011-02-09 src/ANTS/Utilities/itkVectorGaussianInterpolateImageFunction.h

@@ -1,0 +1,290 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkVectorGaussianInterpolateImageFunction.h,v $
+  Language:  C++
+  Date:      $Date: 2009/07/01 12:59:34 $
+  Version:   $Revision: 1.5 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkVectorGaussianInterpolateImageFunction_h
+#define __itkVectorGaussianInterpolateImageFunction_h
+
+#include "itkInterpolateImageFunction.h"
+#include "vnl/vnl_erf.h"
+#include "itkImageRegionConstIteratorWithIndex.h"
+namespace itk
+{
+
+/** \class VectorGaussianInterpolateImageFunction
+ * \brief Gaussianly interpolate an image at specified positions.
+ *
+ * VectorGaussianInterpolateImageFunction linearly interpolates image intensity at
+ * a non-integer pixel position. This class is templated
+ * over the input image type and the coordinate representation type 
+ * (e.g. float or double).
+ *
+ * This function works for N-dimensional images.
+ *
+ * \ingroup ImageFunctions ImageInterpolators 
+ */
+template <class TInputImage, class TCoordRep = double>
+class ITK_EXPORT VectorGaussianInterpolateImageFunction : 
+  public InterpolateImageFunction<TInputImage,TCoordRep> 
+{
+public:
+  /** Standard class typedefs. */
+  typedef VectorGaussianInterpolateImageFunction Self;
+  typedef InterpolateImageFunction<TInputImage,TCoordRep> Superclass;
+  typedef SmartPointer<Self> Pointer;
+  typedef SmartPointer<const Self>  ConstPointer;
+  
+  /** Run-time type information (and related methods). */
+  itkTypeMacro(VectorGaussianInterpolateImageFunction, InterpolateImageFunction);
+
[…]

### 2011-02-09 src/ANTS/Utilities/itkVectorImageFileWriter.h

@@ -1,0 +1,203 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkVectorImageFileWriter.h,v $
+  Language:  C++
+  Date:      $$
+  Version:   $ $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even 
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR 
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef __itkVectorImageFileWriter_h
+#define __itkVectorImageFileWriter_h
+
+#include "itkProcessObject.h"
+#include "itkImageIOBase.h"
+#include "itkExceptionObject.h"
+#include "itkSize.h"
+#include "itkImageIORegion.h"
+
+namespace itk
+{
+
+/** \brief Base exception class for IO problems during writing. */
+class VectorImageFileWriterException : public ExceptionObject 
+{
+public:
+  /** Run-time information. */
+  itkTypeMacro( VectorImageFileWriterException, ExceptionObject );
+
+  /** Constructor. */
+  VectorImageFileWriterException(const char *file, unsigned int line, 
+                           const char* message = "Error in IO",
+                           const char* loc = "Unknown" ) : 
+    ExceptionObject(file, line, message, loc)
+  {
+  }
+
+  /** Constructor. */
+  VectorImageFileWriterException(const std::string &file, unsigned int line, 
+                           const char* message = "Error in IO",
+                           const char* loc = "Unknown" ) :
+    ExceptionObject(file, line, message, loc)
+  {
+  }
+};
+
+/** \class VectorImageFileWriter
+ * \brief Writes the deformation field as component images files.
+ *
+ * \sa VectorImageFileWriter
+ * \sa ImageSeriesReader
+ * \sa ImageIOBase
+ *
+ * \ingroup IOFilters 
+ */
+template <class TVectorImage, class TImage>
+class ITK_EXPORT VectorImageFileWriter : 
+  public ProcessObject
+{
+public:
+  /** Standard class typedefs. */
+  typedef VectorImageFileWriter                  Self;
+  typedef ProcessObject    Superclass;
[…]

### 2011-02-09 src/ANTS/Utilities/itkWarpImageMultiTransformFilter.h

@@ -1,0 +1,328 @@
+/*=========================================================================
+
+  Program:   Advanced Normalization Tools
+  Module:    $RCSfile: itkWarpImageMultiTransformFilter.h,v $
+  Language:  C++
+  Date:      $Date: 2009/01/08 21:36:48 $
+  Version:   $Revision: 1.18 $
+
+  Copyright (c) ConsortiumOfANTS. All rights reserved.
+  See accompanying COPYING.txt or 
+ http://sourceforge.net/projects/advants/files/ANTS/ANTSCopyright.txt for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notices for more information.
+
+=========================================================================*/
+#ifndef ITKWARPIMAGEMULTITRANSFORMFILTER_H_
+#define ITKWARPIMAGEMULTITRANSFORMFILTER_H_
+
+#include "itkImageToImageFilter.h"
+#include "itkInterpolateImageFunction.h"
+#include "itkLinearInterpolateImageFunction.h"
+#include "itkVectorLinearInterpolateImageFunction.h"
+#include "itkVectorGaussianInterpolateImageFunction.h"
+#include "itkPoint.h"
+#include "itkFixedArray.h"
+#include "itkRecursiveGaussianImageFilter.h"
+#include <list>
+
+namespace itk
+{
+
+/** \class WarpImageMultiTransformFilter
+ * \brief Warps an image using an input deformation field.
+ *
+ * WarpImageMultiTransformFilter warps an existing image with respect to
+ * a given deformation field.
+ *
+ * A deformation field is represented as a image whose pixel type is some
+ * vector type with at least N elements, where N is the dimension of
+ * the input image. The vector type must support element access via operator
+ * [].
+ *
+ * The output image is produced by inverse mapping: the output pixels
+ * are mapped back onto the input image. This scheme avoids the creation of
+ * any holes and overlaps in the output image.
+ *
+ * Each vector in the deformation field represent the distance between
+ * a geometric point in the input space and a point in the output space such
+ * that:
+ *
+ * \f[ p_{in} = p_{out} + d \f]
+ *
+ * Typically the mapped position does not correspond to an integer pixel
+ * position in the input image. Interpolation via an image function
+ * is used to compute values at non-integer positions. The default
[…]

### 2011-02-10 doc/MVSeg.tex
ENH: updating atropos and MVSeg doc

@@
 
+\subsection{EM image segmentation, briefly} 
+Consider a classic image segmentation problem where one wants to
+automatically label each image voxel into one of three classes, $A, B, C$.  We
+use a Bayesian model for the probability of each part of the image being a member of a
+specific class, $Q$, 
+$$
+P( Q | i ) \propto P ( i | Q ) P( Q ) 
+$$
+where $i$ is the intensity.
+We then model 
+$$ P( i | Q )  \propto \exp( - \| i - \mu_Q \|^2 / \sigma_Q^2 ) $$ 
+where $\mu_Q , \sigma_Q$ are the parameters to a Gaussian describing
+the intensity distribution of class $Q$.  For now, let us assume that
+$P(Q)=\pi_q$ where $\pi_q$ is some constant between 0 and 1.  EM will use this
+type of model for each class and, given initial values for the
+parameters (the means and variances for each class's Gaussian), find
+the maximum a posteriori solution to the class labels and parameters.
+It's basically a gradient descent algorithm.  In this case, one simple EM
+solution will be 
+\begin{enumerate}
+\item Take initial values for the class parameters from the user.
+\item For each voxel in the image, compute the posterior probability
+  for each label, that is, $P(A|i), P(B|i), P(C|i)$.  Assign the voxel
+  to the label with maximal probability.  This is a synchronous
+  updating scheme.  
+\item For each class, recompute the parameters to the Gaussian by 
+$ \mu_Q \leftarrow \frac{\sum_{j=1}^N i P(Q|i)}{\sum_{j=1}^N P(Q|i)}$
+and $ \sigma_Q \leftarrow \frac{\sum_{j=1}^N (i-\mu_Q)^2 P(Q|i)}{\sum_{j=1}^N P(Q|i)}$.
+\item If converged, stop, otherwise go to 2.  
+\end{enumerate}
+This is the basic idea of EM.  Seems simple.  However, it gets quickly complicated when
+you begin to specify EM for a problem domain.  For instance, for
+convergence guarantees, one would use iterated conditional modes (ICM)
+in step 2.  A scientist should use his or her domain knowledge to
+define the prior probabilities $P(Q)$ and to define the distance
+(here, $\| i -\mu_Q \|$) and/or likelihood model in
+$P(i|Q)$.  A common goal is to enforce local sets of voxels to have the same label, if possible.  Markov
+random field frameworks provide many options for achieving label regularity 
+in image segmentation \citep{Besag1986}.  One might also take care in estimating the mixing proportions that are
+used in a standard finite mixture model.  This is all discussed in more detail
+in the Atropos paper.  In the current work, we are concerned mainly
+with the likelihood model and the input data representations which should be
+chosen such that they are suitable for deriving biological and/or anatomical meaning from DTI.
+
+\subsection{Diffusion tensor primer} A diffusion tensor (DT) is a
+symmetric positive definite matrix.  Diffusion tensor imaging derives
+a dense field of DT estimates at each position in a three-dimensional
+imaging volume by measuring the amount of magnetically-induced water
+diffusion in each of $N$ directions.  Directions, here, refers to
+positions on the sphere (e.g. longitude and latitude but only in the
+northern hemisphere).  Each tensor may be described uniquely as
+$$D=\sum_i \lambda_i e_i^T e_i $$ where $e_i$ is a unit 3-vector and
+$\lambda_i$ is a scalar in 0 to 1.  The $\lambda_i$ defines the
+relative amount of diffusion along the axis defined by $e_i$.  A
+highly anisotropic tensor will have $\lambda_1 >> \lambda_{2,3} $, that
+is, the prinicipal eigenvector of the tensor will have much higher
+diffusion than the other eigenvectors.   The variance of the
+$\lambda_i$ is known as the fractional anisotropy (FA) and their mean is
+known as the mean diffusion, apparent diffusion coefficient (ADC) or
+the trace (actually the sum). Below are some of other measurements derived from DTI:
+\begin{description}
+\item[RBG:] The principal direction of the tensor can be visualized in
+  RGB space.  Some segmentation methods rely on this feature 
+  modulated by the FA.  That is, the RBG image voxel may be defined as
+  $\vec{rgb}=[R=255~\text{FA}~e_1^1, G=255~\text{FA}~e_1^2, B=255~\text{FA}~e_1^3$].
+\item[FA:] Fractional anisotropy derived from the tensor.  This is
+  equal to the variance of the eigenvalues of the tensor matrix.  It
+  is the most common measure used in DTI studies, along with
+  tractography. 
+\item[Tensor shape:]  This is determined by the tensor eigenvalues.  As
+  neuroscientists interested in tracts, we may be concerned with how
+  collections of eigenvalues are distributed along tracts and how
+  these distributions differentiate tracts or tissues.
+\item[Tensor orientation:] The principal
+  eigenvector can point to any position on the upper hemisphere.  As
+  communicated by the power of the RGB display of tensor information,
+  this is a powerful feature. 
+\item[Tractography:] White matter tracts provide a physical connection
+  between and within cortical regions.  At the coarse scale of DTI, we mostly
+  look at connections between cortical regions.  A simple way to
+  define a tract from DTI is to trace a curve that follows the principal eigenvector until
+  some stopping criterion based on curvature or FA is reached. 
+\end{description}
+So, in this work, we'd like to use these different features to help us
+automatically differentiate tissues and tracts in not only population
+average images but also individual images.  Being able to compute
+these types of measures will let us do specific, novel studies of
+neural architecture across species, gender, age and disease.
+
+\subsection{A quick tensor segmentation example}
+atropos seg example with dti
+
+two questions --- one of representation , one of distance.  that is,
+distances on what?
+\begin{figure}
+\begin{center}
+\begin{tabular}{c}
+\includegraphics[width=6in]{Figures/macaque_rgb.pdf}
+\end{tabular}
+\caption{We perform an RGB-based segmentation of the macaque
+  population average tensor image.  At left is the manually placed
+  labels, as performed in ITK-Snap.  At right is the Atropos result.
+  You may repeat this study by labeling the macaque image with labels
+  1, 2 and 3 (R, G, B) similarly to this example.  You then
+  save the segmentation and run the script entitled MVSeg/src/scripts/seg\_dti\_manual\_init\_atropos.}
+\label{fig:example1}
+\end{center}
+\end{figure}
+
 \section{Data}
@@
 \section{Methods} Our goal will be to provide reproducible
-measurements derived from multivariate segmentation.
-\begin{description}
-\item[RBG:] The principal direction of the tensor can be visualized in
-  RGB space.  Some segmentation methods rely on this feature 
-  modulated by the FA.
-\item[FA:] Fractional anisotropy derived from the tensor.  This is
-  equal to the variance of the eigenvalues of the tensor matrix.  It
-  is the most common measure used in DTI studies, along with
-  tractography. 
-\item[Tractography:] White matter tracts provide a physical connection
-  between and within cortical regions.  At the coarse scale of DTI, we mostly
-  look at connections between cortical regions.  A simple way to
-  define a tract from DTI is to follow the principal eigenvector until
-  some stopping criterion based on curvature or FA is reached. 
-\end{description}
-An EM segmentation method can use any of these features.  One project
-that may be similar to our own is here
+measurements derived from multivariate segmentation.  An EM
+segmentation method can use any of the DTI features we discussed
+above.  One project that may be similar to our own is here
 \url{http://www.nitrc.org/projects/dots} though there is no clear
@@
 
+\subsection{Ideas}
+One idea to explore is nonparametric joint histogram based likelihood
+models for tensor shape and orientation.  $J_{orientation} = f(i,j)/
+\sum_{ij} f(i,j) $ where $f(i,j)$ is the frequency histogram of the
+upper hemisphere describing orientation.  two parameters: theta and
+psi.  $J_{shape} = g(i,j)/ \sum_{ij} g(i,j) $ where $g(i,j)$ is the
+frequency histogram describing the co-occurrence of first and second
+eigenvalues.  These would become our likelihood models (recalling
+$P(i|Q)$) and would be initialized by priors deformed from previously
+segmented images.
+
 We can also explore the effect of simulated and/or deterministic
 annealing which is a hidden feature of Atropos.
+
+\subsection{Code}
+The class called {\em antsHistogramParzenWindowsListSampleFunction}
+implements a non-parametric histogram based likelihood function.  It is located in MVSeg/src/ANTS/ImageSegmentation.
+One might implement a class called  {\em
+  antsJointHistogramParzenWindowsListSampleFunction} that could
+implement the idea above.   Or one could follow the example {\em
+  antsLogEuclideanGaussianListSampleFunction} to implement a
+parametric likelihood.  Each class should be tested as a stand-alone
+method and should be verified to work before being included in the
+segmentation functionality.
 

### 2011-02-10 src/ANTS/ImageSegmentation/antsAtroposSegmentationImageFilter.h

@@ -120,10 +120,7 @@ public:
     <SampleType, RealType, RealType>                  LikelihoodFunctionType;
   typedef typename LikelihoodFunctionType::Pointer    LikelihoodFunctionPointer;
   typedef typename LikelihoodFunctionType::
-    ListSampleWeightArrayType                         WeightArrayType;
-
-  typedef std::vector<LabelType>                      PartialVolumeLabelSetType;
-  typedef std::vector<PartialVolumeLabelSetType>      PartialVolumeClassesType;
+    WeightArrayType                                   WeightArrayType;
 
   /** Outlier handling typedefs */
   typedef ants::Statistics::
@@ -161,53 +158,16 @@ public:
   // ivars Set/Get functionality
 
   /**
-   * Set the number of tissue classes which is clamped from below at 2.
+   * Set the number of segmentation classes which is clamped from below at 2.
    * Default = 3.
    */
-  itkSetClampMacro( NumberOfTissueClasses, LabelType, 2,
+  itkSetClampMacro( NumberOfClasses, LabelType, 2,
     NumericTraits<unsigned int>::max() );
 
   /**
    * Get the number of segmentation classes.
    */
-  itkGetConstMacro( NumberOfTissueClasses, unsigned int );
-
-  /**
-   * Set the partial-volume-label set one at a time.
-   */
-  void AddPartialVolumeLabelSet( PartialVolumeLabelSetType );
-
-  /**
-   * Get the number of partial volume classes
-   */
-  itkGetConstMacro( NumberOfPartialVolumeClasses, unsigned int );
-
-  /**
-   * The user can specify whether or not to use the partial volume likelihoods,
-   * in which case the partial volume class is considered separate from the
-   * tissue classes.  Alternatively, one can use the MRF only to handle
-   * partial volume in which case, partial volume voxels are not considered
-   * as separate classes.
-   */
-  itkSetMacro( UsePartialVolumeLikelihoods, bool );
-
-  /**
-   * The user can specify whether or not to use the partial volume likelihoods,
-   * in which case the partial volume class is considered separate from the
-   * tissue classes.  Alternatively, one can use the MRF only to handle
-   * partial volume in which case, partial volume voxels are not considered
-   * as separate classes.
-   */
-  itkGetConstMacro( UsePartialVolumeLikelihoods, bool );
-
-  /**
-   * The user can specify whether or not to use the partial volume likelihoods,
-   * in which case the partial volume class is considered separate from the
[…]

### 2011-02-10 doc/MVSeg.tex
ENH: updating atropos and MVSeg doc

@@
 
+So, we have two questions---one of representation , one of distance.  That is,
+what distances should we apply to what represenation of DTI data?
+
 \subsection{A quick tensor segmentation example}
-atropos seg example with dti
-
-two questions --- one of representation , one of distance.  that is,
-distances on what?
+Figure~\ref{fig:macaque} shows a population average of DTI data for
+the macaque monkey.  We segment the image by manually seeding each of
+3 classes (R, G and B) by hand and then calling the Atropos
+algorithm.  This parcellates the white matter within the automatically
+generated mask into regions where the tract directionality is
+primarily left-right (red), anterior-posterior (green) and
+inferior-superior (blue).  While this is somewhat anatomically
+meaningful, we would like to improve upon these results.  See, for
+instance, Figure~\ref{fig:example1} which parcellates the brain into
+anatomically determined regions.  
 \begin{figure}
@@
   save the segmentation and run the script entitled MVSeg/src/scripts/seg\_dti\_manual\_init\_atropos.}
-\label{fig:example1}
+\label{fig:macaque}
 \end{center}

### 2011-02-11 doc/MVSeg.tex
Added a few things to the DT section

@@
 a dense field of DT estimates at each position in a three-dimensional
-imaging volume by measuring the amount of magnetically-induced water
-diffusion in each of $N$ directions.  Directions, here, refers to
-positions on the sphere (e.g. longitude and latitude but only in the
-northern hemisphere).  Each tensor may be described uniquely as
-$$D=\sum_i \lambda_i e_i^T e_i $$ where $e_i$ is a unit 3-vector and
+imaging volume by measuring the amount of water
+diffusion along each of $N$ directions. Directions, here, refers to
+undirected axes on the unit sphere - a measurement with gradient direction $\vec{x}$ is equivalent to a measurement with
+gradient direction $-\vec{x}$. Because of this symmetry, we can describe any measurement
+using only positions on the unit hemisphere. (e.g., longitude and latitude can describe any axis through the Earth's center, using only 
+latitudes in the northern hemisphere).
+
+Each tensor may be described uniquely as
+$$D= \frac1{\mathrm{Tr}(D)}\sum_i \lambda_i e_i^T e_i $$ where $e_i$ is a unit 3-vector and
 $\lambda_i$ is a scalar in 0 to 1.  The $\lambda_i$ defines the
@@
 highly anisotropic tensor will have $\lambda_1 >> \lambda_{2,3} $, that
-is, the prinicipal eigenvector of the tensor will have much higher
+is, the principal eigenvector of the tensor will have much higher
 diffusion than the other eigenvectors.   The variance of the
 $\lambda_i$ is known as the fractional anisotropy (FA) and their mean is
-known as the mean diffusion, apparent diffusion coefficient (ADC) or
-the trace (actually the sum). Below are some of other measurements derived from DTI:
+known as the mean diffusion or apparent diffusion coefficient (ADC). Sometimes you will see the
+tensor trace (sum of the eigenvalues) reported instead of the ADC. Below are some of other measurements derived from DTI:
 \begin{description}
-\item[RBG:] The principal direction of the tensor can be visualized in
+\item[RGB:] The principal direction of the tensor can be visualized in
   RGB space.  Some segmentation methods rely on this feature 
-  modulated by the FA.  That is, the RBG image voxel may be defined as
+  modulated by the FA.  That is, the RGB image voxel may be defined as
   $\vec{rgb}=[R=255~\text{FA}~e_1^1, G=255~\text{FA}~e_1^2, B=255~\text{FA}~e_1^3$].
 \item[FA:] Fractional anisotropy derived from the tensor.  This is
-  equal to the variance of the eigenvalues of the tensor matrix.  It
+  equal to the normalized variance of the eigenvalues of the tensor matrix.  It
   is the most common measure used in DTI studies, along with
@@
   collections of eigenvalues are distributed along tracts and how
-  these distributions differentiate tracts or tissues.
+  these distributions differentiate tracts or tissues. FA is one shape measure, you can define others to describe
+  things like linearity ( $\lambda_1 >> \lambda_{2,3} $) vs planarity ( $\lambda_1 \simeq \lambda_2 >> \lambda_3 $)
+  of the ellipsoid tensor representation.
 \item[Tensor orientation:] The principal

### 2011-02-18 src/ANTS/Documentation/Tex/ants.tex
revert to svn from today

%
% Complete documentation on the extended LaTeX markup used for Insight
% documentation is available in ``Documenting Insight'', which is part
% of the standard documentation for Insight.  It may be found online
% at:
%
%     http://www.itk.org/

\documentclass{InsightArticle}

% \usepackage[dvips]{graphicx}
%\usepackage{listing}						
\usepackage{listings}	
\usepackage{wrapfig}
\usepackage{amssymb,amsmath}
\usepackage{multirow,booktabs,ctable,array}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  hyperref should be the last package to be loaded.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[dvips,
bookmarks,
bookmarksopen,
backref,
colorlinks,linkcolor={blue},citecolor={blue},urlcolor={blue},
]{hyperref}

\graphicspath{{./Figures/}}
\newcommand{\fv}{\mbox{\boldmath $f$}}
\newcommand{\Iv}{\mbox{\boldmath $I$}}
\newcommand{\xv}{\mbox{\boldmath $x$}}
\newcommand{\velocity}{\mbox{\boldmath $v$}}
\newcommand{\velocityinv}{\mbox{\boldmath $v$}^{-1}}
\newcommand{\Velocityinv}{\mbox{\boldmath $V$}^{-1}}
\newcommand{\Velocity}{\mbox{\boldmath $V$}}
\newcommand{\welocity}{\mbox{\boldmath $w$}}
\newcommand{\spatialimagegradient}{\mbox{\boldmath $I$}\mbox{$_{x}$}}
\newcommand{\perpspatialimagegradient}{\mbox{\boldmath $I$}\mbox{$^{\perp}_{x}$}}
\newcommand{\temporalimagederivative}{\mbox{$I_t$}}
\newcommand{\finitestrain}{\mbox{\boldmath $\varepsilon^{\ast}$}}
\newcommand{\smallstrain}{\mbox{\boldmath $\varepsilon$}}
\newcommand{\dispgradient}{{\bf \nabla \displace}}
\newcommand{\displace}{{\bf u}} 
\newcommand{\Id}{\text{\bf Id}}
\newcommand{\ode}{{\em O.D.E.}}
\newcommand{\odes}{{\em O.D.E.}s}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\phiinv}{\phi^{-1}}
\newcommand{\psiinv}{\psi^{-1}}
\newcommand{\dM}{DM}
\newcommand{\DM}{Diffeomorphometry}
\newcommand{\diff}{diffeomorphism}
\newcommand{\Diff}{Diffeomorphism}
\newcommand{\orbit}{\mathcal{O}}
\newcommand{\avg}{\mathcal{A}}
\newcommand{\avgn}{\mathcal{A}_n}
\newcommand{\avgna}{\mathcal{A}^a_{n}}
\newcommand{\nset}{\{ J_i \}_n}
\newcommand{\bari}{\bar{I}}
\newcommand{\bart}{\bar{t}}
\newcommand{\jac}{\mathcal{J}}
\newcommand{\pert}{\mbox{\boldmath $w$}}
\newcommand{\barpert}{\bar{\mbox{\boldmath $w$}}}
\newcommand{\half}{0.5}

\newcommand{\X}{{\bf X}}
\newcommand{\x}{{\bf x}}
\newcommand{\Z}{{\bf Z}}
\newcommand{\z}{{\bf z}}
\newcommand{\p}{{\bf p}}
\newcommand{\Y}{{\bf Y}}
\newcommand{\y}{{\bf y}}
\newcommand{\disp}{{\bf u}}
\newcommand{\ytild}{\tilde{{\bf y}}}
\newcommand{\q}{{\bf q}}
\newcommand{\surf}{\mathcal{S}}
\newcommand{\phij}{\phi_{ij}}
\newcommand{\aphij}{\bar{\phi}_{j}}
\newcommand{\apsij}{\bar{\psi}_{j}}
\newcommand{\aphi}{\bar{\phi}}
\newcommand{\aphiinv}{\bar{\phi}^{-1}}

\newcommand{\domain}{\Omega}
\newcommand{\meanshape}{\bf \bar{x}}
\newcommand{\g}{\mbox{\boldmath $g$}}
\newcommand{\h}{\mbox{\boldmath $h$}}
\newcommand{\yv}{\mbox{\boldmath $y$}}

\newcommand{\group}{Diff}

\newcommand{\evol}{{E_{\text{Vol}}}}
\newcommand{\epair}{{E_{\text{Pair}}}}
\newcommand{\esec}{{E_{S}}}
\newcommand{\inv}{^{-1}}
\newcommand{\vinit}{ \velocity^0_{ij}(\x,\tau=0) }
\newcommand{\avinit}{ \bar{\velocity}^0_{j}(\x,t) }
\newcommand{\avinitz}{ \bar{\velocity}^0_{j}(\x,t_a) }
\newcommand{\gz}{\nabla_{\z}}

%  This is a template for Papers to the Insight Journal. 
%  It is comparable to a technical report format.

% The title should be descriptive enough for people to be able to find
% the relevant document. 
\title{Advanced Normalization Tools (ANTS)}

% 
% NOTE: This is the last number of the "handle" URL that 
% The Insight Journal assigns to your paper as part of the
% submission process. Please replace the number "1338" with
% the actual handle number that you get assigned.
%
%\newcommand{\IJhandlerIDnumber}{}

% Increment the release number whenever significant changes are made.
% The author and/or editor can define 'significant' however they like.
\release{1.5}

% At minimum, give your name and an email address.  You can include a
% snail-mail address if you like.
\author{Brian B. Avants, Nick Tustison and Gang Song}
\authoraddress{Penn Image Computing And Science Laboratory\\
University of Pennsylvania}

\begin{document}

%
% Add hyperlink to the web location and license of the paper.
% The argument of this command is the handler identifier given
% by the Insight Journal to this paper.
% 
%\IJhandlefooter{\IJhandlerIDnumber}

\ifpdf
\else
   %
   % Commands for including Graphics when using latex
   % 
   \DeclareGraphicsExtensions{.eps,.jpg,.gif,.tiff,.bmp,.png}
   \DeclareGraphicsRule{.jpg}{eps}{.jpg.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.gif}{eps}{.gif.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.tiff}{eps}{.tiff.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.bmp}{eps}{.bmp.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.png}{eps}{.png.bb}{`convert #1 eps:-}
\fi

\maketitle

\ifhtml
\chapter*{Front Matter\label{front}}
\fi

%Setup listing environment
\definecolor{listcomment}{rgb}{0.0,0.5,0.0}
\definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
\definecolor{listbackground}{gray}{0.965}

\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listbackground}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = left,
        tabsize = 2,
        language=[ISO]C++,
        floatplacement=!h
        }

% The abstract should be a paragraph or two long, and describe the
% scope of the document.
\begin{abstract}
\noindent We provide examples and highlights of Advanced Normalization Tools (ANTS) that 
address practical problems in real data.  
\end{abstract}

%\IJhandlenote{\IJhandlerIDnumber}

\tableofcontents

\section{Introduction}
The ANTS framework provides open-source functionality for deformable
normalization with large deformations, as shown in
figure~\ref{fig:chalf}.  Small deformation mappings and segmentation 
tools are also available. 
Independent evaluation of ANTS normalization tools, 
applied to ``control'' data, 
placed the toolkit as a top performer amongst 14
methods \cite{Klein2009}.  Developer evaluation showed stronger differences 
with other methodology in neurodegenerative neuroimaging data, 
where large deformation is required \cite{Avants2008b}. 
ANTS serves as both a base for further
algorithm development and also as an application-oriented toolkit.
ANTS enables diffeomorphic normalization with a variety of
transformation models, optimal template construction, multiple types
of diffeomorphisms, multivariate similarity metrics, diffusion tensor
processing and warping, image segmentation with and without priors and
measurement of cortical thickness from probabilistic segmentations.
The normalization tools, alone, provide a near limitless range of
functionality and allow the user to develop customized objective
functions.  Objective functions in ANTS are of the form: $$
\text{Deformation Cost} + \text{Data Terms},$$ and the command
line reflects this balance of two terms.  As mentioned above, the
data term may combine multiple different measures of
similarity that are optimized in parallel, for instance, image similarity 
and landmark terms.  This document seeks to
provide a practical overview of basic functionality and some of the common use cases that
users seek.
Additional information is available online -- see 
\href{www.picsl.upenn.edu/ANTS}{ANTS Homepage : www.picsl.upenn.edu/ANTS}.
For compilation details, see: \href{http://picsl.upenn.edu/ANTS/download.php}{Compile and download: http://picsl.upenn.edu/ANTS/download.php}.
The most important ANTS programs are described below. 
\footnote{This document is a work in progress. Please check for updates with each release.}
\begin{figure}
\includegraphics[width=0.33\textwidth]{Figures/grid1100.pdf} 
\includegraphics[width=0.33\textwidth]{Figures/grid1110.pdf} 
\includegraphics[width=0.33\textwidth]{Figures/grid1119.pdf} 
\itkcaption[ANTS Large Deformation]{The original goal of ANTS was to
  develop public, open source large deformation image registration.
  This is a classic example showing the progress of deforming a half
  C to a full C along a geodesic diffeomorphism.  The deforming 
  grid accompanies each deformed image.}
\vspace{-0.1in}
\label{fig:chalf}
\end{figure}

\subsection{The ANTS Executable}
The ANTS program itself is the central program encapsulating
normalization/registration functionality.  Its main output is an
affine transform file and a deformation field, potentially with
inverse.  Options to ANTS allows the user to navigate the similarity
and transformation options that are available.  ANTS takes multiple
similarity and optimization criteria as options.  The program is
wrapped in \verb ants.sh ~for normalization with ``out of the box''
parameters and in \verb buildtemplateparallel.sh ~for 
computationally distributed optimal template construction.
You can also call  \verb ANTS  \verb --help  to get basic usage.

\subsection{The WarpImageMultiTransform Executable}
The WarpImageMultiTransform program applies ANTS mappings to images.  One may apply an 
arbitrarily long series of transformations to an image through this program.  Thus, it enables one 
to compose a series of affine and deformable mappings -- and/or their inverses -- such that one 
may avoid multiple interpolations of a single image.  This program is used internally in ANTS 
optimization to avoid ever using more than one interpolation.  The use-NN option applies nearest 
neighbor interpolation, otherwise linear interpolation is used. 
You can call  \verb WarpImageMultiTransform \verb --help to get basic usage.

Also, please take a look at 
\href{http://picsl.upenn.edu/ANTS/ioants.php}{http://picsl.upenn.edu/ANTS/ioants.php} 
and the section below on I/O.

\subsection{I/O Data Formats In ANTS}
ANTS support both 2D and 3D images. 
Since ANTS is implemented in ITK, 
it is able to read and write the popular data formats supported through ITK. 
There are four basic types of data for ANTS.
\subsubsection{Image Volumes}
ANTS support both 2D and 3D images, including
\begin{itemize}
\item Nifti (.nii, .nii.gz)
\item Analyze (.hdr + .img / .img.gz)
\item MetaImage (.mha)
\item Other formats through itk::ImageFileWriter / itk::ImageFileWriter such as jpg, tiff, etc.  See ITK documentation. 
\end{itemize}
\subsubsection{Affine Transformation File}
ANTS uses the itk::TransformFileReader / itk::TransformFileWriter to handle affine transformation files. 
These are text files recording the matrix and offset for the affine transform {\bf FIXME NEED MORE DETAIL on definition and geometry of affine matrix}.
The interpretation depends on the header but generally it breaks down like this:

\#Insight Transform File V1.0

\# Transform 0

Transform: MatrixOffsetTransformBase\_double\_3\_3

Parameters:  $ a~b~c~d~e~f~g~h~i~j~k~l $

FixedParameters: $ M~N~O $

The meaning of these different terms:

The matrix component (rotation, scaling, shearing all composed together):

$ \begin{matrix}
a & b & c \\
d & e & f  \\
g & h & i \\ 
\end{matrix} $

is followed by  $~j~k~l~$ --- this is the translation.

$M~N~O$  is the center of rotation .

You can use the --initial-affine  and    --continue-affine  options to control how this is used in registration.

\subsubsection{Deformation Field File}
ANTS writes the deformation field into a multi-component image (latest
release).  Thus, 
one refers to the warp file as ABCD.nii.gz which, in 3D, contains 3 components.  If you want to 
see the details of how a warp is achieved, look at the code in itkWarpImageMultiTransform.txx. 
One issue to be aware of is that if you are directly indexing a vector component image in 
voxel coordinates, then you need to apply the image's direction matrix to the vector.  
An example of this is in ANTSJacobian.cxx, the TransformVector function.

\subsubsection{Labeled Point Sets}
There are three formats supported
\begin{itemize}
\item  VTK format: point + scalar label in a .vtk file. {\bf FIXME NEED MORE DETAIL -- what type of vtk data ?  polydata?}.
\item  Labeled point sets are saved as an image volume with the intensity = 1,2,3 for label = 1,2,3. For example, 
figure~\ref{fig:frown} shows 2D images representing a 3-class labeled point set. 
\item Avants Format. In this simple format, the point sets are saved as a text file. The first and last line is always 0 0 0 0. Each line in between for a point as X Y Z L (L is the label.)
\end{itemize}
The names and formats of the output files are both specified by -o option to the ANTS executable. 
For example, for in 3D with -o RegA2B.nii.gz , the typical output files are:
Affine transform file: 
RegA2BAffine.txt 
Warping deformation field: 
RegA2BWarp.nii.gz.
Inverse warping deformation field:
RegA2BInverseWarp.nii.gz.  
To apply the affine transform and the deformation field to the moving image, use WarpImageMultiTransform. 
For example: to apply the transform to the moving image:

WarpImageMultiTransform 2 A.nii.gz ADeformed.nii.gz -R B.nii.gz RegA2BWarp.nii.gz RegA2BAffine.txt

to apply the inverse transform to the fixed image:

WarpImageMultiTransform 2 B.nii.gz BDeformed.nii.gz -R A.nii.gz -i RegA2BAffine.txt RegA2BInverseWarp.nii.gz

\subsection{The ImageMath Executable} This is a multi-purpose program
that has the following syntax:  ``ImageMath ImageDimension
outputfilename   Operation  InputFileName  parameters ``.  Most basic
scalar image operations -- and some tensor operations -- may be
performed with this program.  Some operations output text files, some
output images and some output only to the terminal.   ImageMath allows
one to multiply images together (m), to negate images (Neg), to take
an image to a power (pow), to test the invertibility of
transformations (InvId), to compute the fractional anisotropy of an
image (TensorFA) and to perform many other feats of image processing strength.   
Like all other ANTS programs, one may call ImageMath from the command
line to see all of its options.   ImageMath is used heavily in ANTS
scripts.   You can call it, in compination with grep or less, to attempt to
find some of its capabilities.

\subsection{ANTS Scripts} 
The ANTS/Scripts directory contains (hopefully) user-friendly
wrappings of ANTS tools that enable higher-level error checking and
combinations of basic ANTS functions.  These scripts are called as
\verb bash ~ \verb antsscriptname.sh ~ and provide usage when called
from the command line.  For instance, try \verb bash ~ \verb ants.sh .
Additionally, if you run ants in tcsh shells, you will need to call it
as follows:
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m CC\[r16slice.nii,r64slice.nii,1,2\]
\end{verbatim}
instead of
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m CC[r16slice.nii,r64slice.nii,1,2]
\end{verbatim}
because tcsh does not interpret the brackets as intended.  Using the
bash shell script interface also avoids this.  The key change involves using the slash to force tcsh to interpet the bracket literally. 
\newpage
\section{ImageRegistration with ANTS}
\begin{wrapfigure}{r}{0.5\textwidth}
\framebox[0.5\textwidth]{
\begin{minipage}[t]{0.45\textwidth}
{\bf Quick Start:} {\color{blue}call ants.sh (from ANTS/Scripts/) to get usage and apply a normalization to some of your existing data.  It is instructional to read the script, modify some of the parameters and re-run to witness the effect of your changes -- image registration is an art as well as science.  Many other ``ready to go'' scripts are available in ANTS/Scripts. The user must set his/her ANTSPATH environment variable  -- which points to the location of ANTS binaries -- within these scripts or in their environment.  Note: All ANTS programs provide usage when called from the command line.  Most require the image dimension to be specified as the first parameter.  E.g. 
  ImageMath ImageDimension 
where ImageDimenion is 2, 3 or 4.}
\end{minipage}}
\end{wrapfigure}
There are two general applications of image registration. The first application is transforming labeled data from a template image space into an individual space. This strategy is important when appearance alone is not enough to locate a structure as, for example, in the case of hippocampus segmentation. The template provides a prediction of the hippocampus-amygdala boundary.
The second application operates in the "inverse" direction of the first: instead of mapping template to individual, we map individual(s) to the template. Voxel-based population studies of either functional or structural variables depend on mapping to a template space. The common coordinate system enables a statistical evaluation of the likelihood of consistent activation across a group or, in other contexts, the differences in anatomy between two groups.

The ANTS toolkit enables both types of mapping. The main challenge in image and brain mapping is defining the way in which images/anatomy are compared. There are two components to the comparison.
The shape transformation space defines the range of shape variation that will be allowed in the optimization.
The appearance similarity space defines the statistical assumptions that determine when one image is considered to appear similar to another.

These two components interact in a weighted optimization within a multiple resolution gradient descent framework. Each component may use either "strict" or "flexible" assumptions about the shape or appearance similarity. The selection of these models should be done in a principled way that fits the problem at hand. No single choice is appropriate for all scenarios (see the "no free lunch" theorem).

Thus, ANTS enables many operating points from both the transformation and appearance domains such that users may make choices appropriate for their problems. The ANTS command-line syntax, shown in figure~\ref{fig:cmd}, reflects these operating points and the various components that interact in the optimization.
\begin{figure}
\includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax.pdf} 
\itkcaption[ANTS Command Line]{The relationship between the variational 
form that defines the optimization goals and the ANTS command line.  The 
SyN option implements the method in \cite{Avants2008b} and evaluated in 
\cite{Klein2009}.}
\vspace{-0.1in}
\label{fig:cmd}
\end{figure}
ANTS may be used to navigate the transformation and similarity metric space. 
\subsection{Use Your Header!} ANTS uses the direction/orientation, spacing
and origin in its definitions of the mapping between images.  That is,
ANTS uses the nifti standard defintions of an image space.  So, the
image orientation (direction), origin and spacing are important!  The
program 'PrintHeader' will give you information about your image's header.

\subsection{ANTS Transformation Models}
The ANTS toolkit provides a hierarchy of transformations with
adjustable levels of complexity, regularization, degrees of freedom
and behavior as optimizers.  The simplest transformation model is the
rigid and/or affine transform.  The most complex -- and most flexible
-- is a symmetric diffeomorphic transformation based on optimizing and
integrating a time-varying velocity field.  Computation time also
increases with transformation model complexity, but most normalization
needs may be met with under an hour of computation.  
We provide an overview of some of the ANTS models below and 
try to communicate intuition on what one gains/loses with each choice.  
An overview of available models and similarity terms is in Table~\ref{table:chart}.
\begin{table*}
  \centering
    \begin{tabular}{c c c c}
    {\bf Category} & {\bf Transformation, $\phi$} & {\bf Similarity Measures} & {\bf Brief Description} \\
    \toprule     
    \multirow{2}*{\bf Linear}
           & Rigid$^\dagger$ & MI, MSQ & Rigid registration. \\
       {} & Affine$^\dagger$ & MI, MSQ & Affine registration. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Elastic}
           & Deformable & CC, PR, MI, MSQ, PSE & Demons-like algorithm. \\
       {} & DMFFD & CC, PR, MI, MSQ, PSE  & FFD variant. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Diffeo.}
           & Exponential$^\dagger$ &  CC,PR, MI, MSQ, PSE  & $~\min~\velocity(\x)$ \\
       {} & Greedy SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  & ~~locally in time~$\min~\velocity(\x,t)$\\
       {} & Geodesic SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  &  $~\min~\velocity(\x,t)$ over all time \\
    \bottomrule
    \end{tabular}
  \caption{Transformations and a subset of the similarity metrics
    available in ANTS.  Similarity metric acronyms:  CC = fast cross
    correlation (the preferred metric), PR = cross correlation (deprecated), MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation \cite{Pluta2008}.  ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.  The brief descriptions of the diffeomorphic algorithms contrast the way in which the velocity field is optimized and used to parameterize $\phi$, the mapping.  All 
ANTS Diff algorithms generate $\phi(\x,t)$ over $t \in [ 0, 1]$ through gradient descent.}
  \label{table:chart}
\end{table*}    
\vspace{0in}

\subsubsection{Affine and Rigid Registration}
\begin{figure}
\includegraphics[width=0.9\textwidth]{./Figures/ANTSanatomy.pdf} 
\itkcaption[ANTS Introduction]{The anatomy of an ANTS optimization and application of the resulting warping.}
\vspace{-0.1in}
\label{fig:aff}
\end{figure}
The most basic mapping between images is an affine mapping. The ANTS 
affine mapping syntax is shown in the affine mapping figure~\ref{fig:aff}. The
parameter breakdown of the ANTS call in the affine figure is:
\begin{itemize}
\item  ANTS -- the normalization program
\item 2 -- the expected image dimension
\item -m MI[...,...,1,32] -- the similarity metric "mutual information" with a 32-bin square joint histogram and weight of 1.  
Metric details will be discussed in a later section. 
\item -i 0 -- no deformable iterations (affine only).
\item -o ab -- the output prefix.  A specific extension may also be specified, e.g.   -o ab.mhd  .
\end{itemize}
The parameter breakdown of the WarpImageMultiTransform call in the affine figure is:
\begin{itemize}
\item WarpImageMultiTransform -- to apply the mapping output from ANTS to an image.
\item 2 -- the expected image dimension.
\item the image to be deformed is input next.
\item the deformed image output file name is next.
\item the transform itself is next -- if is affine and preceded by " -i ", we apply the inverse affine map.
\item the " -R " option dictates the domain you want to warp into -- usually the ``fixed'' image, unless using the inverse mapping, in which case one switches the role of fixed and moving images. 
\end{itemize}
Note that -i 0 (in the ANTS call) means that no deformable mapping is performed. The call "ANTS 2" means to treat the images as 2-dimensional. "ANTS 3" would be called for 3D images. The " -m " term defines the similarity metric. 
Note that no more than one image interpolation is used during optimization.  
That is, ANTS always references the original image before applying a transformation or series of transformations.

Affine mapping may become nontrivial when the region of interest 
occupies only a small portion of the image data.  In such cases, 
ANTS enables the user to define a mask to focus the optimization 
effort on the region of interest.  
Example code for (affine) registration with a mask: 
\begin{verbatim}
ANTS 3 -m MI[AtlasHead.nii,crAnatomical.nii,1,32] 
-o TEST -i 10x10x0  -r Gauss[1.5,0] -t Exp[0.5] -x mask.nii.gz
\end{verbatim}
The difference between this command 
and a regular ANTS call is the \verb"-x" option, which specifies the mask, defined in the 
template space (here, AtlasHead.nii).   The mask option also affects the deformable optimization. 
Other affine registration options include  \verb"--number-of-affine-iterations 5000x5000x5000" which specifies that the affine registration uses a 3 level image pyramid with each level 5000 iterations at most. \verb"--MI-option 48x6000" means to use Mutual Information as similarity metric with 48 bins and 6000 samples. \verb"--affine-gradient-descent-option" defines the options for the gradient descent procedure; the 4 numbers are maximum step length, relaxation factor, minimum step length and translation scale.

\noindent{\bf Rigid Registration: }  ANTS will also perform rigid registration.  
To force the linear component of registration to be purely rigid (no scaling or shearing), 
then pass the parameter  \verb"--rigid-affine true " to 
the ANTS command line.   For example,  
\begin{verbatim}
ANTS 3 -m MI[AtlasHead.nii,crAnatomical.nii,1,32] 
-o TEST -i 0 --rigid-affine true 
\end{verbatim}
The jacobian of the resulting affine mapping should be unity.

\subsubsection{Deformable Registration}
Affine mapping is adequate when the
difference between images involves only rotation, scaling or shearing.
Other data requires more deformable mappings to capture shape
differences and find a good alignment of image features.
Figure~\ref{fig:antshier} shows how deformable mapping may improve the
correspondence of the deformed beetle to the ford image.
\begin{figure}
\includegraphics[width=0.9\textwidth]{./Figures/ANTSElastToDiff.pdf} 
\itkcaption[ANTS Transformation Examples]{This example shows the degree to which the beetle (b.img) may be deformed to the ford (a.img) under different transformation models.  Left to right increases the degrees of freedom in the mapping and thus the registration accuracy.}
\vspace{-0.1in}
\label{fig:antshier}
\end{figure}
Most ANTS-based applications use symmetric diffeomorphic normalization.  However, 
ANTS also enables a simpler parameterization of a deformable mapping via a regularized 
vector space.  We term these types of transformations as ``elastic''.  The original Demons algorithm 
provides a classic example of using a regularized vector space for nonlinear registration.  
Caveats are that a regularized vector space may not preserve the underlying topology 
and may also prove too inflexible to capture the shape changes one is after.  Both of these 
shortcomings motivate the use of diffeomorphisms.

\noindent{\bf Elastic/Vector Space Transformations.} If we assume no affine transformation, then 
an elastic transformation involves computing a mapping from image $I(\x)$ to image $J(\x)$ 
through a deformation field $\disp( \x )$.  
\begin{figure}
\center \includegraphics[width=0.75\textwidth]{./Figures/PullBack.pdf} 
\itkcaption[ANTS Deformation]{A deformation in a digital domain.}
\label{fig:pull}
\end{figure}  
The deformation is defined in the physical space of the image 
and dictates the positional difference between corresponding features in the two images.  
Thus, if a feature defined at $I(\x)$ matches a feature in $J$ at position $\y$ then the deformation 
field at $\x$ should give $\disp ( \x )=\y-\x$.   
Such a deformation field may be applied to deform 
image $J$ into image $I$ by composing the mapping $J_{deformed}(\x)=J(\x+\disp(\x))$.  
In a perfect world, then $I(\x)=J_{deformed}(\x)$, though this is rarely the case.  Figure~\ref{fig:pull} 
visualizes the deformation of an image under this standard model. 
Gradient descent optimization of an elastic mapping may be summarized (crudely) as: 
\begin{align}
\text{Compute the similarity gradient:   }  \nabla E = \partial_{\disp} \Pi(I,J(\x+\disp(\x))). \nonumber \\
\text{Update the deformation field:  }  \disp(\x) \leftarrow \disp(\x)+\delta \nabla E. \nonumber \\ 
\text{Regularize the deformation field:  }  \disp(\x) \leftarrow G_\sigma \star \disp(\x),
\end{align}
where $\Pi$ is the similarity, $\delta$ is a gradient step length and $G_\sigma$ is a gaussian smoother.
This optimization is captured in the following ANTS command:
\begin{verbatim}
ANTS 3 -m CC[AtlasHead.nii,crAnatomical.nii,1,4] -r Gauss[0,3]  
-o ElasticTest.nii.gz -i 30x20x10  -t Elast[1.5] 
\end{verbatim}
Here, we use the CC metric (a cross-correlation implementation) with
window radius 4, weight 1 and gradient step length 1.5.  The
optimization will be performed over three resolutions with a maximum
of 30 iterations at the coarsest level, 20 at the next coarsest and 10
at the full resolution.  We use a Gaussian regularizer with a sigma of
3 that operates only on the deformation field and not on the similarity gradient, as 
0 is passed as the first parameter to \verb Gauss .  One may see the
correspondence, yet again, between the ANTS call and the optimization
scheme.  The optimization will stop when either the energy cannot be 
smoothly minimized or the maximum number of iterations is reached. 
BSpline regularization is also available in ANTS -- see the DMFFD 
section below.

\noindent{\bf Warping and Invertibility.}  ANTS uses physical space to
define mappings. The origin etc is in the coordinates of the world in
which the picture (e.g. MRI) was taken. Then, warping between physical
coordinates is relatively easy. Differences in bounding boxes etc
present no problem -- assuming you avoid inconsistent headers
i.e. origins, directions, data orientation. One may use PrintHeader to
check the data and run simple, fast tests (few iterations) to perform
sanity checks before running through loads of data.
An ANTS deformation consists of
a standard naming prefix plus a standard naming extension. We usually 
assume nii but other file types may be used. The standard naming
extension is Warpxvec.nii.gz for the x component of the deformation with
similar naming for y and z. Each component is stored in a separate
scalar image. The value of a voxel of a deformation/warp component
image is the physical space displacement from that voxel. Note that
the inverse mapping is stored in InverseWarpxvec.nii.gz and provides the
mapping in the opposite direction.  Note that an inverse -- in the 
digital domain -- is only approximate as shown in figure~\ref{fig:inv}.
\begin{figure}
\center \includegraphics[width=0.75\textwidth]{Figures/DeformationsInDigitalDomain.pdf} 
\itkcaption[Digital Invertibility]{Digital invertibility presents some limitations. 
Here, we see that invertibility is not exact but is gained only by interpolation.  
Thus, in three-dimensional scenarios in particular, there are fundamental limits to the 
degree of invertibility that may be achieved.  The second and third voxels -- from left -- 
in image A undergo an expansion by a factor of 2.  That is, under the mapping, 
2 voxels are mapped to 4.  This gives the definition of the Jacobian -- computed 
by  ANTSJacobian -- which is a unitless measure defined by the 
ratio of volumes.  Thus, $\mathcal{J}(\x)=V(\phi(\x))/V(\x)$ where $V$ represents 
the volume operation and $\x$, here, may be a small object.  Thus, if $\phi$ -- the mapping -- 
causes expansion, then $\mathcal{J}(\x) > 1$. }
\vspace{-0.1in}
\label{fig:inv}
\end{figure}  
A few important notes follow. 
(1) Deformation directionality: Warps/deformations applied to images occur in the opposite direction from warps/deformations applied to points. That is, if we map image B to A using 
\begin{verbatim}
ANTS 2 -m CC[A,B,1,2] -o OUTPUT 
\end{verbatim}
 then we get a deformation called \verb OUTPUTWarp+extensions ~and an affine transform called \verb OUTPUTAffine.txt . 
This composite mapping - when applied to B - will transform B into the space of A. 
However, if we have points defined in B that we want to map to A, 
we have to use \verb OUTPUTInverseWarp ~and the inverse of \verb OUTPUTAffine.txt .  
This is because the domain and range of the map's definition need to be dense in a way that is appropriate for the data type. 
Figure~\ref{fig:inv} illustrates the concept. 
(2) Older Image Formats: older image formats (e.g. Analyze) may not
have proper origin/offset/direction support! In these cases, we
recommend converting to nii and verifying that data overlays properly.
(3) Transorm Composition: Composition of transforms may be achieved with
ComposeMultiTransform.  (4) Warping with inverses and concatenations --
viable when using diffeomorphisms -- are described in
figure~\ref{fig:warp}.
 \begin{figure}
\includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax2.pdf} 
\itkcaption[ANTS Warping]{Here, we detail how one applies 
a deformation and associated inverses.  We also see how 
WarpImageMultiTransform can be used to concatenate a series 
of transformations.}
\vspace{-0.1in}
\label{fig:warp}
\end{figure}

\noindent{\bf Diffeomorphic Transformations.}  The elastic 
mapping space -- as indicated above -- may prove inadequate 
for some large deformation mapping scenarios.  Figures~\ref{fig:antshier} 
illustrate the changing performance one may get in switching from affine 
to elastic to diffeomorphic normalization.  
\begin{figure}
\includegraphics[width=0.9\textwidth]{./Figures/ANTSsyn.pdf} 
\itkcaption[ANTS Symmetric Normalization]{This example shows the benefit of the symmetric normalization model -- invertibility, symmetry, highly deformable and accurate registration.  The transformation model used here is highlighted in red.}
\vspace{-0.1in}
\label{fig:antssyn}
\end{figure}
Figure~\ref{fig:antssyn} shows 
how one may use ANTS to achieve a state-of-the-art diffeomorphic mapping. 
The ANTS diffeomorphic model chosen for this example -- symmetric normalization \cite{Avants2008b,Kim2008} -- 
is invariant to the order of the input images (although the affine mapping is not).  
An additional advantage of the diffeomorphic model over the elastic model is 
that both forward and inverse transformations are computed thus allowing one 
to deform fixed to moving and also moving to fixed.   The transformation model 
chosen here -- \verb SyN[0.25] -- may be replaced with other 
diffeomorphic transformation models.  The step-size (0.25) impacts
accuracy.  Smaller is generally more accurate but takes more time to
compute and may not capture as much deformation if the optimization is
caught in a local minimum.  The most general is global-in-time SyN via \verb SyN[0.25,2,0.01]   , 
where the time step for integration is 0.01 (lower is more accurate).  A fast approximate 
solution may be gained through 
exponential mapping via \verb Exp[0.25,10] , where 10 integrations points are used. 
Update schemes for diffeomorphic optimization are similar to the elastic case 
and are described elsewhere \cite{Avants2008b,ANTStheory}.

\subsection{ANTS Similarity Terms and Other Functions}
Here is a script that will let you experiment with different similarity term 
and transformation model combinations.  A few are listed, with parameters 
that are a reasonable starting point.
\begin{verbatim}
# The user should define two images.  here we use ANTS reference data
II=r16slice.nii.gz 
JJ=r64slice.nii.gz
# 
OUTPUTNAME=OUT  # define the output prefix 
ITS=" -i 100x100x30 " # 3 optimization levels

# different transformation models you can choose
TSYNWITHTIME=" -t SyN[0.25,5,0.01] -r Gauss[3,0] " # spatiotemporal (full) diffeomorphism
TGREEDYSYN=" -t SyN[0.15] -r Gauss[3,0] "          # fast symmetric normalization
TELAST=" -t Elast[1] -r Gauss[0.5,3] "             # elastic 
TEXP=" -t Exp[0.5,10] -r Gauss[0.5,3] "            # exponential

# different metric choices for the user 
INTMSQ=" -m MSQ[${II},${JJ},1,0] "
INTMI=" -m MI[${II},${JJ},1,32] " 
INTCC=" -m CC[${II},${JJ},1,4] " 
#
#
# these are the forward and backward warps. 
#
INVW=" -i ${OUTPUTNAME}Affine.txt ${OUTPUTNAME}InverseWarp.nii.gz " 
FWDW=" ${OUTPUTNAME}Warp.nii.gz ${OUTPUTNAME}Affine.txt " 
#
INT=$INTCC  # choose a  metric ( here , cross-correlation )
TRAN=$TGREEDYSYN  # choose a transformation 
if [[ $OPT == "Reg" ]] ; then

# run the registration
${ANTSPATH}/ANTS $DIM -o $OUTPUTNAME $ITS $TRAN $INT 
# this is how you apply the output transformation 
${ANTSPATH}/WarpImageMultiTransform $DIM ${II} ${OUTPUTNAME}IItoJJ.nii.gz -R ${JJ} $INVW 
${ANTSPATH}/WarpImageMultiTransform $DIM ${JJ} ${OUTPUTNAME}JJtoII.nii.gz -R ${II} $FWDW

elif [[ $OPT == "RegSeg" ]] ; then

# try something different -- segment the images and use them in ANTS 
# to run multivariate registration.
IISEG=${OUTPUTNAME}A_seg.nii.gz
JJSEG=${OUTPUTNAME}B_seg.nii.gz
#
# get masks for atropos 
#
${ANTSPATH}ThresholdImage $DIM $II $IISEG 1 1.e9
${ANTSPATH}ThresholdImage $DIM $JJ $JJSEG 1 1.e9
AtroposParams=" -d $DIM -m [0.1,1x1] -c [5,0] -i kmeans[3] "
${ANTSPATH}/Atropos $AtroposParams -a $II -o $IISEG  -x $IISEG 
${ANTSPATH}/Atropos $AtroposParams -a $JJ -o $JJSEG  -x $JJSEG 
# compute some segmentations and use them in labelguided mapping 
LABELGUIDED=" -m PSE[${II},${JJ},${IISEG},${JJSEG},0.75,0.1,25,0,10] "  
#
${ANTSPATH}/ANTS $DIM -o ${OUTPUTNAME} $ITS $TRAN $INT $LABELGUIDED 
${ANTSPATH}/WarpImageMultiTransform $DIM ${II} ${OUTPUTNAME}IItoJJ.nii.gz -R ${JJ} $INVW 
${ANTSPATH}/WarpImageMultiTransform $DIM ${JJ} ${OUTPUTNAME}JJtoII.nii.gz -R ${II} $FWDW

elif [[ $OPT == "BTP" ]] ; then 
#
# now build a template from the two input images .
# in a real application, you would modify the call to btp 
# such that it's more appropriate for your data.  
# call buildtemplateparallel.sh -h to get usage.
# 
${ANTSPATH}buildtemplateparallel.sh -d 2  -o ${OUTPUTNAME}BTP -c 0 $II $JJ
fi 
\end{verbatim}
Convergence occurs under two conditions: (1) either the maximum number of 
iterations are reached at a given optimization level or (2) the slope 
of the change in the optimization objective is negative or very small.

\subsection{Choosing a Metric}
ANTS supports both volumetric registration and point set registration. The image / point set similarity metrics in ANTS are unified in the form of a function on the images or the point sets:       $$\textbf{Similarity}[\textbf{fixedImage}, \textbf{movingImage}, \textbf{weight}, \textbf{parameters}].$$

The similarity type for the deformation transformation is specified by
the \verb -m  option, which contains two parts: simarity type and the parameters inside the brackets. (Note: no white spaces exist between parameters.) The possible similarity metrics for volumetric images are: 
\begin{itemize}
    \item Cross correlation estimate:  \verb " -m  CC[fixedImage,movingImage,weight,radius]". This metric works
      well for intra-modality and some inter-modality image registration. For example, \verb"-m CC[fixed.nii,moving.nii,1,5]" specifies:
    \begin{itemize}
        \item the fixed image: fixed.nii
        \item the moving image: moving.nii             
        \item weight for this metric is 1 (i.e. only this metric drives the registration).      
        \item the region radius for computing cross correlation is 5
    \end{itemize}

    \item Mutual information: \verb"-m MI[fixedImage,movingImage,weight,number-of-histogram-bins]". This metric works both well for intra-modality and inter-modality image registration. For example, the first three parameters in \verb"-m MI[fixed.nii,moving.nii,1,32]" is similar to the example above in cross correlation, except that the last parameter means that the number of bins in computing mutual information is 32.

    \item PR: \verb"-m PR[fixedImage,movingImage,weight,radius]". This
      metric is deprecated.  It still ``works'' but CC is preferred.

    \item Mean square difference: \verb"-m MSQ[fixedImage,movingImage,weight,0]". This metric works for intra-modality image registration. The last parameter \verb"0" is a padding value of no real meaning. For example, \verb"-m MSQ[fixed.nii,moving.nii,1,0]".

\end{itemize}

ANTS also support registration of point sets. The supported formats for point sets can be found in I/O section. The similarity metrics for point sets are:
\begin{itemize}
    \item Point set expectation: 
     \begin{verbatim}
-m PSE [fixedImage,movingImage,fixedPoints,movingPoints,weight,
pointSetPercentage,pointSetSigma,boundaryPointsOnly,
kNeighborhood,PartialMatchingIterations=100000]
    \end{verbatim}
    \begin{itemize}
        \item \verb"fixedImage": defines the space domain of the fixed point set.
        \item \verb"movingImage": defines the space domain of the moving point set.
        \item \verb"fixedPoints/Image": defines the coordinates of the fixed point set or label image. It can be an image with discrete positive labels, a VTK format point set file, or a text file. Details can be found in I/O section (TODO).
        \item \verb"movingPoints/Image": defines the coordinates of the moving point set or label image. 
        \item \verb"weight": weight for this metric. \verb"1" means that only this metric drives the registration.
        \item \verb"pointSetPercentage": the percentage of points to be randomly sampled used in the registration.
        \item \verb"pointSetSigma": the standard deviation of the Parzen window used to estimate the expectation. 
        \item \verb"boundaryPointsOnly":  \verb"1" (or ``true'') means only the boundary points in the label image is used to drive registration.
        \item \verb"kNeighborhood" is a positive discrete number. The first $k$ neighbors are used to compute the deformation during the registration. 
        \item \verb"PartialMatchingIterations" controls the symmtry in the matching. This option assumes the complete labeling is in the first set of label parameters ... more iterations leads to more symmetry in the matching  - 0 iterations means full asymmetry 
    \end{itemize}

    \item Jensen-Tsallis BSpline
    \begin{verbatim}
-m JTB[fixedImage,movingImage,fixedPoints,movingPoints,weight,
pointSetPercentage,pointSetSigma,boundaryPointsOnly,kNeighborhood,alpha,
meshResolution,splineOrder,numberOfLevels,useAnisotropicCovariances] 
    \end{verbatim}
    \begin{itemize}
        \item \verb"fixedImage": defines the space domain of the fixed point set.
        \item \verb"movingImage": defines the space domain of the moving point set.
        \item \verb"fixedPoints": defines the coordinates of the fixed point set. It can be an image with discrete positive labels, a VTK format point set file, or a text file. Details can be found in I/O section (TODO).
        \item \verb"movingPoints": defines the coordinates of the moving point set.
        \item \verb"weight": weight for this metric. \verb"1" means that only this metric drives the registration.
        \item \verb"pointSetPercentage": the percentage of points to be randomly sampled used in the registration.
        \item \verb"pointSetSigma": the sigma for the Parzen window used to estimate probabilities.  
        \item \verb"boundaryPointsOnly": [TODO] \verb"1" (or ``true'') means only the boundary points in the point sets are used to drive registration.
        \item \verb"kNeighborhood" is a positive discrete number. The first $k$ neighbors are used to compute the deformation during the registration. 
        \item \verb"alpha"
        \item \verb"meshResolution"
        \item \verb"splineOrder"
        \item \verb"numberOfLevels"
        \item \verb"useAnisotropicCovariances"
    \end{itemize}

\end{itemize}

ANTS has three affine registration similarity metrics for volumetric images, which are specified using \verb"--affine-metric-type":
\begin{itemize}
    \item Mutual information, specified as \verb"--affine-metric-type MI". This usually works for both inter and intra-modalities in 3D. Also, the options used in computing mututal information in affine registration can be specified as \verb"--MI-option N1xN2". The first parameter \verb"N1" specifies the number of bins. The second parameter \verb"N2" specifies the number of samples. For example: \verb"--MI-option 32x8000".
    \item Cross-correlation  \verb"--affine-metric-type CC".  CC has a
      wide capture range.  
    \item Mean square difference, specified using \verb"--affine-metric-type MSE". There is no options for this metric.
\end{itemize}

Fig. \ref{fig:metric_example} shows the registration result using intensity difference with the following command:
\begin{verbatim}
ANTS 2 -m MSQ[r16slice.nii,r64slice.nii,1,0] -r Gauss[3,0] -t SyN[0.5] -i 50x50x30
\end{verbatim}

%        \begin{wrapfigure}{r}{0.5\textwidth}
\begin{figure}
    \label{fig:metric_example}
    \centering
    \begin{tabular}[h]{c|c|c}
%        \includegraphics[width=0.16\textwidth]{Figures/r16slice.jpg} &
%        \includegraphics[width=0.16\textwidth]{Figures/r64slice.jpg} &
 %       \includegraphics[width=0.16\textwidth]{Figures/resMSQ.jpg} \\
        fixed image &
        moving image & 
        MSE  \\
    \end{tabular} 
    \itkcaption{registration using mean square intensity difference}
\end{figure}
    %    \end{wrapfigure}
Here is also an example script to register a pair of images using mean square intensity difference and computing the metrics of the registration image.
\begin{verbatim}
#use intensity difference with radius 0 -- radius no effect on intensity difference
ANTS 2 -m MSQ[r16slice.nii,r64slice.nii,1,0] -r Gauss[3,0] -t SyN[0.5] -i 50x50x30
WarpImageMultiTransform 2 r64slice.nii resMSQ.nii Warp.nii Affine.txt
MeasureImageSimilarity 2 0 r16slice.nii resMSQ.nii metricexamplelog.txt
MeasureImageSimilarity 2 1 r16slice.nii resMSQ.nii metricexamplelog.txt
MeasureImageSimilarity 2 2 r16slice.nii resMSQ.nii metricexamplelog.txt
ConvertToJpg resMSQ.nii resMSQ.jpg
\end{verbatim}

\subsection{Diffusion Tensor Normalization}
The ImageMath program -- via TensorFA and TensorMeanDiffusion -- 
may derive scalar images from DTI.  Such images may be used to 
distortion correct DTI to T1 or to map DTI together.   See the 
ANTS/Scripts program called \verb antswithdt.sh ~for a 
ready to go example.   Figure~\ref{fig:dterr} shows what might 
happen if your tensor entries are stored in the wrong order. 
ANTS expects the nifti standard ordering of the DT six vector.
\begin{wrapfigure}{r}{0.5\textwidth}
\includegraphics[width=0.5\textwidth]{Figures/dtiordererr.pdf} 
\itkcaption[DT Entry Error]{The FA and mean diffusion of the tensor may be corrupted, 
as at right, if the order of the DT entries is wrong.}
\vspace{-0.1in}
\label{fig:dterr}
\end{wrapfigure}  
\subsection{Multivariate Normalization with ANTS}
\begin{figure}
\includegraphics[width=0.9\textwidth]{Figures/multivariatenorm.pdf} 
\itkcaption[Multivariate Normalization]{The current version of ANTS 
may perform normalization of multiple modalities by combining {\em intrasubject, intermodality} mappings with {\em intersubject, intramodality} maps to a 
group template.  In brain imaging, the intersubject maps are usually guided by the T1 component.}\vspace{-0.1in}
\label{fig:mvnorm}
\end{figure}
Multivariate normalization may be performed in two ways with ANTS.  
First, as shown in figure~\ref{fig:mvnorm}, intra-subject mappings may 
be used to conform all modalities into a common subject space defined by 
the ``anchor'' modality.  In brain imaging, this is usually the T1 component.

A second type of multivariate normalization is able to use the T1 and DTI components 
directly in the optimization of the mapping.  An example of this type of call is below, 
where we assume that one has resampled the DT data to the space of T1: 
\begin{verbatim}
ANTS 3 -m CC[T1template.nii,T1subject.nii,1.25,4]
       -m CC[FAtemplate.nii,FAsubject.nii,1,4] 
-o MultiVar -i 30x30x20  -r Gauss[3,0] -t SyN[0.25]
\end{verbatim}
In this case, dense features within white matter are gained by using the DT component  
in addition to the T1 component.   The T1 CC metric, here, is weighted slightly more 
than the FA CC metric, 1.25 vs. 1.0.  {\em The convergence criterion for 
the multivariate scenario is a slave to the {\bf last} metric you 
pass on the ANTS command line}.

\subsection{Optimal Template Construction with ANTS}
A useful script for setting up optimal template construction
 is available in ANTS/Scripts.  This method is described in \cite{Avants2004,Avants2006d,Kim2008,Yushkevich2009,Avants2009c}, 
where more recent publications are more descriptive and incorporate 
more recent features and optimization strategies. 
Two versions of this algorithm are available : serial and parallel.  
Data and a script for building an average face template is available 
the ANTS/Examples/Data directory.  Two key points about optimal templates:
1. The outcome stabilizes at around ten images, for most populations.
That is, if we randomly select images from individuals within the same 
demographic distribution, we will end up with very similar average images.  
2.  Optimality, for the ANTS-SyN approach, is defined by the minimum shape 
and appearance distance image. 
See the template page for some examples of this and users may 
contact ANTS developers for previously derived template 
images that may be useful.

\subsubsection{A Concrete Template Construction Example, Step-by-Step}
For this example, we will use the data in "ANTS/Examples/Data/B*.tiff."
The files are B1.tiff, B2.tiff, B3.tiff, B4.tiff and B5.tiff.   They are single slices 
from a dataset of real MRI brain images -- individuals around 15 years of age. 
Our goal is to derive a "most representative" single image from this population. 
The steps are: 
\begin{figure}
\center \includegraphics[width=0.7\textwidth]{Figures/templateex.jpg} 
\itkcaption[ANTS Optimal Template]{This example may be recreated by the user.  
The shared anatomy -- 
across this dataset -- is recovered in the derived optimal template.}
\label{fig:template}
\end{figure}
\begin{enumerate}
\item   Make a directory for this example.   Copy or link all the B*tiff images into it.
\item Copy the file ANTS/Scripts/buildtemplateparallel.sh into that directory.  
\item  On the command line, within that directory, run the following command: 
\verb  bash ~ \verb buildtemplateparallel.sh ~ \verb -h ~ to get usage. 
\item Following the usage, you can write a script that contains a call like this: 
\verb buildtemplateparallel.sh ~ \verb  -d ~ \verb  2 ~ \verb -o ~ \verb  BTP ~ \verb -c ~ \verb  0 ~ \verb  B*tiff
% \verb bash ~ \verb ~ \verb 2 ~ \verb B ~ \verb tiff ~ \verb OUT ~ \verb 4 ~ . 
\item  Call the script in step 4 and then wait for the output.  The template will 
be called (in this example)  \verb BTPtemplate.nii.gz .
\item If the script does not produce the desired output, check that
  you defined \$ANTSPATH in your path.
You may also check and or define the path to the ANTs programs
explicitly by editing this variable in buildtemplateparallel and antsIntroduction.
\end{enumerate}
The script offers an option ( -c 1  for qsub , -c 2 to attempt to use
multiple cores on your machine )  so that you can send the whole thing to distributed computing. 
Typically, one would use voxbo or the qsub program available in the Sun Grid Engine.  
Figure~\ref{fig:template} shows the data and a result that may be recreated by the user.
\subsection{Notes on Large Deformation Mapping} 
Figure~\ref{fig:large} defines what one might expect from a 
high-resolution, large deformation, successful normalization.  
Major and many minor features are well aligned between 
these images to an extent that is approaching the limits 
allowable by biological variation.  
\begin{figure}
\includegraphics[width=0.9\textwidth]{Figures/ANTSLargeDef.jpg} 
\itkcaption[ANTS LargeDeformation]{ANTS succeeds in a challenging 
normalization scenario.  Comparing the template mapping to the
individual shows that much of the cortex is well-aligned, as is the
hippocampus, despite the relatively large difference between the
initial template and the target.  Here, one might note a limitation of
whole brain mapping: occipital lobe sulcal variation is highly
idiosyncratic and extremely difficult to capture, in particular when
there is such severe atrophy nearby.
}
\label{fig:large}
\end{figure}
\noindent{\bf Turning "Failures" into Successes:} 
Below are some pointers to follow if you are unable to recreate such normalization quality.  
Usually, reasons for registration failure are one of a very few things: 
\begin{enumerate}
\item The initialization is so off mark that affine registration fails -- thus meaning all subsequent deformable registration will not be meaningful.
\item The information within headers is inconsistent e.g. origins are defined in different ways, the "directions" are not correct or some combination of these.    The PrintHeader executable can help one in debugging this type of problem.  Also, the ImageMath function CompareHeadersAndImages can fix some problems of this type. 
\item  The similarity or transformation model is inappropriate or has too small a capture region for the problem. 
\end{enumerate}
An example of this final point is given here.   
A recent study used the following call to ANTS across a dataset of elderly subjects:
\begin{verbatim}
ANTS 3 -m CC[template.nii.gz,subject.nii.gz,1,2] -i 10x50x50x20
 -o subjectmap.nii.gz -t SyN[0.25] -r Gauss[3,0]
\end{verbatim}
This succeeded for all subjects but one.  This subject had severe neurodegeneration which caused "under-normalization" to result.  A larger deformation mapping was required and so we increased the maximum number of iterations (the "-i" parameter).  One might also increase the gradient-descent step-size (SyN[0.25] $\rightarrow$ SyN[0.5]).   Large deformation mapping is challenging not only because of the amount of deformation, but also because the "true" solution is difficult to find.  Keeping this in mind, we also increased the "span" (radius) of the correlation window in the similarity metric (which also increases computation time). These modified parameters succeeded: 
\begin{verbatim}
ANTS 3 -m CC[template.nii.gz,subject.nii.gz,1,4] -i 100x100x100x20
 -o subjectmap.nii.gz -t SyN[0.25] -r Gauss[3,0] 
\end{verbatim}
Key changes were to increase the radius of the correlation and to
allow more iterations at the coarsest resolution.  
This two coarsets levels of the computation take only about
10 percent of the time (about 5-10 minutes depending on the machine)
but accounts for the large majority of the shape variation.  Note that
this is a 3D example -- that is why the eyes appear only in the frame
at left.

\newpage
\section{Image Segmentation and Labeling}
ANTS has tools for both tissue based segmentation and prior-based segmentation that 
leverages spatial priors, usually based on a template mapping.  
\subsection{Basic Segmentation}
An advanced MRF segmentation algorithm is available in ANTS via the
\verb Atropos ~tool, which is still under development.  We apply 
the segmentation to the example data in: ANTS/Examples/Data/r16slice.nii.  
\begin{verbatim}
  ThresholdImage 2 r16slice.nii mask.nii.gz 0.001 Inf
  Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
      -c [10,0.001] -m [0.1,1x1x1] -i kmeans[3] -x mask.nii.gz
\end{verbatim}
The parameter \verb kmeans[3] ~indicates 3 tissues plus background are sought.  The \verb [0.05,1] ~indicates 
that some MRF smoothing will be applied to the labels.  Other
options are defined in the Atropos help.  Output is shown 
in figure~\ref{fig:seg} and -- with priors -- in figure~\ref{fig:seg2}.
The document atropos.pdf in the Documentation directory provides much
more explanation.
\begin{figure}
\includegraphics[width=1\textwidth]{Figures/segmentation.pdf} 
\itkcaption[ANTS Segmentation]{ANTS provides basic 
Markov Random Field regularized Gaussian model tissue segmentation. 
The ANTS program N3BiasFieldCorrection is very valuable as preprocessing 
for this naive approach to segmentation.   The last panel -- far right -- 
shows the thickness derived from the white matter and gray matter 
probabilities where a prior on thickness was used to prevent thickness
overestimation.   Thickness was derived with DiReCT \cite{Das2009}, a software 
tool available in binary form only -- a future release may provide code.  
}
\label{fig:seg}
\end{figure}
\subsection{Prior and Template-based Image Segmentation}
The same algorithm may be augmented 
to perform Prior-based segmentation.  
\begin{verbatim}
  Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
      -c [10,0.001] -m [0.05,1] \ 
      -i  PriorProbabilityImages[3,r16slice_prob_%02d.nii.gz,0.5] -x mask.nii.gz
\end{verbatim}
The \verb PriorProbabilityImages ~parameter~ \verb 3 ~indicates 3 tissues
plus background are sought.  The \verb PriorProbabilityImages ~term \verb 0.5 ~indicates 
we weight the priors equally as the data term and use a spatially varying set of Gaussians 
to estimate the segmentation.  In this way, data from the priors may be used to modify 
and guide the segmentation in a locally varying way, accomodating for both inhomogeneity 
and the different imaging signature that different tissues provide.   Prior images should be 
of the same size and dimension as the input data and should have intensities in the range $[0,1]$.
\begin{figure}
\includegraphics[width=1\textwidth]{Figures/segmentation2.pdf}
\itkcaption[ANTS Segmentation]{The ANTs program {\em Atropos} extends basic Markov Random Field regularized Gaussian model tissue segmentation to include priors,
  which allow spatially varying tissue models to determine
  segmentation.  The ANTS program N4BiasFieldCorrection is 
  critical as preprocessing when using this approach to segmentation.
  Here, we see the improvement in segmentation as the locally varying 
  prior models are weighted more heavily.  
  The last panel -- far right -- shows the segmentation derived from
  data driven with an initialization founded purely on prior models 
  based on template mapping.  The prior model brings out the caudate 
and some CSF -- highlighted by arrows -- as their weight increases. 
}
\label{fig:seg2}
\end{figure}
\subsection{Cortical Thickness}
Two forms of cortical thickness estimation from probability maps are available 
in ANTS: first, the traditional Laplacian cortical thickness estimation and, second, 
the more recently developed Diffeomorphic Registration-based Cortical Thickness 
(DiReCT) \cite{Das2009}.  Both methods estimate thickness of an object 
from probability or binary images that define object boundaries.  This tool 
is mainly of interest in brain mapping and cardiac imaging for morphometric 
studies.  Cortical thickness, for instance, is known to correlate with language 
development and IQ in adolescents.  The laplacian approach to 
cortical thickness may be used via the program LaplacianThickness, 
which implements the method described in ``Three-dimensional mapping of cortical thickness using Laplace's Equation'' by 
Jones, et al \cite{}.  Super-sampling 
and controlling segmentation accuracy for input to this program is up 
to the user.  Otherwise, the Laplacian method may grossly overestimate 
thickness in closed sulci.

\subsection{User/Label-Guided Normalization for Hippocampus Mapping}
See the link for details on hippocampus labeling:
\href{http://picsl.upenn.edu/ANTS/hipptutorial.php}{http://picsl.upenn.edu/ANTS/hipptutorial.php}.
\noindent{\bf Expectation-Based Point-Set Registration.}
Here, we apply the expectation-based point set registration method 
for mapping labeled points sets, as described in \cite{Pluta2008}.   ITK-SNAP may be used to label 
images and exported segmentation images may be input to the 
PSE metric below, as labeled data.  The Frown and Smile data is used 
as example.  This data is available in the 
ANTS/Examples/Data/ directory.
\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/frowntosmile.png}
\includegraphics[width=0.45\textwidth]{Figures/smiletofrown.png}
\itkcaption[Frown To Smile]{Expected output for the frown to smile 
shows a smooth, though large deformation.  The grids are overlaid on 
the deformed images. }
\label{fig:frown}
\end{figure}
\begin{verbatim}
ANTS 2 -o PSEtest  -i 91x70x55x40x30  -r Gauss[3,0.] -t SyN[0.2]
   -m  PSE[Frown.nii,Smile.nii,Frown.nii,Smile.nii,0.75,0.1,11,0,10] 
   -m  MSQ[Frown.nii,Smile.nii,1,0]    --number-of-affine-iterations 0

WarpImageMultiTransform  2 Frown.nii FrownToSmile.nii -R Smile.nii
   -i PSEtestAffine.txt PSEtestInverseWarp.nii

WarpImageMultiTransform  2 Smile.nii SmileToFrown.nii -R Frown.nii 
    PSEtestWarp.nii PSEtestAffine.txt

CreateWarpedGridImage 2 PSEtestInverseWarp.nii grid1.nii
CreateWarpedGridImage 2 PSEtestWarp.nii grid2.nii
\end{verbatim}
This example should run on the downloaded ANTS data so you may see the results.
%\input{ANTSLabeledDataExample}
%\input{ANTSMaskedImageRegistrationExample}
\newpage
\section{Application to Studies}
\subsection{Brain Mapping in the Presence of Lesions}
Many cases violate the basic assumption of a one-to-one mapping from a
template image to a target image. Brain lesions caused by stroke or
traumatic brain injury are a common instance of this.

The ANTS toolkit handles this situation by a constrained cost-function
masking approach. In short, the mapping within an excluded region (as
determined by an inclusive mask) is determined by the solution outside
of the region. To achieve this with ANTS, one would use a call similar
to this:
\begin{verbatim}
ANTS 3 -m CC[tp22_s1.nii,template.nii.gz,1,4] -i 50x20x0
   -o tp22map -t SyN[0.25] -x mask.nii.gz -r Gauss[3,0] 
\end{verbatim}
The additional parameter needed for this approach 
is the \verb -x ~ \verb mask.nii.gz ~ which 
is the mask defined in the "fixed" (or individual) image space.  
Mask values above 0.5 are considered non-outlier or non-lesioned regions.
The labeling may be performed with ITK-SNAP. 
Figure~\ref{fig:lesion} illustrates the approach.
\begin{figure}
\includegraphics[width=1\textwidth]{Figures/lesionstudy.pdf}
\itkcaption[ANTS Lesions]{Here, we see a semi-automated approach for lesion analysis
based on diffeomorphic normalization.  The user outlines the lesion
site in the subject space, as shown in (a).  Diffeomorphic mapping is
then used to deform the subject image in (a) into the space of the
reference template.  The deformed subject is shown in panel (b).
Panel (c) shows the reference template space with the estimated
position of the subject's lesion overlaid on the healthy tissue of the
template.  This approach enables one to compare the subject image
against prior information stored in the template space such as
anatomical labels or statistics on the appearance of lesioned and/or
normal tissue.}
\label{fig:lesion}
\end{figure}
Once the ANTS mapping is solved, 
one is able to estimate the location of the lesion in the template space by mapping the inverse lesion mask to the template.
To take the negative image of the inclusive mask 
and warp the mask to the template :
\begin{verbatim}
ImageMath 3 lesionmask.nii.gz Neg inclusivemask.nii.gz
WarpImageMultiTransform 3 lesionmask.nii.gz lesiontotemplate.nii.gz 
   -R template.nii.gz -i tp22mapAffine.txt tp22mapInverseWarp.nii
\end{verbatim}
More details on this approach are being prepared for publication.
See also: \verb http://www.ncrrn.org/papers/methodology_papers/sp_norm_kim.pdf .

\subsection{Statistical Mapping with ANTS: Morphometry, Function, Jacobian, Thickness}
ANTS has been applied in a wide array of imaging studies
\cite{Avants2009a,Avants2009b,Avants2009,Das2009,Klein2009,Massimo2009,Pluta2008,Tustison2009,Yushkevich2009,Avants2008c,Avants2008b,M.Grossman10282008,Kim2008,Simon2008,Sun2008,Yushkevich2008,Aguirre2007,Avants2007,Avants2007i}.
All of these studies benefit in some way from normalization whether the topic is cortical
thickness, Jacobian-based morphometry, volumetric morphometry or functional studies. 
From a broad perspective, each of these applications requires the same steps:
\begin{enumerate}
\item Preprocess images -- bias correction (e.g. via the \verb N4BiasFieldCorrection ~program), 
segmentation (via \verb Atropos ), etc.  Potentially construct an optimal template. 
\item Normalize images -- run \verb ants.sh ~ to map a population of images to a template and store the deformations. 
\item Derive any data from the deformation that may be necessary,
  e.g. the log-Jacobian, via ANTSJacobian.
\item Apply the warp to any images one may want to analyze in template space.  
\item Perform statistics in template space over the region of interest, e.g. all cortex. 
\end{enumerate}
We now illustrate this procedure using ANTS tools and example images from ANTS/Examples/Data. 
The expected output is in figure~\ref{fig:morph}.
\begin{verbatim}
# 0.  Here, we use r16 as a template and r64 as a subject T1 image. 
# 1. Segment the input images 
  ThresholdImage 2 r16slice.nii mask.nii.gz 0.001 Inf
  Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
      -c [10,0.001] -m [0.05,1] -i kmeans[3] -x mask.nii.gz
  ThresholdImage 2 r64slice.nii mask.nii.gz 0.001 Inf
  Atropos -d 2 -a r64slice.nii -o [r64slice_seg.nii.gz,r64slice_prob_%02d.nii.gz] \
       -c [10,0.001] -m [0.05,1] -i kmeans[3] -x mask.nii.gz
# 2. Compute the deformation
  sh ants.sh 2 r16slice_prob_02.nii.gz r64slice_prob_02.nii.gz SPM 100x100x100
# 3. Smooth the segmented image 
  SmoothImage 2 r64slice_prob_02.nii.gz  1. sr64slice_prob_02.nii.gz  
# 3. Warp the segmented image by the 
  WarpImageMultiTransform 2 sr64slice_prob_02.nii.gz  sr64slice_prob_02_norm.nii.gz \ 
     -R r16slice.nii SPMWarp.nii.gz SPMAffine.txt 
# 4. Compute the log Jacobian from the transformation.
  ANTSJacobian 2 SPMWarp.nii.gz SPM 1  # take the log
  gzip SPMlogjacobian.nii 
# 5. Multiply the log-jacobian by the gray matter probability. 
 ImageMath 2 SPMlogjacobianmask.nii.gz m SPMlogjacobian.nii.gz sr64slice_prob_02_norm.nii.gz
# 6.  Estimate cortical thickness --- use svn revision 684 or greater (not in official release yet). 
 KellySlater 2 r64slice_seg.nii.gz r64slice_prob_03.nii.gz  r64slice_prob_02.nii.gz \
    r64slice_thickness.nii.gz 0.02 50 8 0 0.5
# 7.  Warp to template space. 
  WarpImageMultiTransform 2 r64slice_thickness.nii.gz  r64slice_thickness_norm.nii.gz  \
    -R r16slice.nii SPMWarp.nii.gz SPMAffine.txt 
# 8.  Repeat steps 1-7 for a population and run statistics.
\end{verbatim}
\begin{figure}
\includegraphics[width=1\textwidth]{Figures/morphometry.pdf}
\itkcaption[ANTS Morphometry]{The output from the ANTS morphometry example.}
\label{fig:morph}
\end{figure}
This example uses a standard Jacobian-based approach to morphometry.
Here, we use the log-Jacobian because it is symmetric about zero and
mask with the gray matter segmentation to restrict the analysis to the
cortex.  The Jacobian is discussed in figure~\ref{fig:inv} and visualization of
the Jacobian is shown in figure~\ref{fig:jac}.  This Jacobian comes
from the mapping shown in figure~\ref{fig:large}.
\begin{figure}
\includegraphics[width=1\textwidth]{Figures/Jacobian.pdf}
\itkcaption[ANTS Jacobian]{The Jacobian: Note that the bright values
  in the template image ventricles (left) indicate that the ventricles
  are relatively larger in the subject image. Similarly, the dark
  jacobian values in the individual image show that the template has
  smaller ventricles. This example is described on the large
  deformation page Large Def. Once one gains the Jacobian (or more
  appropriately, the log-Jacobian), then one may compute statistics
  across the population.}
\label{fig:jac}
\end{figure}
Similar analyses may be performed on thickness images, functional 
images or fractional anisotropy images derived from the diffusion 
tensor modality.  In all these cases, steps similar to those above 
would be performed.

\subsection{Statistics with ANTS and {\bf R}}
{\bf R} is an open-source, cutting-edge statistical package 
used by statisticians world-wide.   ANTS is designed to 
interface with {\bf R} via an ImageMath tool that converts 
image sets into an {\bf R} compatible format.  
Three steps are required.  First, convert image data into 
vector and matrix formats through a mask of your statistical 
region of interest.  Second, read that data into {\bf R} and 
run your statistics.  Third, write out the data from {\bf R} and 
convert back to an image.  Here are the steps, in code examples,
\begin{enumerate}
\item Create a vector for your output: 
\begin{verbatim} 
ImageMath 3 testresultvec.nii ConvertImageSetToMatrix 1 mask.nii image1.nii
\end{verbatim}
Here, we assume all images are called image1.nii, image2.nii, ... , imageN.nii so we 
can use a wildcard to pass to ImageMath and 
create a matrix for all your input data: 
\begin{verbatim} 
 ImageMath 3 datamatrix.nii ConvertImageSetToMatrix 1 mask.nii image*.nii 
\end{verbatim}
\item   Run {\bf R} by reading your data above, performing whatever statistics you want and then 
writing out.  
\item   Convert the test result back to an image.  
\begin{verbatim} 
ImageMath 3 testresult.nii ConvertVectorToImage mask.nii testresultvec.nii
\end{verbatim}
\end{enumerate}
Most statistical requirements may be met with this setup. 
\newpage
\section{Dependencies and Related Software}
\subsection{Dependencies and Compilation}
ANTS depends on the most recent version of the Insight ToolKit (ITK)
-- www.itk.org.   File types that may be read and written are 
all of those available in itk and a few more \href{http://picsl.upenn.edu/ANTS/ioants.php}{http://picsl.upenn.edu/ANTS/ioants.php}.

ANTS and ITK are both compiled through pointing CMake -- www.cmake.org
-- to a CMakeLists.txt file.  The ANTS CMakeLists.txt is in the
Examples directory.

See below for more details on compilation and binaries: \newline
\href{http://picsl.upenn.edu/ANTS/download.php}{Compile and download: http://picsl.upenn.edu/ANTS/download.php}.
We maintain a dashboard to give users an idea of what to expect 
from compilation:\newline
\href{http://picsl.upenn.edu/cdash/index.php?project=ANTS}{http://picsl.upenn.edu/cdash/index.php?project=ANTS}

\subsection{Visualization and Quantification of ANTS Results}
Use \href{www.itksnap.org}{itk SNAP : www.itksnap.org } or mricro / mricron for viewing
images.  The ANTS program StackSlices can be used to conveniently scan
through a dataset in any of its axes.  It is useful for checking both
input data quality and output normalization quality.  MeasureImageSimilarity 
will supply numerical measures of image similarity via three different metrics.

\subsection{Statistics Beyond ANTS}
Use SPM, {\bf R} or npm (from Chris Rorden's mricron) for computing population statistics.

\subsection{Pipelining with ANTS}
Use PipeDream for automating large-scale processing:\newline
\href{https://sourceforge.net/projects/neuropipedream/}{PipeDream
  Homepage :
  https://sourceforge.net/projects/neuropipedream/} \newline PipeDream
is integrated with ANTS and automates more complex studies such as
multivariate diffusion tensor and T1 cortical thickness studies,
longitudinal mapping and reconstruction of nifti images from Dicom.
\newpage
\section{Annotated Bibliography}
The {\em de facto} ANTs reference is now
\href{http://www.ncbi.nlm.nih.gov/pubmed/20851191}{``A reproducible
  evaluation of ANTs similarity metric performance in brain image
  registration.'' (this is a link).}  The original statement of the symmetric normalization and template
construction methodology was given in \cite{Avants2004} with a more 
recent template paper here \cite{Avants2009c}. 
A follow up study that used landmark guidance to compare the chimpanzee cortex to
the human cortex was published here \cite{Avants2006} -- this study
used {\em in vivo} MRI and template-based normalization to confirm
volumetric numbers derived from an early 20th century post-mortem
study comparing one human and one chimp.  This conference article has
some additional detail and alternative updates to the methodology, in
particular application to shape-based interpolation
\cite{Avants2005b}.  Network based studies were performed here
\cite{duda08miccai,duda08cvpr}.  The main SyN paper is here
\cite{Avants2008}.  Applications to neurodegeration are here
\cite{Avants2005,Avants2008a,Grossman2008,Avants2009,Das2009,Yushkevich2009,Massimo2009}.
Hippocampus focused work is here \cite{Pluta2009,Yushkevich2009}.  The
main evaluation papers include \cite{Avants2008} and \cite{Klein2009}
for the cortex and deep brain structures whereas \cite{Pluta2009}
evaluates the use of automated and semi-automated normalization for
high-throughput hippocampus morphometry.  An additional evaluation
paper is being developed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  Insert the bibliography using BibTeX
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{plain}
\bibliography{ants}

\end{document}

### 2011-02-18 src/ANTS/Documentation/atropos/atropos.tex
revert to svn from today

--- 2011-02-09 src/ANTS/Documentation/atropos/_region_.tex
@@
-\message{ !name(atropos.tex)}%\documentclass[11pt,english]{article}
+%\documentclass[11pt,english]{article}
 %
@@
 \usepackage{lscape}
-\usepackage[round,semicolon,authoryear,]{natbib}
+\usepackage[square,comma,numbers,sort&compress]{natbib}
 
@@
 
-% \doublespacing
+\doublespacing
 
@@
         caption={\small \baselineskip 12pt Atropos short command line menu which is invoked using the `{\ttfamily -h}' option. 
-        The expanded menu, which provides additional details regarding the possible parameters and usage 
+        The expanded menu, which provides details regarding the possible parameters and usage 
         options, is elicited using the `{\ttfamily {-}{-}help}' option.},
@@
         caption={\small \baselineskip 12pt N4 short command line menu which is invoked using the `{\ttfamily -h}' option. 
-        The expanded menu, which provides additional details regarding the possible parameters and usage 
+        The expanded menu, which provides details regarding the possible parameters and usage 
         options, is elicited using the `{\ttfamily {-}{-}help}' option.  },
@@
 
-\message{ !name(atropos.tex) !offset(-3) }
-
 \normalem
@@
 \begin{center}
-{\Large \bf An Open Source Framework for Multivariate $n$-Tissue
-  Segmentation and Evaluation Using Open Data Sets} \\
+{\Large \bf An Open Source Multivariate Framework for $n$-Tissue
+  Segmentation with Evaluation on Public Data} \\
 \vspace*{0.5cm}
@@
 We introduce {\em Atropos}, an ITK-based multivariate $n$-class
-Expectation Maximization (EM) segmentation tool based on either parametric or non-parametric finite mixture modeling (FMM) of the
-class intensities.  Atropos, available in ANTs
+open source segmentation algorithm distributed
+with ANTs.%
 \footnote{
 http://www.picsl.upenn.edu/ANTs
-}, is capable of incorporating spatial prior probability maps,
+}  The Bayesian formulation of the segmentation problem is 
+solved using the Expectation Maximization (EM) algorithm
+with the modeling of the class intensities 
+ based on either parametric or non-parametric finite mixtures.  
+Atropos is capable of incorporating spatial prior probability maps,
 (sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
@@
 First, we use the BrainWeb dataset from Montreal Neurological
-Institute to evaluate three tissue segmentation performance via (1)
-K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
-derived from a group template; (3) Prior-based segmentation with use
+Institute to evaluate three-tissue segmentation performance via (1)
+K-means segmentation without use of template data; (2) MRF
+segmentation with initialization by prior probability maps derived from a group template; (3) Prior-based segmentation with use
 of spatial prior probability maps derived from a group template.  We
-also evaluate Atropos performance by using spatial priors to drive a 69 class EM segmentation
-problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
+also evaluate Atropos performance by using spatial priors to drive a 69-class EM segmentation
+problem derived from the Hammers atlas from University College London.  These evaluation studies, combined with
 illustrative examples that exercise Atropos options, demonstrate both
@@
 \section{Introduction} As medical image acquisition technology has advanced,
-concomitant investment has been made towards adapting classification
-techniques for neuroanatomy.  Some of the earliest work appropriated NASA
+significant investment has been made towards adapting classification
+techniques for neuroanatomy.  Early work appropriated NASA
 satellite image processing software for statistical classification of
@@
 The problem of accurately delineating the white matter, grey matter
-and cerebrospinal fluid of the primate brain continuously spurs
+and cerebrospinal fluid (and subdivisions) of the human brain continuously spurs
 technical development in segmentation.  Following \cite{Vannier1985},
 many researchers adopted statistical methods for $n$-tissue anatomical
-brain segmentation.  Given the ``missing data'' aspect of this
-problem, use of the Expectation-Maximization (EM) framework is natural\citep{Dempster1977}.  The
-work described in \cite{Wells1996} was one of the first to use EM for
+brain segmentation.  The Expectation-Maximization (EM) framework is
+natural \citep{Dempster1977} given the ``missing data'' aspect of this
+problem.  The work described in \cite{Wells1996} was one of the first to use EM for
 finding a locally optimal solution by iterating between bias field
@@
 and T1 coronal images, which continues to find utility in contemporary
-developments.  A secondary component was an extended nonparametric
+developments.  A secondary component was an extended non-parametric
 probability model, also influenced by earlier work \citep{Kikinis1992}, where
@@
 
-EM may easily be caught in local optima. Consequently, a good
-initialization is critical.  Common low-level initialization steps 
+Several initialization strategies have been proposed to overcome the characteristic
+susceptibility of EM algorithms to local optima.   Common low-level initialization steps 
 include uniform probability assignment \citep{Wells1996},
@@
 neuroanatomical layout with subsequent processing used to conjoin
-subsets of Gaussians belonging to the same tissue classes.  Recently,
-anatomically specific strategies for incorporating prior knowledge
-into the segmentation solution have included the use of spatial prior
-probability maps of the various structures of interest
-\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
-probability maps can also be used to provide an initial segmentation.
+subsets of such Gaussians belonging to the same tissue classes.
+Recently, reseachers have begun to rely on spatial prior probability
+maps of anatomical structures of interest to encode domain knowledge\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
+probability maps may also provide an initial segmentation.
 Related technological developments model partial volume effects for
@@
 optimization scheme. The roots of this very popular software package
-stem back to early work by Karl Friston in which the basis for
-statistical parametric mapping was developed \citep{Friston1990} and
-subsequently provided to the research community.  Similar integrative brain processing was provided in \cite{Pohl2006} in which
+stem back to early work by Karl Friston which laid the basis for
+statistical parametric mapping \citep{Friston1990}. 
+ Similar integrative brain processing was provided in \cite{Pohl2006} in which
 segmentation and registration parameters were optimized simultaneously
@@
 manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
-\citep{Pieper2006}---an open source package with developmental
+\citep{Pieper2006}---an open source medical image computation and
+visualization package with developmental
 contributions from multiple agencies including both private and
-academic institutions. Researchers in aging often focus on
+academic institutions.
+
+Related neuroanatomical research concerns the selection of geometric
+features of the cortex \citep[e.g.][]{Goualher1999} which aims at
+understanding the functional-anatomical relationship of the human
+brain. Recent endeavors produce a dense cortical labeling in which
+every point of the cortex is classified, i.e. a cortical parcellation
+\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Various techniques
+have been proposed to reduce the manual effort required to densely
+label a high-resolution neuroimage; one example is the popular
+software package known as Freesurfer
+\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
+approach detailed in this work, Freesurfer is primarily a
+surface-based technique in which the brain structures such as the
+grey-white matter interface and pial surfaces are processed, analyzed,
+and displayed as tessellated surfaces \citep{Dale1999,Fischl1999}.
+Advantages of surface representations include the ability to map
+processed neuroanatomy to simple geometric primitives such as spheres
+or planes and the ease of including topological constraints in the
+analysis workflow.  These types of methods, including Klein's
+Mindboggle \citep{Klein2005}, would usually follow an initial
+segmentation by a volumetric method such as Atropos.
+
+%{\color{red}{{\em Needs work}:  Mindboggle, Heckemann  One advantage
+%of the pipeline we propose is that we parcellate cortex entirely in
+%the image space, thus avoiding the difficulty of transferring labels
+%from the mesh space back to the image space---a problem that is a confound of surface-based methods\citep{Klein2010}.}}
+
+Researchers in aging often focus on
 accurately segmenting the T1 MRI of elderly controls and subjects
@@
 single method performs best under every measurement and, along with the No
-Free Lunch theorem, highlight the need for segmentation tools that are tunable for different
+Free Lunch theorem \citep{Wolpert1997}, highlight the need for segmentation tools that are tunable for different
 problems and research goals.
 
-Related neuroanatomical research concerns the selection of geometric
-features of the cortex \citep[e.g.][]{Goualher1999} with aims at
-understanding the functional-anatomical relationship of the human
-brain. Recent endeavors produce a dense cortical labeling in which
-every point of the cortex is classified, i.e. a cortical parcellation
-\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
-time-consumption and tedium associated with manual effort towards the
-latter, various techniques have been proposed of which a small subset
-has been publicly availed to the research community such as the
-popular software package known as Freesurfer
-\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
-approach detailed in this work, Freesurfer is primarily a
-surface-based technique in which the brain structures such as the
-grey-white matter interface and pial surfaces are processed, analyzed,
-and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
-Advantages of surface representations include the ability to map
-processed neuroanatomy to simple geometric primitives such as spheres
-or planes and the ease of including topological constraints in the
-analysis workflow.  These types of methods, as well as Klein's
-Mindboggle \cite{Klein2005}, would usually follow an initial segmentation by a volumetric method such as Atropos.
-
-%{\color{red}{{\em Needs work}:  Mindboggle, Heckemann  One advantage
-%of the pipeline we propose is that we parcellate cortex entirely in
-%the image space, thus avoiding the difficulty of transferring labels
-%from the mesh space back to the image space---a problem that is a confound of surface-based methods\citep{Klein2010}.}}
-
 Our open source segmentation tool, which we have dubbed {\em
-Atropos}, 
-\footnote{ Atropos is one of the three Fates from Greek
+Atropos},%
+\footnote{Atropos is one of the three Fates from Greek
 mythology characterized by her dreaded shears used to decide the
 destiny of each mortal.  Also, consistent with the entomological motif
-of our ANTs toolkit, {\it Acherontia atropos} is a species of large
+of our ANTs, {\it Acherontia atropos} is a species of large
 moth known for the skull-like pattern visible on its thorax.  }
-efficiently and flexibly implements a $n$-tissue paradigm for
+efficiently and flexibly implements an $n$-tissue paradigm for
 voxel-based image segmentation.  Atropos allows users to harness its
 generalized EM algorithm for standard tissue classification of
-the brain into gray matter, white matter and cerebrospinal fluid.
+the brain into gray matter, white matter and cerebrospinal fluid even
+in cases of multivariate image data---relevant when more than one 
+view of anatomy aids segmentation, as in neonatal brain tissue classification
+\citep[e.g.][]{Prastawa2005,Weisenfeld2009}.  
 Atropos equally allows incarnations that use EM to simultaneously
-maximize the posterior probabilities of many classes, for instance,
+maximize the posterior probabilities of many classes with minimal
+random access memory requirements, for instance,
 when parcellating the brain into hemispheres, cortical regions and
@@
 to provide a segmentation toolbox that may be modified,
-tuned and refined by the user for different problems.
-
-We also allow Atropos to address the bias field issue by incorporating
-the recently developed N4 bias correction software \citep{Tustison2010} in an adaptive manner.  Coupled with the
-registration \citep{Avants2010b} and template building 
-\citep{Avants2010} already included in the ANTs toolkit, Atropos is a
+tuned and refined for different use scenarios.
+
+Coupled with the registration \citep{Avants2011} and template building
+\citep{Avants2010} already included in the ANTs, Atropos is a
 versatile and powerful software tool which touches multiple aspects of
-our brain processing pipeline from brain extraction
-\citep{Avants2010a} to grey matter/white matter/cerebrospinal fluid
-segmentation to label fusion and propagation for achieving a refined
-cortical parcellation based on both prior information and
-subject-specific image data.
-
-Atropos is multi-platform, free, stand-alone and able to
-manage a wide range of scalar and vector image datatypes.  To highlight the
-value of this open-source contribution, we performed a search of
-software attributes on NITRC and found that as of November 2010 no
-stand-alone EM methods are currently listed.
-Atropos's memory efficiency enables EM-segmentation of many classes at
-once, a capability that is due to a sparse internal representation of
-prior probability images that minimizes the need for random access
-memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
-neuroimages into 69 neuroanatomical regions to illustrate the practical
-value of the low-memory implementation within this paper.  Atropos
-also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
-segmentation, as in neonatal brain tissue classification
-\cite{Weisenfeld2009}.  Although Atropos may be applied to
-multivariate data from arbitrary modalities, we limit our evaluation
-to tissue-classification in T1 neuroimaging in part due to the
-abundance of ``gold-standard'' data for this modality.  Consistent
+our brain processing pipeline.  We use Atropos to address brain
+extraction \citep{Avants2010a}, grey matter/white matter/cerebrospinal
+fluid segmentation, label fusion/propagation and cortical
+parcellation.  We also allow Atropos to interact with the recently
+developed N4 bias correction software \citep{Tustison2010} in an
+adaptive manner.  To further highlight the value of this open source
+contribution, we performed a search of software attributes on NITRC
+and found that as of November 2010 no stand-alone EM methods are
+currently listed.  We also evaluate Atropos performance on two brain
+MRI segmentation objectives.  First, three-tissue classification.
+Second, we test our ability to parcellate the brain into 69
+neuroanatomical regions to illustrate the practical value of the
+low-memory implementation within this paper.  Although Atropos may be
+applied to multivariate data from arbitrary modalities, we limit our
+evaluation to tissue classification in T1 neuroimaging in part due to
+the abundance of ``gold-standard'' data for this modality.  Consistent
 with our advocacy of open science (not to mention the facilitation of
@@
 Organization of this work is as follows: we first describe the theory
-behind the various components of Atropos while
-acknowledging that theoretical discussion is available elsewhere,
-e.g. in previously cited literature.  This is followed by a thorough
-discussion of implementation which, though often overlooked, is of
-immense practical utility.  We then report evaluation results on
-public T1 neuroimaging datasets that have reference gold/silver
-standard labelings.  Finally, we provide a discussion of our results
-and our open-source contribution in the context of the remainder of
-this paper and of previous and future work.
+behind the various components of Atropos while acknowledging that more
+theoretical discussion is available elsewhere.  This is followed by a
+thorough discussion of implementation which, though often overlooked,
+is of immense practical utility.  We then report results on the
+BrainWeb and Hammers dataset.  Finally, we
+provide a discussion of our results and our open source contribution
+in the context of the remainder of this paper and of previous and
+future work.
 
@@
 
-Atropos encodes a family of Bayesian segmentation techniques that may be configured in an application-specific manner.
-The theory underlying Atropos dates back 20$+$ years and is representative of some of the most innovative work in the field.
-Although we synthesize some of the theoretical work in this section, we recommend to the interested reader consultation with the literature cited previously for further insight.
+Atropos encodes a family of Bayesian segmentation techniques that may
+be configured in an application-specific manner.  The theory
+underlying Atropos dates back 20$+$ years and is representative of
+some of the most innovative work in the field.  Although we summarize
+some of the theoretical work in this section, we recommend that 
+the interested reader consult the deep literature in this field for
+additional perspective and proofs behind the major concepts.
 
@@
 prior probability terms.  Bayesian paradigms for brain image
-segmentation employ a user-selected observation model defining the
+segmentation employ a user selected observation model defining the
 likelihood term and one or more prior probability terms.  The product
@@
 given in the literature, has often been formed either as
-MRF-based or template-based.  An image segmentation solution in this
-context is an assignment of one label to each voxel
+MRF-based and/or template-based.  An image segmentation solution in this
+context is an assignment of one label to each voxel%
 \footnote{
@@
 
-\subsection{Segmentation Objective Function} Atropos optimizes a class
-of user-selectable maximization objective functions each of which may
+\subsection{Segmentation Objective Function}
+
+ Atropos optimizes a class
+of user selectable objective functions each of which may
 be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
@@
 p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
-p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{\mathbf{y}}, 
+p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{p(\mathbf{y})} 
 \end{equation} 
@@
 solutions to this problem.  The following sections detail the 
-Atropos EM along with choices for the likelihood(s), prior(s).
+Atropos EM along with choices for the likelihood and prior terms.
 
@@
 the mean and variance of the $k^{th}$ model.  When $y_i$ is a vector
-quantity, we replace the Euclidean distance by by Mahalanobis distance
+quantity, we replace the Euclidean distance by Mahalanobis distance
 and define multivariate Gaussian parameters via a mean vector,
-$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma_k}$.
+$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma}_k$.
 
@@
 \end{align}
-where $N_B$ is the number of bins used to define the histogram of the sample observations (in Atropos the default is $N_B = 32$) and $c_j$ is the center of $j^{th}$ bin in the histogram.
+where $N_B$ is the number of bins used to define the histogram of the
+sample observations (in Atropos the default is $N_B = 32$) and $c_j$
+is the center of the $j^{th}$ bin in the histogram.
 $\sigma_j$ is the width of each of the $N_B$ Gaussian kernels.
@@
 possible).  The MRF locality condition is then, 
-\begin{align}
-  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right),
+\begin{align}\label{eq:cond}
+  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i |
+    x_{\mathcal{N}_i}\right) 
 \end{align}
@@
 \begin{align} \label{eq:gibbs}
-p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
-\end{align}
-$Z$ is a normalization factor known as the {\em partition function}
-and $U(\mathbf{x})$ is the {\em energy function} which can take
+p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right) 
+\end{align}
+with $Z$ a normalization factor known as the {\em partition function}
+and $U(\mathbf{x})$ the {\em energy function} which can take
 several forms \citep{Li2001}. In Atropos, as is the case with many
@@
 \begin{align}
-U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
+U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in
+    \mathcal{N}_i} V_{ij}( x_i, x_j ) \right) \nonumber
 \end{align}
@@
                           \begin{array}{ll}
-                            1 & \text{if } x_i = x_j \\
-                            0 & \text{otherwise}
+                            0 & \text{if } x_i = x_j \\
+                            1 & \text{otherwise}
                           \end{array}
@@
 \end{align}
-and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
+and $\beta$ is a granularity term which weights the contribution of
+the MRF prior on the segmentation solution.  
 Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that
@@
 \end{align}
-so that sites in the neighborhood closer to $i$ are weighted more heavily than sites further away.
+so that sites in the neighborhood closer to $i$ are weighted more
+heavily than distant sites.
 
@@
 population from which a template is constructed \citep[e.g.][which is
-also available in the ANTs toolkit]{Avants2010}.  Each labeling can
+also available in ANTs]{Avants2010}.  Each labeling can
 then be warped to the template where the synthesis of warped labeled
@@
 the spatial distribution of labeled anatomy which can be harnessed in
-joint segmentation/registration or, specifically, Atropos/ANTs hybrids involving unlabeled subjects.
+joint segmentation/registration or Atropos/ANTs hybrids involving unlabeled subjects.
 
@@
 \begin{align}
-\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}}.
+\mathrm{Pr}(x_i = l_k) = \frac{t_{ik}}{\sum_{j=1}^K t_{ij}}
 \end{align}
@@
 posterior) at a subset of $\mathcal{I}$ to be $1$ for the known label
-and $0$ for each other label at the same site.  The user-input
+and $0$ for each other label at the same site.  The user input
 therefore not only initializes the optimization, but also gives boundary conditions
@@
 classes.  Thus, in most practical applications, EM is a
-reasonable and efficient alterative to Boykov's solution.
-Furthermore, one may automate the
-initialization process.
-
+reasonable and efficient alternative to Boykov's solution.
+Furthermore, one may automate the initialization process.  
+{\color{blue}{We provide this capability to allow the user to implement
+    an interactive editing and segmentation loop.  The user may
+    run Atropos with sparse manual label guidance, evaluate the
+    results, update the manual labels and repeat until achieving the desired
+    outcome.  This processing loop may be performed easily with, e.g., ITK-SNAP.}}
 \begin{comment}
@@
 \end{align}
-The probability calculations are illustrated in Fig.~\ref{fig:distancePrior} and were designed so that the most interior point, or set of points, of region $R_k$ would have a distance prior probability of 1.0 which would linearly decrease within the region to the boundary with a user-selected probability value of $\alpha_k$.
-Outside $R_k$, the distance prior exponentially decays with a decay rate governed by another user-selected parameter $\sigma_k$.
+The probability calculations are illustrated in
+Fig.~\ref{fig:distancePrior} and were designed so that the most
+interior point, or set of points, of region $R_k$ would have a
+distance prior probability of 1.0 which would linearly decrease within
+the region to the boundary with a user selected probability value of
+$\alpha_k$.  Outside $R_k$, the distance prior exponentially decays
+with a decay rate governed by another user selected parameter $\sigma_k$.
 This permits a dense segmentation throughout the segmentation region of $\mathcal{I}$ even where the template prior probabilities are zero.
@@
 Atropos uses expectation maximization to find a locally optimal
-solution for the user-selected version of the Bayesian segmentation
+solution for the user selected version of the Bayesian segmentation
 problem (cf. Eqn. (\ref{eq:bayes})).
@@
 as a constant.  The Atropos implementation estimates $\gamma_k$ at
-each iteration as in \cite{Ashburner2005}.
+each iteration, similar to \cite{Ashburner2005}.%
 \footnote{
@@
 Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.
-Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
+For large $K \gg 3$, exact optimization is even more intractable than for the traditional 3-tissue scenario.
 }
-Although many optimization techniques exist \citep[see the
-introduction in][for a concise summary of the myriad optimization
-possibilities]{Marroquin2002}---each with their characteristic
+Although many optimization techniques exist (see the introduction in 
+\cite{Marroquin2002} for a concise summary of the myriad optimization
+possibilities)---each with their characteristic
 advantages and disadvantages in terms of computational complexity and
@@
 classes to segment.  The simplest initialization is by the classic
-K-means algorithm with only the number of classes as a prior.
+K-means or Otsu thresholding algorithms with only the number of classes 
+specified by the user.
 Otherwise, the user must provide prior information for each class in
 the form of either a single $n$-ary prior label image or a series of
-prior probability images, one for each class.
-\item[E-Step: ICM Label Update:] Given the initialization and fixed
-model parameters, Atropos will evaluate the posterior probability for
-each possible label and assign, to each voxel, the label that
-maximizes the posterior.  {\bf \textcolor{red}{ FIXME more specific?}}.
-\item[M-Step: Parameter Update:] 
-The posteriors used in this step are called $p^{old}$ to make it
-clear that we use the previous parameter estimates.
+prior probability images, one for each class.  The initialization also
+provides starter parameters.
+\item[Label Update (E-Step):] Given the initialization and fixed
+model parameters, {\color{blue}{Atropos is capable of updating the current label 
+estimates using either a {\em synchronous} or {\em asynchronous} scheme.
+The former is characterized by iterating through the image and determining
+which label maximizes the posterior probability without updating any
+labels until all voxels in the mask have been visited at which point
+all the voxel labels are updated simultaneously (hence the descriptor 
+``synchronous'').  This option is specified with \texttt{--icm [0]}. 
+However, unlike asynchronous schemes characteristic of ICM, synchronous 
+updates lack convergence guarantees.  To determine the labeling 
+which maximizes the posterior probability for the asynchronous option, 
+an ``ICM code'' image is created once for all iterations by iterating 
+through the image and assigning an ICM code label to each voxel in the 
+mask such that each MRF neighborhood has a non-repeating code label set.  
+Thus each masked voxel in the ICM code image is assigned a value in the 
+range $\{1,\ldots,C\}$ where $C$ is the maximum code label.
+Such an image can be created and viewed with Atropos by assigning a valid
+filename in the \texttt{--icm [1]} set of options.  An example adult brain
+slice and the associated code image is given in 
+Figure \ref{fig:r16icm} for an MRF neighborhood of $5 \times 5$ pixels.
+This produces a maximum code label of `13'.
+For each iteration, one has the option to permute the
+set $\{1,\ldots,C\}$ which prescribes the order in which the
+voxel labels are updated asynchronously. After the first pass through the
+set of code labels, additional passes can further increase the posterior
+probability until convergence (in $\sim$5 iterations).  One can specify
+a maximum number of these ``ICM iterations'' on the command line.  For
+our example in Figure \ref{fig:r16icm}, this means that for each
+ICM iteration, we would iterate through the image 13 times only updating
+those segmentation labels associated with the current ICM code.}}
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{c}
+\includegraphics[width=160mm]{ICMfigure.pdf}
+%\includegraphics[width=70mm]{r16slice.png} &
+%\includegraphics[width=70mm]{r16icm.png} \\
+%(a) & (b) \\
+\end{tabular}
+\end{center}
+\caption{\baselineskip 12pt \small {\color{blue}{An adult brain image slice is shown 
+with its ICM code image corresponding to a $5\times5$ MRF neighborhood.  
+To the right of the ICM code image, we focus on a single neighborhood
+with a center voxel associated with the ICM code label of '10'.  Each 
+center voxel in a specified neighborhood exhibits a unique ICM code
+label which does not appear elsewhere in its neighborhood.  When performing
+the segmentation labeling update for ICM, we iterate through the set
+of ICM code labels and, for each code label, we iterate through the image
+and update only those voxels associated with the current code label.}}
+  }
+\label{fig:r16icm}
+\end{figure}
+
+% {\bf \textcolor{red}{ FIXME more specific?}}.
+\item[Parameter Update (M-Step):] 
+Note that the posteriors used in the previous iteration are used to 
+estimate the parameters at the current iteration.  
 We use a common and elementary estimate of the mixing parameters:
 \begin{align}
-\gamma^{new}_k \leftarrow  \sum_i \frac{p^{old}_{ik}}{\sum_{j=1}^K p^{old}_{ij}}. 
-\end{align}
-For parametric model update, for each of
-  $K$ labels, we compute the mean,
-$$ \mu^{new}_k \leftarrow \frac{ \sum_{i=1}^N  y_i p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) },
-$$
-and variance {\bf \textcolor{red}{ FIXME should this be the full
-    covariance matrix?}},
-$$ \sigma^{new}_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) }.
-$$
+\gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i).
+\end{align}
+We update the parametric model parameters by computing, for each of
+  $K$ labels, the mean,
+\begin{align}  
+ \mu_k \leftarrow \frac{ \sum_{i=1}^N  y_i p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) }
+\end{align}
+and variance,
+\begin{align}  
+ \sigma^2_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k )^T
+   p_k(l_k|y_i) (y_i - \mu_k )}{ \sum_{i=1}^N p_k(l_k|y_i) }.
+\end{align}
+The latter two quantities are modified, respectively, in the case of multivariate data as follows:
+\begin{align}
+ \boldsymbol{\mu}_k \leftarrow \frac{ \sum_{i=1}^N  \mathbf{y}_i p_k(l_k|\mathbf{y}_i)}{ \sum_{i=1}^N p_k(l_k|\mathbf{y}_i) }
+\end{align}
+and the $k^{th}$ covariance matrix, $\boldsymbol{\Sigma}_k$, is calculated from
+\begin{align}
+ \boldsymbol{\Sigma}_{k} \leftarrow \frac{ \sum_{i=1}^N  p_k(l_k|\mathbf{y}_i) ( \mathbf{y}_{i} - \boldsymbol{\mu}_{k} )^{\mathrm{T}}  ( \mathbf{y}_{i} - \boldsymbol{\mu}_{k} )}{1 - \sum_{i=1}^N p^2_k(l_k|\mathbf{y}_{i}) }.
+\end{align}
 This type of update is known as {\em soft} EM.  Hard EM, in
-contrast, only uses sites containing label $l_k$ to update model $k$ parameters.
-A similar pattern is used in the multivariate and non-parametric cases.
+contrast, only uses sites containing label $l_k$ to update the 
+parameters for the $k^{th}$ model.
+A similar pattern is used in non-parametric cases.
 \end{description}
-EM will iterate toward a local maximum.  We track
-convergence by summing up the maximum posterior
-probability at each site over the segmentation domain.  
+
+EM will iterate toward a local maximum.  We track convergence by
+summing up the maximum posterior probability at each site over the
+segmentation domain.  The E-step, above, depends upon the selected
+coding strategy \citep{Besag1986}.  Atropos may use either a
+classical, sequential checkerboard update or a synchronous update of
+the labels, the latter of which is commonly used in practice.
+Synchronous update does not guarantee convergence but we employ it by
+default due to its intrinsic parallelism and speed.  The user may
+alternatively select checkerboard update if he or she desires
+theoretical convergence guarantees.  However, we have not identified
+performance differences, relative to ground truth, that convince us of
+the absolute superiority of one approach over the other.
 %\begin{align}
@@
 \section{Implementation} 
-
+Organization of the implementation section roughly follows that of the
+theory section.  
+%Thus, the reader can refer to, for instance, 2.2 for
+%the likelihood theory and 3.3 for likelihood implementation.
 \begin{figure}
@@
 distributed with the rest of the ANTs package, we plan to contribute
-the Atropos to the Insight Toolkit where it can be vetted and further
-improved by other interested researchers.
-
-An overview of the core components of Atropos can be gleaned, in part, from the flowchart depicted in 
+its core elements to the Insight Toolkit where it can be vetted and improved by other interested researchers.
+
+An overview of Atropos components can be gleaned, in part, from the flowchart depicted in 
 Fig.~\ref{fig:flowchart}.
+{\color{blue}{Given a set of input images and a mask images, each is preprocessed using N4 to correct for intensity inhomogeneity.
+For our brain processing pipeline, the mask is usually obtained from the standard skull-stripping preprocessing step which also uses Atropos.
+Initialization can be performed in various ways using standard clustering techniques, such as K-means, to prior-based images.
+This initialization is used to provide the initial estimate of the parameters of the likelihood model for each class.
+These likelihoods combine with the current labeling to generate the current estimate of the posterior probabilities at each voxel for each class.
+At each iteration, one can also integrate N4 by using the current posterior probability estimation of the white matter to update the estimate of bias field.
+}}
+
 To provide a more intuitive interface without the overhead costs of a graphical user interface, a set of unique command line parsing classes were developed which can also provide insight to the functionality of Atropos.
-The short version of the command line help menu is given in Listing \ref{listing:command} which is invoked by typing `{\ttfamily Atropos -h}' at the command prompt.
-Both short and long option flags are available and each option has its own set of possible values and parameters introduced in a more formal way in both previous discussion and related papers cited in the introduction.
-Here we  describe these options from the unique perspective of  implementation.
+The short version of the command line help menu is given in Listing
+\ref{listing:command} which is invoked by typing `{\ttfamily Atropos
+  -h}' at the command prompt.  Both short and long option flags are
+available and each option has its own set of possible values and
+parameters introduced in a more formal way in both the previous discussion and related papers cited in the introduction.  Here we describe these options from the unique perspective of  implementation.
 
@@
                           PriorLabelImage[numberOfClasses,labelImage,priorWeighting]
+     -p, --posterior-formulation Socrates[<useMixtureModelProportions=1>,
+                                   <initialAnnealingTemperature=1>,<annealingRate=1>,
+                                   <minimumTemperature=0.1>]
+                                 Plato[<useMixtureModelProportions=1>,
+                                   <initialAnnealingTemperature=1>,<annealingRate=1>,
+                                   <minimumTemperature=0.1>]
      -x, --mask-image maskImageFilename
@@
      -m, --mrf [<smoothingFactor=0.3>,<radius=1x1x...>]
+     -g, --icm [<useAsynchronousUpdate=1>,<maximumNumberOfICMIterations=1>,
+                <icmCodeImage=`'>]     
      -o, --output [classifiedImage,<posteriorProbabilityImageFileNameFormat>]
@@
 
+\doublespacing
 \subsection{Initializing the Atropos Objective}
@@
 \item[Input image mask:] This binary image
-  will define the spatial segmentation domain.  Regions outside the
-  mask receive the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
-\item[Convergence criteria:] The algorithm terminates if it reaches max-iterations or
-  produces a change less than the minimum change in the posterior.
+  defines the spatial segmentation domain.  Voxels outside the
+  masked region are designated with the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
+\item[Convergence criteria:] The algorithm terminates if it reaches  the maximum number of iterations or
+  produces a change less than the minimum threshold change in the posterior.
   E.g. {\ttfamily -c [5,1.e-5]}. 
@@
   spatial smoothness of the label map is $\beta$.  A useful range of
-  $\beta$ values is $0$ to $0.5$ where we usually use $0.05, 0.1, 0.2$
-  in brain segmentation.  E.g.  {\ttfamily -m [ 0.1, 1x1x1 ]} would
+  $\beta$ values is $0$ to $0.5$ where we usually use 0.05, 0.1 or 0.2
+  in brain segmentation.  E.g.  {\ttfamily -m [0.1, 1x1x1]} would
   define $\beta=0.1$ with a MRF radius of one voxel in each of three dimensions.
@@
 \begin{itemize}
-\item  {\ttfamily -i  kmeans[3]} standard kmeans initialization for
-  three classes.
-\item  {\ttfamily -i  PriorLabelImage[3, label\_image.nii.gz , w ]}
-  where scalar $w \in [0,1]$ is unused. 
-\item  {\ttfamily -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz ,
-    w ]} where $w=0$~(use the prior probability images only for
-  initialization) or $w=0.5$ (use the prior probability images
-  throughout the optimization).  If one chooses $0 < w < 0.5$ then one
+\item  {\ttfamily -i  Kmeans[3]} standard K-means initialization for
+  three classes,
+\item  {\ttfamily -i  PriorLabelImage[3,label\_image.nii.gz]} and
+%  where scalar $w \in [0,1]$ is unused. 
+\item  {\ttfamily -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,w]} 
+where $w=0$~(use the prior probability images only for
+  initialization) or $w>0.0$ (use the prior probability images
+  throughout the optimization).  If one chooses $0 < w < 1.0$ then one
   will increase (from zero) the weight on the priors.  These images,
-  like the PriorLabelImage, should be defined with the same domain as
+  like the {\ttfamily PriorLabelImage}, should be defined with the same domain as
   the input images to be segmented.  
@@
   mixture proportions (or not) by setting  {\ttfamily -p Socrates[1]}
-  or {\ttfamily -p Socrates[0] }.  Fixed label boundary conditions may be employed by
-  selecting the PriorLabelImage initialization and   {\ttfamily -p Plato[0] }.
+  or {\ttfamily -p Socrates[0]}.  Fixed label boundary conditions may be employed by
+  selecting the {\ttfamily PriorLabelImage} initialization and   {\ttfamily -p Plato[0]}.
 \item[Output:] Atropos will output the hard segmentation and the
   probability image for each model.  E.g.  {\ttfamily -o
-    [segmentation.nii.gz , seg\_prob\%02d.nii.gz] } will write
+    [segmentation.nii.gz,seg\_prob\%02d.nii.gz]} will write
   out the hard segmentation in the first output parameter and a
-  probability image for each class named, here, seg\_prob01.nii.gz ,
+  probability image for each class named, here, seg\_prob01.nii.gz,
   seg\_prob02.nii.gz, etc. 
@@
 Higher dimensions than 4 are possible although we have not yet encountered such 
-application-specific need.  Multiple images (assumed to be of the same
+an application-specific need.  Multiple images (assumed to be of the same
 dimension, size, origin, etc.), will automatically enable multivariate
-likelihoods.  In that case, the first image specified on the 
-command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
-refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
-from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
-follow-up multivariate K-means labeling.    More details on each of
-the key implementation options are given below.
+likelihoods.  In that case, the first image specified on the command
+line is used to initialize the Random, Otsu, or K-means labeling with
+the latter initialization refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is
+determined from the first intensity image which, along with the other
+images, provides the starting multivariate cluster centers for a
+follow-up multivariate K-means labeling.  More details on each of the
+key implementation options are given below.
 
@@
 opted for different likelihood models which have included either
-parametric, specifically Gaussian, or nonparametric variations.
+parametric (Gaussian) or non-parametric variations.
 However, these approaches are similar in that they require a list
@@
 segmentation) to the modeling function, regardless of whether the
-function is parameter or non-parametric.  These similarities permitted
+function is parametric or non-parametric.  These similarities permitted
 the creation of a generic plug-in architecture where classes
-describing both parametric and nonparametric observational models are
+describing both parametric and non-parametric observational models are
 all derived from an abstract list sample function class.  Three
 likelihood classes have been developed, one parametric and two
-nonparametric, and are available for usage although one of the
-nonparametric classes is still in experimental development.  The
+non-parametric, and are available for usage although one of the
+non-parametric classes is still in experimental development.  The
 plug-in architecture even permits mixing likelihood models with
@@
 \subsection{Prior Probability Models}
+\subsubsection{Label Regularity}
 Consistent with our previous discussion, we offer both an MRF-based
@@
 neighborhood radius of 1 in all 3 dimensions.  This radius is defined
-such that voxels including but not limited to those that are face
-connected will influence the MRF.
+such that voxels including but not limited to those that are face-connected will influence the MRF.
 
@@
 }\end{comment}
-
-\subsection{Integration with Registration and N4 Bias Correction} 
-Much of the image processing in our lab takes advantage of multiple
-open source tools.  Most critically, these involve registration and
-bias correction, in addition to segmentation. 
-\subsubsection{Registration of Probability Maps} 
-Image registration enables the ability to transform information
+\doublespacing
+\subsubsection{Registration and Probability Maps} 
+Image registration enables one to transfer information
 between spatial domains which may aid in both segmentation and bias
@@
 \end{command}
-\subsubsection{Integration of N4 Bias Correction}
-The typical segmentation processing pipeline begins with an intensity
-normalization/bias correction step using the recently developed N4
-algorithm \citep{Tustison2010} which is based on the popular
-nonparametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
-several methods proposed in the literature have taken an integrative
-view of the segmentation problem by incorporating an intrinsic bias
-correction step into the actual workflow.  The
-advancements introduced with N4 permit such an adaptive
-integration with Atropos.  Recent demonstrations suggest improved
+\doublespacing
+\subsubsection{Integrating N4 Bias Correction}
+Assumptions about bias correction may be thought of as another prior
+model.  As such, the typical segmentation processing pipeline begins with an intensity
+normalization/bias correction step using a method such as the recently developed N4
+algorithm \citep{Tustison2010}.  {\color{blue}{N4 extends the popular
+nonparametric nonuniform intensity normalization (N3) algorithm \citep{Sled1998}
+in two principal ways:
+\begin{itemize}
+  \item We replace the least squares B-spline fitting with a parallelizable alternative 
+  (which we also made publicly available in the Insight Toolkit)---
+  the advantages being that 1) computation is much faster 
+  and 2) smoothing is not susceptible to outliers as is characteristic with
+  standard least squares fitting algorithms.
+  \item The fitting algorithm permits a multi-resolution approach so whereas standard 
+  N3 practice is to select a single resolution at which bias correction occurs, the N4
+  framework permits a multi-resolution correction where a base resolution is chosen and
+  correction can then occur at multiple resolution levels each resolution being twice 
+  the resolution of the previous level.
+\end{itemize}  
+Specifically, with respect to segmentation, there exists a third advantage with N4 over
+N3 in that the former permits the specification of a probabilistic mask as opposed 
+to a binary mask.}}
+%It was previously mentioned that
+%several methods proposed in the literature have taken an integrative
+%view of the segmentation problem by incorporating an intrinsic bias
+%correction step into the actual workflow.  The
+%advances introduced with N4 permit such an adaptive
+%integration with Atropos.  
+Recent demonstrations suggest improved
 white matter segmentation produces better gain field estimates using
@@
 corresponding to the white matter produced as output in the Atropos
-call, i.e. `{\ttfamily Atropos --output}'.  N4 is
-currently being added as a standard part of the Insight ToolKit.  The
+call, i.e. `{\ttfamily Atropos --output}'.  N4 was recently added  
+to the Insight Toolkit repository%
+\footnote{
+http://www.itk.org/Doxygen/html/classitk\_1\_1N4MRIBiasFieldCorrectionImageFilter.html
+} 
+where it is built and tested on
+multiple platforms nightly.  The
 evaluation section will illustrate inclusion of Atropos, N4 and ANTs
-in a brain processing pipeline.  An example of this procedure, using brainweb data with
-40\% RF bias, is in figure~\ref{fig:bwebrf40}.  We supply the
-information necessary to repeat this example in the script entitled
-`{\ttfamily  atroposBwebRF40FigureExample.sh}' which is available in the ANTs toolkit's Atropos
-documentation folder as of svn commit 711.
+in a brain processing pipeline.
+
+\subsection{Running the Atropos Optimization} 
+The Atropos algorithm is cross-platform and compiles on, at minimum,
+modern OSX, Windows and Linux-based operating systems.  The
+user interface may be reached through the operating system's user
+terminal.  Because of its portability and low-level efficiency,
+Atropos can easily be called from within other packages, such as
+Matlab or Slicer, or, alternatively, integrated at compile time as a
+library.  A typical call to the algorithm, illustrated here with ANTs
+example data, is:
+{\ttfamily Atropos -d 2 -a r16slice.nii.gz -i kmeans[3] -c [5,0] -x
+mask.nii.gz -m [0.2,1x1] -o
+[r16\_seg.nii.gz,r16\_prob\_\%02d.nii.gz]}.  In this case, Atropos
+will output the segmentation image, the per-class probability images
+and a listing of the parameters used to set up the algorithm.  A
+useful feature is that one may re-initialize the Atropos EM via the
+{\ttfamily -i PriorProbabilityImages[...]} option.  This feature
+allows one to compute an initial segmentation via K-means, alter the
+output probabilities by externally computed
+functions (e.g. Gaussian smoothing, image similarity or edge maps) and re-estimate the
+segmentation with the modified priors.  Finally, the functionality
+that is available to parametric models is equally available to the
+non-parametric models enabled by Atropos.
+
+\section{Evaluation}
+Atropos~encodes a family of segmentation techniques that may be
+instantiated for different applications but here we evaluate only two
+of the many possibilities. First, we perform an evaluation on the
+BrainWeb dataset using both the standard T1 image with multiple bias
+and noise levels and also the BrainWeb20 data
+\citep{Aubert-Broche2006a,Battaglini2008}.  In combination, these data
+allow one to vary not only noise and bias but also the underlying
+anatomy.  Second, we evaluate the use of Atropos in improving whole-brain parcellation and exercise its ability to efficiently solve {\em
+many-class} expectation maximization problem.  We choose this
+evaluation problem in part to illustrate the flexibility of Atropos
+and also the benefits of the novel, efficient implementation that
+allows many-class problems to be solved with low memory usage ($<$2GB
+for a 69-class model on 1 mm$^3$ brain data).
+
+\subsection{BrainWeb Evaluation}
+\label{sec:bweb} The BrainWeb data is freely available at
+\url{http://mouldy.bic.mni.mcgill.ca/BrainWeb/}.  We employ both the
+individual subject data and the BrainWeb20 data in this evaluation.
+\subsubsection{Single-Subject Evaluation} We use the single-subject
+data with 3\% noise and three levels of bias referred to as 0, 20 and 40\% RF inhomogeneity.  We
+study the effect of the MRF prior term and initialization on
+the Dice overlap between ground truth and the segmentation result for
+each tissue.  We test both K-means and prior label image
+initialization with MRF $\beta \in \{ 0.00 , 0.05 , 0.10 , 0.15 , 0.20, 
+0.25 , 0.30 \}$ at each bias field.  We also feed the white matter
+probability map derived from K-means into N4 to guide the bias
+correction.%
+\footnote{
+A comprehensive evaluation of N4 reported in \cite{Tustison2010} used the BrainWeb data set
+to compare performance with the original N3 algorithm \citep{Sled1998}.
+}
+Segmentation is then repeated, with the same parameters,
+but with the N4-corrected image as input.  The resulting algorithm is
+similar to those that fix segmentation parameters while
+estimating bias and fix bias while estimating segmentation parameters.
+Thus, with this simple evaluation, we are able to compare the impact
+of bias on the combination of N4 and Atropos and also the validity of our
+prior label image initialization.  
+% The prior label image manual initialization was based on a crude
+% ITK-SNAP labeling of a small set of voxels for each tissue.  
+Results of these evaluation scenarios, in terms of
+Dice overlap, are shown in Figure~\ref{fig:bweb1}. Because overlap
+ratios with N4 bias correction approximate those of the zero bias data, we may
+conclude that simple N4 pre-processing is adequate to correct even the 40\%
+RF bias level.  An example of this procedure, using BrainWeb data with 40\% RF bias,
+is in Figure~\ref{fig:bwebrf40}.  We supply the information necessary
+to repeat the results in this figure in the script entitled
+`{\ttfamily atroposBwebRF40FigureExample.sh}' which is available in
+the ANTs Atropos documentation folder as of SVN commit 711.
+The script may be easily modified to run the whole
+evaluation.  Figure~\ref{fig:bwebrf40} shows the results of
+simultaneously using proton density and T1-weighted BrainWeb data to
+perform the segmentation.  This multivariate input data outperforms
+the univariate T1-weighted data alone.
 \begin{figure}
@@
 \caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
-sequential processing and apply to brainweb single-subject data with
-40\% RF bias and 3\% noise.  The $\beta$ for the MRF term is, here,
-$0.2$.  Slice 71 of the input data is in (a).  The initial K-means
-($K=3$) segmentation is in panel (b).  We use the brain mask to guide
-N4 bias correction and produce the image, shown in (c).  We repeat the
-K-means segmentation, but with the N4 corrected image as input, and
-produce the segmentation in (d).  The average 3-tissue Dice overlap of
-result (b) is 0.906 while the average overlap for (d) is 0.954.
-Arrows highlight a region of large before-after segmentation
-discrepancy.}
+sequential processing and apply to BrainWeb T1-weighted single-subject
+data with 40\% RF bias and 3\% noise.  The $\beta$ for the MRF term
+is, here, $0.2$.  Slice 71 of the input data is in (a).  The initial
+K-means ($K=3$) segmentation is in panel (b).  We use the brain mask
+to guide N4 bias correction and produce the image in (c).  We
+repeat the K-means segmentation, but with the N4-corrected image as
+input and produce the segmentation in (d).  The average 3-tissue Dice
+overlap of result (b) is 0.906 while the average overlap for (d) is
+0.954.  Arrows highlight a region of large before-after segmentation
+discrepancy.  In (e) we see the BrainWeb proton density image with no
+inhomogeneity and 3\% noise.  Its segmentation is in (f) with average
+3-tissue Dice overlap of 0.895.  In (g) we use both proton density
+data and T1 data as multiple modality input to Atropos.  The segmentation
+of this two-modality input data, using a multivariate Gaussian model,
+produces average 3-tissue Dice overlap of 0.958, which exceeds
+the univariate solution.  An arrow highlights one region where there
+is small, visually recognizable improvement in sulcal segmentation
+relative to the result from T1 data alone.  A second area of
+improvement is the putamen segmentation.  The ground truth
+segmentation is in (h).  The multivariate segmentation result, in
+combination with the low PD segmentation performance, suggests PD and
+T1 provide complementary information that may improve 3-tissue
+segmentation and serves to validate the multivariate Atropos
+implementation.  In this case, the benefit is likely to derive from
+the fact that the PD image has no bias.
+}
 \label{fig:bwebrf40}
@@
 
-\section{Evaluation}
-Atropos~encodes a family of segmentation techniques that may be
-instantiated for different applications but here we evaluate only two
-of the many possibilites. First, we perform an evaluation on the
-BrainWeb dataset using both the standard T1 image with multiple bias
-and noise levels and also the BrainWeb20 data
-\citep{Aubert-Broche2006a,Battaglini2008}.  In combination, this data
-allows one to vary not only noise and bias but also the underlying
-anatomy.  Second, we evaluate the use of Atropos in improving whole
-brain parcellation and exercise its ability to efficiently solve {\em
-many class} expectation maximization problem.  We choose this
-evaluation problem in part to illustrate the flexibility of Atropos
-and also the benefits of the novel, efficient implementation that
-allows many-class problems to be solved with low memory usage ($<$2GB
-for a 69 class model on 1mm$^3$ brain data).
-
-\subsection{BrainWeb Evaluation}
-\label{sec:bweb} The BrainWeb data is freely available at
-\url{http://mouldy.bic.mni.mcgill.ca/brainweb/}.  We employ both the
-individual subject data and the BrainWeb20 data in this evaluation.
-\subsubsection{Single-Subject Evaluation} We use the single-subject
-data with 3\% noise and bias 0, 20 and 40 \% RF inhomogeneity.  We
-study the effect of the MRF prior term and initialization on
-the Dice overlap between ground truth and the segmentation result for
-each tissue.  We test both K-means and prior label image
-initialization with MRF $\beta \in \{ 0.00 , 0.05 , 0.10 , 0.15 , 0.20
-, 0.25 , 0.30 \}$ at each bias field.  We also feed the white matter
-probability map derived from K-means into N4 to guide the bias
-correction.  Segmentation is then repeated, with the same parameters,
-but with the N4 corrected image as input.  The resulting algorithm is
-similar to those that fix segmentation parameters while
-estimating bias and fix bias while estimating segmentation parameters.
-Thus, with this simple evaluation, we are able to compare the impact
-of bias on our N4 and Atropos and also the validity of our
-prior label image initialization.  The prior label image manual
-initialization was based on a crude ITK-SNAP labeling of a small
-set of voxels for each tissue.  Results, in terms of
-Dice overlap, are shown in figure~\ref{fig:bweb1}  Because overlap
-ratios with N4 bias correction approximate those of the zero bias data, we may
-conclude that simple N4 pre-processing is adequate to correct even the 40\%
-bias level.
 \begin{figure}
@@
 \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
-\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
-\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
-\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
-\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
+\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}
+% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
+% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
+% \includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
 \end{tabular}
-\caption{\baselineskip 12pt \small Brainweb single-subject results for
+\caption{\baselineskip 12pt \small BrainWeb single-subject results for
   each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
   bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
-  appears to be between $0.1$ and $0.2$.  The prior label image
-  initialization is competitive with but inferior to K-means.  The
-  legend is in the same position in each graph, allowing a visual
-  comparison of the results.  As one may see, the N4 assisted overlap
+  appears to be between $0.1$ and $0.2$.  
+% The prior label image initialization is competitive with but inferior to K-means.  This is
+%  sensible in that the prior labels image outcome is sensitive to the
+%  user choices made.  In this study, we labeled the brain web template
+%  very sparsely, quickly and crudely and did not expect highly
+%  accurate results.  
+  The legend is in the same position in each graph, allowing a visual
+  comparison of the results.  As one may see, the N4-assisted overlap
   values are consistent across bias field/RF inhomogeneity.}
@@
 \end{figure}
-\subsubsection{20 Subject Evaluation} The single-subject brainweb
+\subsubsection{20-subject Evaluation} The single-subject BrainWeb
 study in the previous section tested the basic Atropos options and the
 benefit of N4 for segmentation in the presence of bias.  The 20
-subject brainweb data allows us to use 2-fold cross-validation to test
+subject BrainWeb data allows us to use 2-fold cross-validation to test
 our ability to segment different individuals reliably.  In this study,
@@
 the model parameters but not during subsequent EM iterations.  
-When $w=1$, the priors are only maintained in the product with the
+When $w=0.5$, the priors are maintained in the product with the
 likelihood during all EM iterations.  Results, in terms of bar
-plots for Dice overlap mean and standard deviation, are shown in figure~\ref{fig:bweb20}.
+plots for Dice overlap mean and standard deviation, are shown in Figure~\ref{fig:bweb20}.
 \begin{figure}
@@
 \begin{tabular}{c}
-\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
-\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
-\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
+\includegraphics[width=2.2in]{Figures/bw20_kmeans.pdf}
+\includegraphics[width=2.2in]{Figures/bw20_priorIni.pdf}
+\includegraphics[width=2.2in]{Figures/bw20_priorSeg.pdf}
 \end{tabular}
-\caption{\baselineskip 12pt \small Brainweb 20-subject results for
+\caption{\baselineskip 12pt \small BrainWeb 20-subject results for
   each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
@@
 label.  Following \cite{Avants2010a}, the initialization of our
-evaluation applies the script {\verb ants_multitemplate_labeling.sh }
-(available in the ANTs toolkit) to the 19 Hammers evaluation datasets
-located at \url{http://www.brain-development.org/}
+evaluation applies the script {\ttfamily ants\_multitemplate\_labeling.sh }
+(available in ANTs) to the 19 Hammers evaluation datasets
+located at \url{http://www.brain-development.org/} and currently under
+the adult atlases section 
 \citep{Hammers2003,Heckemann2006}.  These initial majority voting
@@
 subject.  The following parameters are the most relevant to this discussion:
-{\ttfamily  -i  PriorProbabilityImages[3, label\_prob\%02d.nii.gz , 0.5
-  ] -m [0.1,1x1x1] -c [5,0] -p Socrates[1] }.  Results, in terms of
-Dice overlap, are shown in figure~\ref{fig:hammers}.
+{\ttfamily  -i  PriorProbabilityImages[69,label\_prob\%02d.nii.gz,0.5] 
+-m [0.2,1x1x1] -c [5,0] -p Socrates[1]}.  Results, in terms of
+Dice overlap, are shown in Figure~\ref{fig:hammers}.
 \begin{figure}
@@
 \subsection{Reproducibility of this Evaluation}
-The brainweb data is freely available.  We used single-subject
-brainweb data as is.  The 20 subject data, however, required excluding
-non-cerebrum tissue classes.  The Hammers data was also used as is.
-The scripts for the single-subject brainweb study is available in
-supplementary material.  The initialization procedure for the brainweb 20 and
-the Hammers evaluation data is based on freely available scripts
-included with the ANTs toolkit, specifically  {\verb ants_multitemplate_labeling.sh } and  {\verb atroposBwebRF40FigureExample.sh }.
+\noindent{\bf Data:} The BrainWeb data is freely available.  We used
+single-subject BrainWeb data as is but added a metaformat data header
+to the raw binary files.  An example copy of this header is contained
+in {\ttfamily atroposBwebRF40FigureExample.sh }.  The 20-subject data,
+however, required excluding non-cerebrum tissue classes.  The Hammers
+data was also used as is (\url{http://www.brain-development.org/}).
+
+\noindent{\bf Software:} The ANTs software is available at \url{http://www.picsl.upenn.edu/ANTs} with download and compilation
+instructions at \url{http://www.picsl.upenn.edu/ANTS/download.php}.
+SVN release 711 was used for the examples and evaluations performed in
+this paper.  Some components of ANTs depend on the Insight ToolKit.
+The most critical dependency, for Atropos, is the ITK statistics framework
+used to implement the univariate and multivariate parametric models.
+We linked to the git version of ITK current as of Dec. 1, 2010.  See
+\url{http://www.itk.org/Wiki/ITK/Git/Download} for instructions on git ITK.
+
+\noindent{\bf Scripts:} The complete script for the single-subject
+BrainWeb study is based on generalizing {\ttfamily atroposBwebRF40FigureExample.sh}, which is available in the ANTs
+toolkit (SVN release 711 or greater) and which reproduces Figure 2
+results.  The template-based normalization procedure for the BrainWeb
+20 and the Hammers evaluation data is based on freely available
+scripts included with ANTs, {\ttfamily
+  ants\_multitemplate\_labeling.sh} and {\ttfamily
+  buildtemplateparallel.sh}.  A release version of ANTs---with a final version of Atropos---will
+be prepared with the final version of this manuscript.
 
@@
 details and documented its performance in a variety of use cases.  We
-also showed evidence that the openly available N4 bias correction can easily used with Atropos to improve segmentation.
-Furthermore, we used multiple subject brainweb data to build dataset
-specific priors that provided the most consistent segmentation
+also showed evidence that the openly available N4 bias correction can
+easily be used with Atropos to improve segmentation.
+Furthermore, we used multiple subject BrainWeb data to build dataset-specific priors that provided the most consistent segmentation
 performance across tissues.  Finally, we used majority voting to
-initialize an Atropos EM solution to a 69 class brain parcellation
+initialize an Atropos EM solution to a 69-class brain parcellation
 problem.  Significant improvements were gained in multiple brain
@@
 
-\subsection{Performance on Brainweb Data}
+\subsection{Performance on BrainWeb Data}
 Atropos results are competitive with the state of the art.  For
-instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF (bias field) 3\%
-noise brainweb single subject data finding 0.932 (GM) and .961 (WM)
-Dice. Results on 40\% RF were .934 (GM) and .961 (WM). SPM5 exhibits
-insensitivity to bias similar to our own best results on the 40\%RF, 3
-\% noise case (MRF-$\beta$=0.2, Kmeans $+$ N4) with Dice for GM is
-0.950737 and for WM is 0.962586.  \cite{Nakamura2009} gave GM Dice
-results (brainweb single 3\% noise) of 0.962 (0\%RF), 0.964(20\%RF)
-and .956(40\%RF) which are slightly higher than either SPM5 or Atropos
+instance, \cite{Ashburner2005} (SPM5) evaluated on 0\% RF bias field, 3\%
+noise BrainWeb single subject data finding 0.932 (GM) and .961 (WM)
+Dice overlap. Results on 40\% RF bias were 0.934 (GM) and 0.961 (WM). SPM5 exhibits
+insensitivity to bias similar to our own best results on the 40\% RF bias, 3\% 
+noise case (MRF-$\beta$=0.2, K-means $+$ N4) with Dice overlap for GM is
+0.951 and for WM is 0.963.  \cite{Nakamura2009} gave GM Dice overlap
+results (BrainWeb single 3\% noise) of 0.962 (0\% RF bias), 0.964
+(20\% RF bias)
+and 0.956 (40\% RF bias) which are slightly higher than either SPM5 or Atropos
 results.  However, \cite{Nakamura2009} do not report WM or CSF numbers
-for comparison.  Topology preserving methods also perform well.
-\cite{Shiee2010} achieved Dice results for 3\% noise 20\% RF brainweb
-single subject with 0.927, 0.912 and 0.900 for WM, GM and CSF,
-respectively.  These are excellent numbers given the additional
-constraint applied to the segmentation.  
-\cite{Bazin2007c} proposed TOADS and, estimating from the
-paper's graph, showed that the average Dice accuracy for 3\% noise various RF was
-.95-.96 for WM, .93-.95 for GM and .92-.94 for CSF.
+for comparison.  Topology-preserving methods also perform well.
+\cite{Shiee2010} achieved Dice overlap for 3\% noise 20\% RF bias BrainWeb
+single subject with 0.912 (GM), 0.927 (WM) and 0.900 (CSF) Dice
+overlap.  These are excellent numbers given the topological constraint applied to the segmentation.  \cite{Bazin2007c} proposed TOADS and, estimating from the paper's
+graph, showed that the average Dice overlap accuracy for 3\% noise for various RF
+was  0.930--0.950 (GM), 0.950--0.960 (WM), and 0.920--0.940 (CSF).
 Perhaps the most recent balanced evaluation was performed in
@@
 than Dice overlap.  Because the absolute true number of GM and WM
-voxels for brainweb are known, we can convert the confusion matrix to
-Dice.  In that case, the SPM5 Dice for brainweb single GM and WM is .8847
-and .9085, respectively, while FreeSurfer and FSL's accuracy is
-lower. The best GM result for 20 subjects is obtained by SPM5: 0.9295;
-best WM Dice is from FSL: 0.9499.  We note that \cite{Klauschen2009}
-used a comprehensive evaluation where quality of brain extraction also
-contributed to the outcome.  Thus, the results must be interpreted
-slightly differently than those from other papers.
+voxels for BrainWeb are known, we can convert the confusion matrix to
+Dice overlap.  In that case, the SPM5 Dice overlap for BrainWeb
+single-subject data is 0.885 (GM) and 0.909 (WM), while FreeSurfer and FSL's accuracy
+is lower. The best GM Dice overlap result for the 20 subject BrainWeb data is obtained by SPM5:
+0.930; the best WM Dice overlap is from FSL: 0.950.  We note that
+\cite{Klauschen2009} used a comprehensive evaluation where quality of
+brain extraction also contributed to the outcome.  Thus, the results
+must be interpreted slightly differently than those from other papers.
+Finally, in our evaluation of 20-subject BrainWeb data, the prior
+probability models performed best of all the models used.  Compared to
+the K-Means based segmentation, the prior based segmentation
+performance also peaked at lower values of the MRF-$\beta$ term (0.0
+and 0.05).  This is reasonable in that the spatial priors themselves
+impose a degree of regularity on the segmentation, as in SPM5.
 
 \subsection{Performance on Hammers Data}
-Our prior work, \citep{Avants2010b}, showed that the majority vote
+Our prior work, \citep{Avants2011}, showed that the majority vote
 initialization provided to Atropos by ANTs template mapping is
-competitive with \cite{Heckemann2006}.  The Atropos EM extension of
-improved these results further.  However, the Atropos EM segmentation
+competitive with \cite{Heckemann2006}.  Overall, the Atropos EM
+extension improved these results further.  However, in a few regions
+of the mid-brain, the Atropos EM segmentation
 performed significantly worse.  This is not surprising, in that
@@
 alternative solution would be to use boundary conditions for these
-structures, as in the PriorLabelImage Atropos option.
+structures, as in the {\ttfamily PriorLabelImage} Atropos initialization option.
 
 \subsection{Clinically-Related Evaluation}
-While specifying performance on brainweb is highly
+While specifying performance on BrainWeb is highly
 valuable, clinical validation is a second important aspect of
@@
 respect to a known neurobiological outcome measure.  Atropos is
-currently used in clinical studies and a number of clinically-focused,
+currently used in clinical studies and a number of clinically focused,
 application-specific evaluations are ongoing and will constitute
 future work.  One early example of a clinically-focused Atropos
-neuroimaging application is in \citep{Avants2010c}.  An additional successful
+neuroimaging application is in \citep{Avants2010c}.  A second successful
 application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
-\citep{Tustison2010a} which also used the open-source  Glamorous Glue
-algorithm to impose topology constraints.  Thus, future work may
+\citep{Tustison2010a} which also used the open source Glamorous Glue
+algorithm to impose topology constraints \citep{Tustison2010b}.  Thus, future work may
 incorporate topology more closely into the Atropos methodology.
@@
 segmentation is non-trivial and there are numerous design alternatives
-available not only in the models selected but also in the ICM coding, alternatives to ICM and the specific method in which prior
-and likelihood are combined.  Due to the flexibility of Atropos, we
-also hope that some of its capabilities, though not evaluated here,
-are explored by the segmentation or clinical community.
+available not only in the models selected but also in the ICM coding,
+alternatives to ICM and the method in which prior and likelihood are
+combined.  Due to the flexibility of Atropos, we also hope that some
+of its capabilities, though not evaluated here, are explored by the segmentation or clinical community.
+
+\paragraph{Information Sharing Statement}
+{\color{blue}{Atropos software is available in ANTs
+    \url{http://www.picsl.upenn.edu/ANTs} 
+which depends on ITK \url{http://www.itk.org/Wiki/ITK/Git/Download}.
+The data used in this work is available in the ANTs software
+repository, BrainWeb 
+\url{http://mouldy.bic.mni.mcgill.ca/BrainWeb/} and at \url{www.brain-development.org}.
+We employed itk-SNAP for visualization \url{www.itksnap.org}.}}
 
 \paragraph{Acknowledgments}
-{This work was supported in part by NIH  {\bf {\textcolor{red}{ FIXME}}} }
+{This work was supported in part by NIH (AG17586, AG15116, NS44266, and
+NS53488).}
 
@@
 
-\message{ !name(atropos.tex) !offset(-1465) }
+%# T1_RF40_3%_N4 reference results 
+ Label 1 DICE 0.941891  RO 0.890165 TP1 0.904786 TP2 0.98217
+ Label 3 DICE 0.966776  RO 0.935688 TP1 0.962814 TP2 0.97077
+ Label 2 DICE 0.953862  RO 0.911794 TP1 0.971043 TP2 0.937279
+ AvgDice  : 0.954176
+%# T1_RF40_3%_N4 + T2_RF0_3% noise reference results 
+Label 1 DICE 0.937508  RO 0.882367 TP1 0.889406 TP2 0.99111
+ Label 3 DICE 0.96701  RO 0.936128 TP1 0.958398 TP2 0.975778
+ Label 2 DICE 0.951639  RO 0.90774 TP1 0.977645 TP2 0.926981
+ AvgDice  : 0.952052
+%# T1_RF40_3%_N4 + PD_RF0_3% noise reference results 
+ Label 1 DICE 0.944732  RO 0.895254 TP1 0.912413 TP2 0.979425
+ Label 3 DICE 0.970348  RO 0.942405 TP1 0.969169 TP2 0.971531
+ Label 2 DICE 0.957976  RO 0.919341 TP1 0.97094 TP2 0.945353
+ AvgDice  : 0.957685
+%# PD_RF0_3% noise reference results 
+ Label 2 DICE 0.905959  RO 0.828085 TP1 0.892396 TP2 0.919941
+ Label 3 DICE 0.942477  RO 0.891212 TP1 0.943874 TP2 0.941084
+ Label 1 DICE 0.837847  RO 0.720944 TP1 0.87262 TP2 0.805739
+ AvgDice  : 0.895428
+%# T2 
+ Label 1 DICE 0.931616  RO 0.871986 TP1 0.984176 TP2 0.884385
+ Label 2 DICE 0.882997  RO 0.790505 TP1 0.79822 TP2 0.987921
+ Label 3 DICE 0.901592  RO 0.820817 TP1 0.99405 TP2 0.82487
+ AvgDice  : 0.905402

### 2011-02-18 src/ANTS/ImageSegmentation/antsAtroposSegmentationImageFilter.h

@@ -120,7 +120,10 @@ public:
     <SampleType, RealType, RealType>                  LikelihoodFunctionType;
   typedef typename LikelihoodFunctionType::Pointer    LikelihoodFunctionPointer;
   typedef typename LikelihoodFunctionType::
-    WeightArrayType                                   WeightArrayType;
+    ListSampleWeightArrayType                         WeightArrayType;
+
+  typedef std::vector<LabelType>                      PartialVolumeLabelSetType;
+  typedef std::vector<PartialVolumeLabelSetType>      PartialVolumeClassesType;
 
   /** Outlier handling typedefs */
   typedef ants::Statistics::
@@ -158,16 +161,53 @@ public:
   // ivars Set/Get functionality
 
   /**
-   * Set the number of segmentation classes which is clamped from below at 2.
+   * Set the number of tissue classes which is clamped from below at 2.
    * Default = 3.
    */
-  itkSetClampMacro( NumberOfClasses, LabelType, 2,
+  itkSetClampMacro( NumberOfTissueClasses, LabelType, 2,
     NumericTraits<unsigned int>::max() );
 
   /**
    * Get the number of segmentation classes.
    */
-  itkGetConstMacro( NumberOfClasses, unsigned int );
+  itkGetConstMacro( NumberOfTissueClasses, unsigned int );
+
+  /**
+   * Set the partial-volume-label set one at a time.
+   */
+  void AddPartialVolumeLabelSet( PartialVolumeLabelSetType );
+
+  /**
+   * Get the number of partial volume classes
+   */
+  itkGetConstMacro( NumberOfPartialVolumeClasses, unsigned int );
+
+  /**
+   * The user can specify whether or not to use the partial volume likelihoods,
+   * in which case the partial volume class is considered separate from the
+   * tissue classes.  Alternatively, one can use the MRF only to handle
+   * partial volume in which case, partial volume voxels are not considered
+   * as separate classes.
+   */
+  itkSetMacro( UsePartialVolumeLikelihoods, bool );
+
+  /**
+   * The user can specify whether or not to use the partial volume likelihoods,
+   * in which case the partial volume class is considered separate from the
+   * tissue classes.  Alternatively, one can use the MRF only to handle
+   * partial volume in which case, partial volume voxels are not considered
+   * as separate classes.
+   */
+  itkGetConstMacro( UsePartialVolumeLikelihoods, bool );
+
+  /**
+   * The user can specify whether or not to use the partial volume likelihoods,
+   * in which case the partial volume class is considered separate from the
[…]

### 2011-02-18 src/ANTS/Scripts/antsaffine.sh

@@ -1,5 +1,12 @@
 #!/bin/sh 
-export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"}
+if [ ${#ANTSPATH} -le 3 ] ; then 
+  echo we guess at your ants path 
+  export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"} # EDIT THIS 
+fi 
+if [ ! -s ${ANTSPATH}/ANTS ] ; then 
+  echo we cant find the ANTS program -- does not seem to exist.  please \(re\)define \$ANTSPATH in your environment.
+  exit
+fi 
 
 NUMPARAMS=$#
 

### 2011-02-18 src/ANTS/Scripts/buildtemplateparallel.sh

@@ -6,7 +6,14 @@ VERSION="0.0.13"
 trap control_c SIGINT
 
 # Uncomment the line below in case you have not set the ANTSPATH variable in your environment. 
- export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"}
+if [ ${#ANTSPATH} -le 3 ] ; then 
+  echo we guess at your ants path 
+  export ANTSPATH=${ANTSPATH:="$HOME/bin/ants/"} # EDIT THIS 
+fi 
+if [ ! -s ${ANTSPATH}/ANTS ] ; then 
+  echo we cant find the ANTS program -- does not seem to exist.  please \(re\)define \$ANTSPATH in your environment.
+  exit
+fi 
 
 #assuming .nii.gz as default file type. This is the case for ANTS 1.7 and up
 

### 2011-02-18 src/ANTS/Tensor/TensorFunctions.h

@@ -125,7 +125,8 @@ TensorType TensorLogAndExp( TensorType dtv, bool takelog , bool success=true)
   double e1 = D(0,0);
   double e2 = D(1,1);
   double e3 = D(2,2);
-
+  if ( e1 < 0 ) e1=e2;
+  if ( e3 < 0 ) e3=e2;
   //float peigeps=1.e-12;   
 
   if ( fabs(e3) < eps ) { success=false; std::cout << "-4" << std::flush; return dtv; }
@@ -177,63 +178,27 @@ TensorType TensorExp( TensorType dtv, bool takelog , bool success=true)
 template <class TensorType>
 float  GetTensorFA( TensorType dtv )        
 {
-
-// inner product
-  float xx = dtv[0];
-  float xy = dtv[1];
-  float xz = dtv[2];
-  float yy = dtv[3];
-  float yz = dtv[4];
-  float zz = dtv[5];
-  float isp = ( xx*xx + yy*yy + zz*zz + 2.0*(xy*xy + xz*xz + yz*yz) );
-
-  // Computed as 
-  // FA = vcl_sqrt(1.5*sum(sum(N.*N))/sum((sum(D.*D))))
-  // where N = D - ((1/3)*trace(D)*eye(3,3))
-  // equation (28) in http://lmi.bwh.harvard.edu/papers/pdfs/2002/westinMEDIA02.pdf
-
-  if( isp > 0.0 )
-    {
-      float trace = dtv[0];
-      trace += dtv[3];
-      trace += dtv[5];
-      
-      float anisotropy = 3.0 * isp - trace * trace;
-      float fractionalAnisotropy =( vcl_sqrt(anisotropy / ( 2.0 * isp ) ) );
-      return fractionalAnisotropy;
-    }
- 
-   return 0.0 ;
-
-   /*
-   // Computed as 
-   // FA = vcl_sqrt(1.5 * ( \sum_i ( lambda_i - lambda_mean )^2 ) / \sum_i ( lambda_i^2 ) )
-   // as in http://splweb.bwh.harvard.edu:8000/pages/papers/martha/DTI_Tech354.pdf
-   // [lambda = eig(A)].
-   EigenValuesArrayType eigenValues;
-   ComputeEigenValues( eigenValues );
-   eigenValues[0] = vnl_math_abs(eigenValues[0]);
[…]

### 2011-02-19 src/ANTS/ImageRegistration/ANTS_affine_registration2.h
ENH: fixing some stuff that prevents windows compilation --- using visual studio .net 2003 ... ANTS_aff2 , xcor , prob , SMI round and log all needed changes to eliminate errors or warnings

@@ -153,7 +153,7 @@ void WriteAffineTransformFile(TransformPointerType &transform, StringType filena
     transform_writer->Update();
   }
   catch( itk::ExceptionObject &err){
-    std::cout << "!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!" << std::endl
+    std::cout << "!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!" << err << std::endl
         <<"Exception in writing tranform file: " << std::endl
         << filename << std::endl;
     return;
@@ -575,7 +575,7 @@ double TestCostValueMMI(ImagePointerType fixedImage, ImagePointerType movingImag
     rval = mattesMutualInfo->GetValue(para);
   }
   catch (itk::ExceptionObject &err){
-    std::cout << "!!!!!!!!!!!!!!!!!!!!!!!!!!!!!" << std::endl
+    std::cout << "!!!!!!!!!!!!!!!!!!!!!!!!!!!!!" << err << std::endl
         << "Exception caught in computing mattesMutualInfo after registration" << std::endl
         << "Maybe: Too many samples map outside moving image buffer" << std::endl
         << "Set the cost value = 0 (max for MutualInfo) " << std::endl;

### 2011-02-19 src/ANTS/ImageRegistration/itkCrossCorrelationRegistrationFunction.h

@@ -203,12 +203,12 @@ public:
   double fixedGradientSquaredMagnitude = 0;
   fixedValue = (double) Superclass::Superclass::m_FixedImage->GetPixel( index );
   fixedGradient = m_FixedImageGradientCalculator->EvaluateAtIndex( index );
-  for( unsigned int j = 0; j < ImageDimension; j++ )
+  unsigned int j=0;
+  for( j = 0; j < ImageDimension; j++ )
     {
     fixedGradientSquaredMagnitude += vnl_math_sqr( fixedGradient[j] );
     } 
   double movingValue;
-  int j;
   PointType mappedPoint;
   for( j = 0; j < ImageDimension; j++ )
     {

### 2011-02-19 src/ANTS/ImageRegistration/itkProbabilisticRegistrationFunction.h

@@ -201,12 +201,13 @@ public:
   double fixedGradientSquaredMagnitude = 0;
   fixedValue = (double) Superclass::Superclass::m_FixedImage->GetPixel( index );
   fixedGradient = m_FixedImageGradientCalculator->EvaluateAtIndex( index );
-  for( unsigned int j = 0; j < ImageDimension; j++ )
+  unsigned int j=0;
+  for( j = 0; j < ImageDimension; j++ )
     {
     fixedGradientSquaredMagnitude += vnl_math_sqr( fixedGradient[j] );
     } 
   double movingValue;
-  int j;
+  
   PointType mappedPoint;
   for( j = 0; j < ImageDimension; j++ )
     {

### 2011-02-19 src/ANTS/ImageRegistration/itkSpatialMutualInformationRegistrationFunction.h

@@ -409,7 +409,7 @@ public:
 		
 	//	std::cout << " JE " << JointEntropy << " JEXuY " << JointEntropyXuY << " JEXYu " << JointEntropyXYu << " JEXuYr " << JointEntropyXuYr << std::endl; 
 		
-		this->m_Energy = (-1.0) * fabs(SMI/log(2));
+		this->m_Energy = (-1.0) * fabs(SMI);
 		return this->m_Energy;
 	}
 
@@ -474,7 +474,7 @@ public:
 	unsigned int FitIndexInBins( double windowTerm ) 
 	{
 		unsigned int movingImageParzenWindowIndex  = 
-			static_cast<unsigned int>( this->m_Padding + round( windowTerm * (float)(this->m_NumberOfHistogramBins - 1 - this->m_Padding))) ;
+			static_cast<unsigned int>( this->m_Padding + (unsigned int)( windowTerm * (float)(this->m_NumberOfHistogramBins - 1 - this->m_Padding + 0.5))) ;
 
 		// Make sure the extreme values are in valid bins
 		if ( movingImageParzenWindowIndex < this->m_Padding )

### 2011-02-19 src/ANTS/Utilities/itkN3MRIBiasFieldCorrectionImageFilter.h
ENH: had to redefine function declarations in N3 filter because of vs stupidity

@@ -274,10 +274,11 @@ private:
   N3MRIBiasFieldCorrectionImageFilter( const Self& ); //purposely not implemented
   void operator=( const Self& ); //purposely not implemented
 
-  typename RealImageType::Pointer SharpenImage(
-    typename RealImageType::Pointer );
-  typename RealImageType::Pointer SmoothField(
-    typename RealImageType::Pointer );
+  typename N3MRIBiasFieldCorrectionImageFilter<TInputImage, TMaskImage, TOutputImage>::RealImageType::Pointer
+    SharpenImage( typename N3MRIBiasFieldCorrectionImageFilter<TInputImage, TMaskImage, TOutputImage>::RealImageType::Pointer
+ );
+ typename N3MRIBiasFieldCorrectionImageFilter<TInputImage, TMaskImage, TOutputImage>::RealImageType::Pointer
+ SmoothField( typename N3MRIBiasFieldCorrectionImageFilter<TInputImage, TMaskImage, TOutputImage>::RealImageType::Pointer);
   RealType CalculateConvergenceMeasurement(
     typename RealImageType::Pointer,
     typename RealImageType::Pointer );
